---
title: "Model diagnostics with dynhr"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Model diagnostics with dynhr}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>",
  eval     = FALSE   # most diagnostics require estimation; show code only
)
```

## Overview

dynhr ships an extended diagnostic battery (D1–D30) that evaluates a DSGE model
across six stages of the workflow:

| Stage | Diagnostics | When to run |
|-------|-------------|-------------|
| Pre-estimation | D1–D4, D19–D26, Expectations | Before drawing from the posterior |
| MCMC quality | D5–D7 | After sampling |
| Post-estimation fit | D8–D18 | After sampling |
| OBC identification | D27 | After OBC model setup |
| Regime-switching | D28 | After Markov-switching specification |
| Data-driven constraints | D29 | When external data moments are available |
| Arbitrary-precision | D30 | For high-precision rank validation |
| Model comparison | Bayes factor, parent comparison | When comparing two models |

Every diagnostic returns a `dynhr_diagnostic` object with a standardised
shape:

```r
list(
  pass    = TRUE / FALSE / NA,   # NA means "informational only"
  summary = "text description",
  result  = <diagnostic-specific data>,
  plots   = list(...)            # ggplot2 objects, if any
)
```

`run_diagnostics()` runs all of them conditionally — skipping those whose
required inputs are absent — and returns a named list of class
`dynhr_diagnostic_suite`.

---

## 1  Running all diagnostics

```{r run-all}
library(dynhr)

# Assume you have completed estimation (see vignette("estimation"))
diags <- run_diagnostics(
  model            = model,        # from parse_mod()
  compiled         = compiled,
  ss               = steady$values,
  params           = theta_mode,
  priors           = priors,
  data             = Y,
  draws            = chains$chain,
  chains_list      = list(chains$chain),   # for Gelman-Rubin
  irf              = irfs,
  vd               = moms$variance_decomposition,
  hd               = hd,           # from historical_decomposition()
  shocks           = shocks,       # smoothed shocks
  obs_names        = colnames(Y),
  param_names      = colnames(chains$chain),
  data_moments     = list(sigma_y = empirical_sd, acf_y = empirical_acf),
  model_moments    = moms,
  dates            = dates,        # optional Date vector for plots
  model_name       = "nk_hs2016",
  verbose          = TRUE
)
```

### Printing the suite

```{r print-diags}
print(diags)
```

The console output shows one line per diagnostic with a `[PASS]`,
`[FAIL]`, or `[INFO]` badge plus a truncated summary.

---

## 2  Pre-estimation diagnostics

### D1 — Local identification

```{r d1}
d1 <- diag_identification <- diags$d1
cat(d1$summary)
```

Checks that the model is locally identifiable by examining whether the
Jacobian of model-implied moments with respect to estimated parameters
has full column rank.  A rank defect flags unidentified parameter
combinations.

**Inputs required:** `model_solve_fn` (a function `theta -> moments`),
calibrated `params`, `param_names`, `moment_names`.

### D2 — Spectral identification (retired)

D2 was a placeholder stub for the Komunjer–Ng (2011) spectral identification
test and has been retired.  The full implementation is now in D23 (see
below); running `diag_meta()` will not list a D2 entry.

### D3 — Sensitivity analysis (Morris screening)

```{r d3}
cat(diags$d3$summary)
```

Uses Morris one-at-a-time screening to identify which parameters have
negligible influence on model moments, suggesting they cannot be
identified from data.

**Inputs required:** `model_solve_fn`, `params`, `param_bounds`,
`moment_names`.

### D4 — Prior predictive checks

```{r d4}
cat(diags$d4$summary)
```

Draws parameter vectors from the prior and checks that the resulting
model moments are consistent with the data moments (means, standard
deviations).  A prior that consistently places moment predictions far
outside the data should be tightened.

**Inputs required:** `model_solve_fn`, `prior_draw_fn`, `data_moments`.

### Expectations check

```{r expectations}
cat(diags$expectations$summary)
```

Checks user-declared expectations from the `@dynhr:expectations` block
in the `.mod` file.  Typical checks:

- **`data_mean`**: Is the mean of an observed series in a plausible range?
- **`data_ratio`**: Is one series's mean a plausible multiple of another?
- **`param_range`**: Does an estimated parameter fall in a reasonable range?
- **`irf_sign`**: Does an IRF have the expected sign at a given horizon?

Add expectations to any `.mod` file:

```
// @dynhr:expectations
// infl_mean:  type="data_mean",   variable="INFL", min=0.0, max=6.0,
//             description="Annualised inflation should be positive and moderate"
// rho_R_hi:   type="param_range", variable="rho_R", min=0.5, max=0.99,
//             description="Interest rate smoothing should be moderate to high"
// @dynhr:end
```

Run it directly:

```{r expectations-direct, eval = TRUE}
library(dynhr)
mod_path <- system.file("extdata", "models", "nk_hs2016.mod", package = "dynhr")
model_nk <- parse_mod(mod_path, verbose = FALSE)

# Check param_range expectations with calibrated values
theta_calib <- c(rho_R = 0.85, rho_g = 0.70, rho_z = 0.65)
exp_result  <- diag_expectations(model_nk, params = theta_calib)
cat(exp_result$summary, "\n")
```

---

## 3  MCMC quality diagnostics

### D5 — Convergence (R-hat, ESS, trace)

```{r d5}
cat(diags$d5$summary)
```

Computes the Gelman–Rubin potential scale reduction factor (R̂) and
effective sample size (ESS) for each parameter.  The diagnostic
**passes** when `max(R̂) < 1.05` and `min(ESS) > 100`.

Plots (if `ggplot2` is installed):
- Trace plots for each parameter
- R̂ bar chart

**Rule of thumb:**
- R̂ > 1.1: chain mixing is poor — increase draws or run more warmup
- ESS < 50: proposal scale is too large or the posterior is highly correlated

### D6 — Posterior vs prior

```{r d6}
cat(diags$d6$summary)
```

For each estimated parameter, overlays the prior and posterior densities
and reports the **overlap coefficient** (fraction of shared density area).
Large overlap (> 0.8) may indicate the data contain little information
about that parameter.

The ridge-plot output shows all parameters on one panel for easy
comparison.

### D7 — Mode robustness

```{r d7}
cat(diags$d7$summary)
```

If `find_mode()` was run with `n_starts > 1`, D7 checks whether all
starting points converged to the same mode.  Dispersion in modal log-posterior
values greater than 2 natural-log units suggests a multimodal posterior.

---

## 4  Post-estimation fit diagnostics

### D8 — IRF plausibility

```{r d8}
cat(diags$d8$summary)
```

Checks that each IRF benchmark declared in the `@dynhr:benchmarks`
block of the `.mod` file is met.  For example:

```
// @dynhr:benchmarks
// mp_output: shock="eps_R", variable="y", horizon=4, sign="negative",
//            description="Monetary contraction depresses output"
// @dynhr:end
```

Returns **PASS** when all declared benchmarks are satisfied; lists the
failing ones otherwise.

### D9 — Moment matching

```{r d9}
cat(diags$d9$summary)
```

Compares theoretical moments (standard deviations, cross-correlations,
autocorrelations) from the model at the posterior mode against empirical
moments from the data.  Reports the ratio of model SD to data SD for
each observable — ratios outside [0.5, 2.0] flag a systematic fit
problem.

### D10 — Variance decomposition

```{r d10}
cat(diags$d10$summary)
```

Reports the fraction of variance in each observable attributable to each
structural shock.  Useful for identifying whether the estimated model is
driven by implausibly few (or many) shocks.

### D11 — Historical decomposition

```{r d11}
cat(diags$d11$summary)
```

Decomposes the historical path of each observed variable into shock
contributions.  Uses the smoother output from `kalman_smoother()`.
The plot shows stacked bar charts of shock contributions for each
observable through the sample period.

### D12 — Smoothed shocks

```{r d12}
cat(diags$d12$summary)
```

Checks distributional properties of smoothed shocks: are they roughly
serially uncorrelated (Ljung–Box test) and normally distributed
(Shapiro–Wilk test)?  Persistent or fat-tailed shocks suggest model
misspecification.

### D13 — Cross-equation restrictions

D13 compares the DSGE-implied VAR coefficients (derived via Yule–Walker
from the state-space representation) against an unrestricted OLS VAR on
the same data.  The diagnostic reports the Frobenius norm of the
difference between DSGE and OLS coefficient matrices and computes a
Wald-type test p-value.  A low p-value suggests the model's
cross-equation restrictions are rejected by the data.

**Inputs required:** `dr`, `data`, `obs_names`.

### D14 — Bayes factor model comparison

```{r d14}
# Compare two models
models_list <- list(
  baseline  = list(log_ml = chains_baseline$log_marginal_lik),
  augmented = list(log_ml = chains_augmented$log_marginal_lik)
)
d14 <- d14_bayes_factor(models_list)
cat(d14$summary)
```

Computes log Bayes factors between all pairs in `models_list`.  SMC
provides the log marginal likelihood `log p(Y | M)` directly; for RWMH
chains, use the Harmonic Mean estimator (with caution).

| log₁₀ BF | Interpretation |
|----------|----------------|
| 0 – 0.5  | Barely worth mentioning |
| 0.5 – 1  | Substantial |
| 1 – 2    | Strong |
| > 2      | Decisive |

### D15 — DSGE-VAR tightness

D15 searches over a grid of hyperparameter λ (the tightness of the
model prior relative to the unrestricted VAR) and computes the optimal
λ* that maximises the log marginal likelihood.  Returns the optimal
tightness, marginal likelihoods across the grid, and a comparison of
DSGE-VAR coefficients vs the unrestricted OLS VAR.

**Inputs required:** `dr`, `data`, `obs_names`.

### D16 — Subsample stability

```{r d16}
# Split chains at the midpoint and compare
n_mid <- nrow(chains$chain) %/% 2L
results_sub <- list(
  first_half  = chains$chain[1:n_mid, ],
  second_half = chains$chain[(n_mid+1):nrow(chains$chain), ]
)
d16 <- d16_subsample_stability(chains$chain, results_sub, param_names)
cat(d16$summary)
```

Compares posterior means and standard deviations across two subsamples.
Instability flags parameters whose posterior means shift by more than
one posterior standard deviation across subsamples.

### D17 — Narrative identification

```{r d17}
d17 <- d17_narrative_identification(hd, shocks, episodes = model$metadata$narratives)
cat(d17$summary)
```

Evaluates narrative sign restrictions from the `@dynhr:narratives` block.
For each episode (e.g. a financial crisis), checks whether the shocks from
that category dominate the smoothed shocks during that period.

### D18 — Welfare plausibility

D18 consumes `ramsey_result` when available (for example from
`run_full_estimation(..., run_ramsey = TRUE)`) and reports:

- simulated mean planner objective
- discounted steady welfare
- discounted unconditional welfare
- welfare gap relative to steady state

If no Ramsey output is provided, D18 remains informational and explains how to
enable the welfare check.

---

## 4  Extended diagnostics (D19–D30)

The extended diagnostic suite covers identification, calibration,
nonlinear, and data-driven checks that go beyond the core battery.

### D19 — IRF comparison

```{r d19}
cat(diags$d19$summary)
```

Compares IRFs from the estimated model against empirical IRFs (from a
VAR or local projections) or against a reference model.  Reports the
mean-squared deviation across horizons and flags IRFs that fall outside
the empirical confidence bands.

**Inputs required:** `irf` (model IRFs), `data` (for empirical IRFs).

### D20 — Fisher information strength

```{r d20}
cat(diags$d20$summary)
```

Computes the Fisher information matrix of the model moments with respect
to the estimated parameters.  Reports the determinant, condition number,
and per-parameter strength (the inverse of the Cramér–Rao lower bound).
Low strength indicates parameters that are difficult to identify from the
data.

### D21 — KPS precision

```{r d21}
cat(diags$d21$summary)
```

Implements the KPS (Koop–Pesaran–Smith) precision measure, which
evaluates how precisely parameters can be estimated at different sample
sizes.  Useful for experimental design and for determining the data
requirements for reliable estimation.

### D22 — Observable information

```{r d22}
cat(diags$d22$summary)
```

Decomposes the Fisher information by observable variable, identifying
which observed series contribute the most information for each parameter.
This helps focus data collection and assess whether the observable set
is sufficient for the parameter set.

### D23 — Spectral identification

```{r d23}
cat(diags$d23$summary)
```

Full implementation of the Komunjer–Ng (2011) spectral identification
test.  Assesses local identification in the frequency domain by
examining the rank of the spectral density matrix at each frequency.
Complements the time-domain D1 check.

### D24 — Global KL identification

```{r d24}
cat(diags$d24$summary)
```

Computes the Kullback–Leibler divergence between the spectral density
at the baseline parameter vector and at perturbed values (Qu & Tkachenko
2017).  Parameters whose KL divergence is small over a wide range are
globally unidentified.  Accepts Ramsey-augmented decision rules when a
`ramsey_result` is provided.

### D25 — Higher-order identification

```{r d25}
cat(diags$d25$summary)
```

Extends the D1/D19 local identification analysis to higher-order moments:
skewness (third cumulant) and, when available, excess kurtosis (fourth
cumulant).  Reports the gain in identification strength from including
higher-order information versus first-order moments alone.

### D26 — Calibration sensitivity

```{r d26}
cat(diags$d26$summary)
```

Screens all calibrated (non-estimated) parameters for their influence on
the model moments, using Morris one-at-a-time screening.  Parameters with
high sensitivity should be closely scrutinised or moved to the estimated
set; parameters with negligible sensitivity can be safely calibrated.

### D27 — OBC identification

```{r d27}
cat(diags$d27$summary)
```

Assesses identification across OBC slack and binding regimes.  Builds
regime-specific Jacobians and computes per-regime identification strength.
Parameters that are well identified in one regime but not another may
require regime-specific priors or data splits.

**Inputs required:** `obc_specs` (a list of OBC specification objects).

### D28 — Regime-switching identification

```{r d28}
cat(diags$d28$summary)
```

For Markov-switching models, computes per-regime and ergodic-weighted
identification diagnostics.  Uses the ergodic distribution over regimes
to weight the regime-specific Jacobians into an overall identification
assessment.

### D29 — Data-driven constraints

```{r d29}
cat(diags$d29$summary)
```

Implements the Lanne & Luoto (2017) approach: compares model-implied
identification constraints against data-driven constraints derived from
a HAC (Newey–West) covariance matrix of the data moments.  Parameters
that are identified by the data but constrained by the model (or vice
versa) are flagged.

### D30 — Arbitrary-precision rank checks

```{r d30}
cat(diags$d30$summary)
```

Recomputes the SVD-based rank of the moment Jacobian at elevated
precision using either the **Rmpfr** package (preferred) or **JuliaCall**
(fallback).  Default precision is 128 bits (~38 decimal digits).  A rank
that increases at high precision suggests near-rank deficiency that may
be masked by floating-point rounding at standard double precision.

---

## 5  OBC diagnostics

The OBC diagnostic section summarises binding episodes from a solved
OBC path:

```{r obc, eval = FALSE}
obc_diag <- d_obc_binding_summary(
  regime_path      = occbin_result$regimes,
  constraint_names = c("zlb"),
  var_paths        = occbin_result$paths,
  var_names        = model$var_names
)
cat(obc_diag$summary)
```

This is informational (`[INFO]`) when called via `run_diagnostics()`
without explicit OBC simulation data; pass OBC results directly for
a full binding-episodes analysis.

---

## 6  Writing a report

```{r write-report}
write_report(diags, file = "output/report.md")
```

`write_report()` writes a structured plain-text report optimised for
LLM consumption: each diagnostic's pass/fail/error status, numeric
summary, and a brief interpretation.  HTML and PDF report formats
are also supported via `format_llm_report()`.

---

## 7  Running individual diagnostics

Every diagnostic is also callable directly for targeted inspection:

```{r individual}
# Run only D5 (convergence) and D8 (IRF plausibility)
d5 <- d5_mcmc_convergence(chains$chain, param_names, chains_list = list(chains$chain))
d8 <- d8_irf_plausibility(irfs)

cat("D5:", d5$summary, "\n")
cat("D8:", d8$summary, "\n")
```

---

## 8  Custom expectations

The `@dynhr:expectations` block is the recommended way to encode
model-specific quantitative checks.  These checks replace ad hoc Stage 4
logic in pipeline scripts:

```
// @dynhr:expectations
// cy_ratio:    type="data_ratio",  numerator="c_obs", denominator="y_obs",
//              min=0.5, max=2.0,   description="C/Y plausibility"
// rho_a_hi:    type="param_range", variable="rho_a",
//              min=0.5, max=0.999, description="Tech AR persistence"
// mp_irf_sign: type="irf_sign",    shock="eps_R", variable="y",
//              horizon=4, sign="negative",
//              description="Monetary contraction depresses output at h=4"
// @dynhr:end
```

Call `diag_expectations()` directly or let `run_diagnostics()` invoke it
automatically via the model's metadata.

---

## References

Antolin-Diaz, J., & Rubio-Ramirez, J. F. (2018). Narrative sign restrictions for DSGE models. *Journal of Monetary Economics*, 102, 95-112.

Gelman, A., & Rubin, D. B. (1992). Inference from iterative simulation using multiple sequences. *Statistical Science*, 7(4), 457-472.

Gelman, A., Meng, X.-L., & Stern, H. (1996). Posterior predictive assessment of model fitness via realized discrepancies. *Statistica Sinica*, 6(4), 733-760.

Geweke, J. (1992). Evaluating the accuracy of sampling-based approaches to the calculation of posterior moments. In *Bayesian Statistics 4*, 169-193. Oxford University Press.

Iskrev, N. (2010). Local identification in DSGE models. *Journal of Monetary Economics*, 57(2), 189-202.

Kass, R. E., & Raftery, A. E. (1995). Bayes factors. *Journal of the American Statistical Association*, 90(430), 773-795.

Komunjer, I., & Ng, S. (2011). Dynamic identification of dynamic stochastic general equilibrium models. *Econometrica*, 79(6), 1995-2032.

Lanne, M., & Luoto, J. (2017). Data-driven constraints for DSGE models. *Journal of Applied Econometrics*, 32(3), 609-624.

Ljung, G. M., & Box, G. E. P. (1978). On a measure of lack of fit in time series models. *Biometrika*, 65(2), 297-303.

Lubik, T. A., & Schorfheide, F. (2004). Testing for indeterminacy: An application to U.S. monetary policy. *American Economic Review*, 94(1), 190-217.

Morris, M. D. (1991). Factorial sampling plans for preliminary computational experiments. *Technometrics*, 33(2), 161-174.

Qu, Z., & Tkachenko, D. (2017). Global identification in DSGE models. *Journal of Econometrics*, 200(2), 271-292.

Schorfheide, F. (2000). Loss function-based evaluation of DSGE models. *Journal of Applied Econometrics*, 15(6), 645-670.

Sims, C. A. (1980). Macroeconomics and reality. *Econometrica*, 48(1), 1-48.

---

```{r session, eval = TRUE}
sessionInfo()
```
