---
title: "Getting started with dynhr"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting started with dynhr}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse  = TRUE,
  comment   = "#>",
  fig.width = 7,
  fig.height = 4,
  eval = TRUE
)
```

## Overview

**dynhr** is a self-contained R toolkit for DSGE modelling.  It parses
Dynare-format `.mod` files, solves the model by first- through
fifth-order perturbation, runs the Kalman filter and smoother (including
block-Kalman and OBC PKF variants), and provides a suite of Bayesian
estimation samplers (RWMH, SMC, SMC², NUTS, DIME, PMMH).  A diagnostic
battery (D0–D41) helps you evaluate whether the estimated model is
numerically and economically plausible.

The package ships with a small corpus of reference models (RBC, three-equation
New Keynesian, Herbst–Schorfheide 2016, Lubik–Schorfheide 2003, the
three-equation New Keynesian model `nk_demo` — which also ships its
own 192-quarter dataset — the `03j` NZ production model, and the
Caldara et al. 2012 financial-accelerator model) with regression goldens
produced from Dynare.jl and Octave Dynare 7.0.

```{r library}
library(dynhr)
```

---

## 1  Parse a model

The entry point is `parse_mod()`.  Point it at any Dynare `.mod` file:

```{r parse}
mod_path <- system.file("extdata", "models", "rbc.mod", package = "dynhr")
model    <- parse_mod(mod_path, verbose = FALSE)
print(model)
```

The result is a `dynhr_mod` list with named slots for endogenous
variables, exogenous shocks, parameters, equations, lead/lag incidence,
and (optionally) estimated-parameter priors and `@dynhr:` metadata blocks.

---

## 2  Compile

Compilation builds the static and dynamic residual/Jacobian functions via
symbolic differentiation:

```{r compile}
compiled <- compile_model(model, verbose = FALSE)
```

---

## 3  Solve

### 3a  Steady state

```{r steady}
params <- model$param_values
steady  <- solve_steady(compiled, params,
                        endo_names = model$var_names,
                        exo_names  = model$varexo_names,
                        verbose    = FALSE)
print(steady)
```

### 3b  First-order perturbation (decision rules)

```{r dr}
dr <- solve_perturbation(model, compiled, steady$values, params,
                         verbose = FALSE)
print(dr)
```

`dr$bk_satisfied` is `TRUE` when the model satisfies Blanchard–Kahn
conditions.  `dr$ghx` (state transition) and `dr$ghu` (shock impact)
are the standard decision-rule matrices.

---

## 4  Simulate

### 4a  Impulse response functions

```{r irfs}
irfs <- compute_irfs(dr, model, n_periods = 40L)
# irfs$eps_a is a 40 × 7 matrix (horizons × variables)
head(irfs$eps_a[, c("y", "c", "k")])
```

### 4b  Theoretical moments

```{r moments}
moms <- compute_moments(dr, model)
round(moms$correlation[1:4, 1:4], 3)
```

### 4c  Stochastic simulation

```{r stochsimul}
sim <- stoch_simul(model, verbose = FALSE)
```

`stoch_simul()` wraps all three in one call and returns a list with
`$dr`, `$irfs`, and `$moments`.

### 4d  Monte Carlo simulation paths

```{r simulate}
set.seed(42)
paths <- simulate_model(dr, n_periods = 100L, model = model, burn_in = 20L)
head(paths[, c("y", "c", "k")], 3)
```

---

## 5  Kalman filter

Given observed data, evaluate the log-likelihood:

```{r kf}
# Use simulated data as a stand-in for real observations
obs_vars <- c("y", "c")
Y <- as.matrix(paths[, obs_vars])

suppressWarnings(
  kf <- kalman_filter(Y, dr, model, params,
                      obs_vars        = obs_vars,
                      return_filtered = TRUE,
                      me_variance     = 1e-6)
)
cat("Log-likelihood:", round(kf$loglik, 3), "\n")
```

---

## 6  Full estimation in one call

For a model with an `estimated_params` block, `run_full_estimation()`
runs the entire pipeline — parse → compile → mode-finding → MCMC →
optional diagnostics — and saves outputs to a directory.

A model *and* a dataset ship with the package, so this runs from a clean
install with no data of your own — `nk_demo.mod` is a three-equation New
Keynesian model (nine estimated parameters) and `nk_demo_data.csv` is
192 quarters of US output- and price-growth observations:

```{r rfe, results = "hide"}
mod_file <- system.file("extdata/models/nk_demo.mod", package = "dynhr")
est_data <- read.csv(system.file("extdata/models/nk_demo_data.csv",
                                 package = "dynhr"))

result <- run_full_estimation(
  mod_file    = mod_file,
  data        = est_data,                    # T × n_obs data.frame or matrix
  obs_vars    = c("ygr", "infl", "intr"),
  output_dir  = tempdir(),
  sampler     = "smc",                       # no mode required; gives log p(Y|M)
  n_particles = 500L,                        # raise for a production run
  seed        = 1L,
  verbose     = FALSE
)
```

```{r rfe-summary}
summary(result$chains)              # posterior mean / sd / 5-50-95%
result$chains$log_marginal_lik      # SMC estimate of log p(Y | M)
```

For a random-walk Metropolis run instead, pass `sampler = "rwmh"` with
`n_draws`, `n_warmup` and `n_chains` (multi-chain runs add Gelman–Rubin
convergence diagnostics).

The returned `dynhr_estimation_result` object carries the parsed model,
compiled model, prior specification, mode, chains, convergence
diagnostics, and run metadata.

Alternatively, use the stepwise API: `run_mode_finding()` for mode-finding
only, or `run_posterior_estimation()` for sampling from a pre-computed mode.
`make_posterior()` also accepts `likelihood = "cumulant"` to estimate
using third- and fourth-order cumulants (skewness and kurtosis) alongside
the usual Gaussian likelihood, following Mutschler (2015).

---

## 6b  One pipeline object instead of six threaded calls

The stepwise API above asks you to re-thread the same pieces — model,
compiled model, data, `obs_vars`, prior spec — through every call.
`dynhr_model()` bundles them once, and the `dm_*()` verbs take that object
as their first argument and hand back a new one with the step's result
stored in a slot, so the pipeline pipes:

```{r dm-object, results = "hide"}
obs_vars <- c("ygr", "infl", "intr")
Y <- as.matrix(read.csv(system.file("extdata/models/nk_demo_data.csv",
                                    package = "dynhr"))[, obs_vars])

dm <- dynhr_model(system.file("extdata/models/nk_demo.mod", package = "dynhr"),
                  data = Y, obs_vars = obs_vars) |>
  dm_solve() |>
  dm_posterior() |>
  dm_mode(n_iter = 200L, verbose = FALSE)
```

```{r dm-print}
dm
dm$mode$logpost
```

The verbs are `dm_solve()`, `dm_posterior()`, `dm_mode()`, `dm_sample()`
(`sampler = "rwmh"`, `"nuts"`, `"smc"` or `"dime"`), `dm_diagnostics()`,
`dm_irf()` and `dm_forecast()`.  Each one is a pure re-threading of the
stored fields into the function documented above — `dm_solve()` *is*
`solve_perturbation()`, `dm_sample()` *is* `mcmc()`/`nuts()`/`smc()`/`dime()`
— so results are identical to the functional pipeline, and the raw result
objects (`dm$solution`, `dm$log_post`, `dm$chains`, …) are the same objects
those functions return.  Nothing runs implicitly: a verb whose prerequisite
is missing errors and names the step to run first.

---

## 7  Diagnostics

After estimation, run the diagnostic battery (D0–D41) covering
pre-estimation identification, MCMC quality, post-estimation fit, OBC
identification, regime-switching identification, and more.  Each stage is
skipped when its inputs are absent, so you can call it with whatever you have:

```{r diag, eval = FALSE}
diags <- run_diagnostics(
  model    = model,       # the RBC pieces built above
  compiled = compiled,
  dr       = dr,
  data     = Y
)
```

Write a structured report (LLM-optimised or plain Markdown):

```{r report, eval = FALSE}
write_report(diags, file = "output/report.md")
```

---

## 8  What's next

- `vignette("solving")` — steady-state solvers, perturbation through order 5, OBC, IRFs, moments, Kalman filter variants
- `vignette("estimation")` — prior specification, mode-finding, RWMH/SMC/NUTS, cumulant-based likelihood, exact-Hessian curvature (`posterior_hessian()`), Laplace evidence, and weak-identification tools
- `vignette("hank")` — heterogeneous-agent (HANK) models: solving, mixture estimation over macro / cross-sectional-level / response channels, and consumption-equivalent welfare
- `vignette("diagnostics")` — the D0–D41 diagnostic battery, plus standalone helpers (`chain_diagnostics()`, `bk_distance()`, `kf_innovation_diagnostics()`)
- `?ramsey_model`, `?ramsey_nn1` — Ramsey optimal policy (augmented-system and Gross–Hansen (n, n+1) approximation)
- `?osr`, `?discretionary_policy` — optimal simple rules and Markov-perfect discretionary policy
- `?nash_ramsey_cooperative`, `?nash_ramsey_openloop` — Nash–Ramsey games
- `?ramsey_obc_pf`, `?ramsey_obc_pwlinear` — Ramsey with occasionally binding constraints
- `?ramsey_regime_deterministic`, `?ramsey_regime_independent` — regime-dependent Ramsey
- `?opp_sufficient_stats` — sufficient-statistics optimal policy (Barnichon & Mesters 2023)
- `?run_full_estimation` — complete argument reference for the one-call pipeline
- `?run_diagnostics` — orchestrator and individual diagnostic docs
- `?welfare_compute`, `?welfare_ce_diff` — welfare evaluation toolkit

---

## 9  References

Andreasen, M. M., Fernandez-Villaverde, J., & Rubio-Ramirez, J. F. (2018). The pruned state-space system for non-linear DSGE models. *Review of Economic Studies*, 85(1), 1-49.

Barnichon, R., & Mesters, G. (2023). A sufficient statistics approach for macro policy evaluation. *Journal of Political Economy*, 131(1), 159-197.

Binning, A. (2013). Underidentified SVAR models: A framework for combining short-run and long-run restrictions. *Norges Bank Working Paper*.

Bodenstein, M., & Guerrieri, L. (2011). The welfare-optimal degree of central bank transparency. *Journal of Money, Credit and Banking*, 43(5), 857-889.

Boehl, G. (2020). Efficient solution and computation of models with occasionally binding constraints. *Deutsche Bundesbank Discussion Paper*, 38/2020.

Gross, T., & Hansen, J. (2021). Optimal policy design in nonlinear DSGE models: An n-order accurate approximation. *Working Paper*.

Guerrieri, L., & Iacoviello, M. (2015). OccBin: A toolkit for solving dynamic models with occasionally binding constraints easily. *Journal of Monetary Economics*, 70, 22-38.

Klein, P. (2000). Using the generalized Schur form to solve a multivariate linear rational expectations model. *Journal of Economic Dynamics and Control*, 24(10), 1405-1423.

Levintal, O. (2017). Fifth-order perturbation solution of DSGE models. *Journal of Economic Dynamics and Control*, 80, 1-16.

Mutschler, W. (2015). Identification of DSGE models — The effect of higher-order approximation and pruning. *Journal of Economic Dynamics and Control*, 56, 1-33.

Schmitt-Grohé, S., & Uribe, M. (2004). Solving dynamic general equilibrium models using a second-order approximation to the policy function. *Journal of Economic Dynamics and Control*, 28(4), 755-775.

Villemot, S. (2011). Solving rational expectations models by first-order perturbation. *Dynare Working Paper Series*, 6.

---

```{r session}
sessionInfo()
```
