---
title: "Bayesian estimation with dynhr"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Bayesian estimation with dynhr}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE, purl = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>",
  eval     = FALSE   # estimation is slow; show code without running
)
## These chunks are deliberately illustrative -- they refer to YOUR data file
## and YOUR estimation output, so they cannot run standalone. `eval = FALSE`
## alone is not enough: R CMD check also RUNS the tangled code, which
## knitr::purl() emits whatever `eval` says. Hence `purl = FALSE` in every
## chunk HEADER -- it cannot be set here, because during tangling knitr never
## evaluates this chunk.
```

## Overview

This vignette walks through Bayesian estimation of a DSGE model in dynhr:

1. Specify priors from the `.mod` file
2. Build a log-posterior evaluator (Gaussian or cumulant-based)
3. Find the posterior mode
4. Draw from the posterior (RWMH, SMC, or NUTS)
5. Inspect convergence and run diagnostics (D0–D41)
6. (Optional) Evaluate Ramsey, OSR, discretionary, or Nash welfare

For a one-call pipeline skip to §7.

The running example is `nk_demo.mod` — a three-equation New Keynesian
model — together with `nk_demo_data.csv`, 200 quarters of simulated
output-growth observations.  Both ship with the package, so every chunk below
runs from a clean install with no data of your own.

---

## 1  Prior specification

Priors live in the `estimated_params` block of the `.mod` file:

```
// inst/extdata/models/nk_demo.mod
estimated_params;
// name,        distribution,  p1,    p2,    lower, upper
  kappa,        beta_pdf,      0.10,  0.05,  0.001, 0.95;   // Phillips slope
  phi_pi,       normal_pdf,    1.50,  0.25,  1.010, 4.00;   // Taylor: inflation
  phi_y,        normal_pdf,    0.25,  0.10,  0.000, 2.00;   // Taylor: output gap
  rho_r,        beta_pdf,      0.75,  0.10,  0.010, 0.99;   // rate smoothing
  rho_g,        beta_pdf,      0.85,  0.08,  0.010, 0.99;   // demand persistence
  rho_u,        beta_pdf,      0.50,  0.15,  0.010, 0.99;   // cost-push persistence
  stderr e_g,   inv_gamma_pdf, 0.30,  2.00;
  stderr e_u,   inv_gamma_pdf, 0.15,  2.00;
  stderr e_m,   inv_gamma_pdf, 0.20,  2.00;
end;
```

The optional fifth and sixth columns are hard bounds; `phi_pi`'s lower bound
of 1.01 keeps the sampler inside the determinacy region.

Extract the prior specification in R:

```{r priors, eval = TRUE, purl = FALSE}
library(dynhr)
mod    <- system.file("extdata", "models", "nk_demo.mod", package = "dynhr")
model  <- parse_mod(mod, verbose = FALSE)
priors <- prior_spec(model)   # extracts estimated_params block
print(priors)
```

`priors` is a data.frame with columns `name`, `distribution`, `p1`, `p2`,
`lower`, `upper`, `mean`, `std`.

---

## 2  Build the log-posterior

```{r posterior, eval = TRUE, purl = FALSE}
compiled <- compile_model(model, verbose = FALSE)

# Observed data (T × n_obs matrix; column names must match obs vars in model)
obs_vars <- c("ygr", "infl", "intr")
Y <- read.csv(system.file("extdata", "models", "nk_demo_data.csv",
                          package = "dynhr"))
Y <- as.matrix(Y[, obs_vars])

# Build the closure
log_post <- make_posterior(model, data = Y, prior_spec = priors,
                           obs_vars    = obs_vars,
                           compiled    = compiled,
                           me_variance = 0)

# Test at prior means
theta0 <- setNames(priors$mean, priors$name)
val    <- log_post(theta0)
cat("logpost:", val$logpost, " loglik:", val$loglik, " logprior:", val$logprior)
```

`make_posterior()` caches the compiled model structure; evaluating the
returned closure is fast (one Kalman filter pass per call).

#### Likelihood initialization for (near-)unit-root models

The Kalman filter initialization is controlled by `lik_init`, which defaults
to `"auto"` and almost never needs to be set by hand. `"auto"` picks the
cheapest *valid* initialization: it uses the stationary Lyapunov
initialization whenever that yields a valid (finite, positive semi-definite)
initial covariance, and falls back to the exact-diffuse filter only for a
genuine unit or explosive root (e.g. a random-walk trend), where the
stationary variance does not exist.

This matters for any model whose feasible region is *near* a unit root (debt,
net foreign assets, trends). A root just inside the unit circle is still
stationary, so `"auto"` keeps the fast stationary path (~10 ms/eval) rather
than the per-step exact-diffuse phase (~500–800 ms/eval) — and returns the
*exact* stationary likelihood there, not a diffuse approximation. Because
global mode-finders (CMA-ES) and MCMC sample the feasibility boundary
heavily, this is frequently the difference between a tractable and an
intractable estimation. If you genuinely want the exact-diffuse treatment for
every near-unit root, pass it explicitly:
`make_posterior(..., lik_init = "diffuse")` (`run_full_estimation()` takes the
same `lik_init` argument).

### Cumulant-based likelihood

Instead of the standard Gaussian (Kalman) likelihood, pass
`likelihood = "cumulant"` to match first through fourth moments
(mean, variance, skewness, kurtosis) of the pruned state space
against the data (Mutschler 2015).  This is particularly useful
when the data exhibit non-Gaussian features:

```{r cumulant, eval = FALSE, purl = FALSE}
log_post_cumulant <- make_posterior(model, data = Y,
  prior_spec   = priors,
  obs_vars     = obs_vars,
  compiled     = compiled,
  likelihood   = "cumulant",
  orders       = 1:4        # which cumulants to match
)
```

The underlying `make_log_posterior_cumulant()` is also callable directly,
and the individual cumulant functions `compute_third_cumulant()` and
`compute_fourth_cumulant()` can be used standalone for analysis.

### Nonlinear likelihood: the global (projection) particle filter

Every likelihood above evaluates a *perturbation* solution. When the
nonlinearity is the object of interest rather than a nuisance, the
`"global_pf"` likelihood pairs a projection solve with a bootstrap particle
filter, so the model is never linearised:

```{r globalpf, eval = FALSE, purl = FALSE}
log_post_gpf <- make_log_posterior(model, data = Y,
  prior_spec  = priors,
  obs_vars    = obs_vars,
  compiled    = compiled,
  likelihood  = "global_pf",
  n_particles = 1000L,
  poly_degree = 3L,     # Chebyshev degree of the policy approximation
  n_quad      = 5L,     # quadrature nodes for the expectation
  me_variance = 0.01    # particle filters need observation noise
)
```

The estimate is unbiased for the marginal likelihood, so `pmmh()` over this
closure is a valid pseudo-marginal sampler — provided you keep the default
`seed = NULL`, which draws a fresh particle cloud per evaluation. A fixed
seed makes the filter deterministic and breaks pseudo-marginal validity.
`global_pf_sbc()` runs the rank-uniformity certification for this posterior.

The underlying solver, `solve_global()`, is described in
`vignette("solving")`, and its accuracy diagnostics (`euler_errors()`,
`den_haan_marcet()`) are in the same place.

### Mixed-frequency observations

`obs_aggregation` declares that an observable is the *k*-period temporal
aggregate of a higher-frequency model variable, so a monthly model can be
estimated on quarterly data without leaving the monthly frequency:

```{r mixedfreq, eval = FALSE, purl = FALSE}
agg <- list(
  gdp_q = list(of = "gdp_m", type = "flow_sum", k = 3L)
)

log_post_mf <- make_posterior(model, data = Y,
                              prior_spec = priors,
                              obs_vars   = obs_vars,
                              compiled   = compiled,
                              obs_aggregation = agg)
```

The aggregated observable is `NA` in the periods where it is not observed;
the filter's existing missing-data path handles those unchanged. Four
aggregators are supported, all of them fixed linear filters of the latent:

| `type` | weights | use |
|---|---|---|
| `flow_sum` | `(1, …, 1)` | *k*-period sum of a flow |
| `flow_mean` | `(1/k, …, 1/k)` | *k*-period average |
| `stock_end` | `(1)` | end-of-period stock |
| `triangle` | `(1, 2, …, k, …, 2, 1)/k` | Mariano–Murasawa growth-rate approximation |

Implementation is fixed-weight state augmentation (Harvey 1989 §6.3;
Mariano & Murasawa 2003): the *m*−1 lags of the latent are carried as extra
states, so `ZZ` and `TT` stay constant and the filter's hot path is
untouched. The cost is *m*−1 extra states per aggregated observable. A model
with no `obs_aggregation` returns a byte-identical log-likelihood.
`mf_augment_state_space()` exposes the augmentation directly, and
`mf_expand_observations()` places a low-frequency series onto the
high-frequency grid.

---

## 3  Find the posterior mode

```{r mode, purl = FALSE}
mode <- find_mode(
  log_post_fn = log_post,
  theta_init  = theta0,
  prior_spec  = priors,
  n_iter      = 10000L,
  method      = "newrat",     # default: csminwel quasi-Newton (= Dynare mode_compute 4)
  verbose     = TRUE
)
cat("Mode logpost:", mode$logpost, "\n")
cat("Mode estimate:\n")
print(round(mode$theta_mode, 4))
```

The `mode` list contains `theta_mode`, `logpost`, `hessian` (numerical),
`V_mode` (inverse-Hessian as proposal covariance), and `se_mode`.

#### Choosing the optimizer

The default `method = "newrat"` is the csminwel quasi-Newton optimizer —
the same algorithm as Dynare's `mode_compute = 4`. It is gradient-based
(using the analytic gradient, built automatically for Gaussian/cumulant/
Whittle likelihoods) with an infeasibility-backtracking line search, which
makes it both the most robust and the cheapest choice on near-unit-root
models: it typically converges in hundreds of evaluations, where the
gradient-free CMA-ES search needs thousands and tends to stall on the
feasibility boundary. If the posterior is multimodal or the starting point is
poor, use `method = "cmaes_newrat"` (CMA-ES global search to escape poor
starts, then newrat polish) or the previous default `method = "combined"`
(CMA-ES then L-BFGS-B).

#### Hessian performance levers

The proposal-covariance Hessian (the inverse of which is `V_mode`) is the
main cost after the optimizer itself. Two knobs control it, and the
performance-optimal choice depends on model size:

- **`parallel = TRUE`** (default `FALSE`) is the single biggest lever. The
  default computes the finite-difference Hessian serially
  (\eqn{O(n_{par}^2)} likelihood evaluations); `parallel = TRUE` spreads
  those over a mirai daemon pool for a roughly \eqn{n}-fold speed-up on
  \eqn{n} cores. It is opt-in (not auto-enabled) because the daemon pool has
  setup/teardown cost and is undesirable in CI or embedded runs — but for a
  large parameter vector it is the recommended setting.

- **`use_exact_hessian = TRUE`** (default `FALSE`) swaps the finite-difference
  Hessian for the analytic `posterior_hessian`. This is *not* a universal
  speed-up: the analytic Hessian costs
  \eqn{O(n_{par}^2 \cdot n_{state}^3)} and is the faster route only for a
  small parameter vector / long sample. For a large state and many parameters
  (e.g. \eqn{n_{par} \gtrsim 60}, \eqn{n_{state} \gtrsim 70}) the
  finite-difference Hessian wins, which is why it is the default — reach for
  `parallel = TRUE` rather than `use_exact_hessian = TRUE` on big models. (A
  related option, `use_analytic_hess`, default `TRUE`, uses the analytic
  Hessian only to seed the newrat optimizer's initial curvature; set
  `dynhr_set_options(use_analytic_hess = FALSE)` to skip that expensive seed
  on a large model.)

Both can be set per call or globally via `dynhr_set_options()`.

#### Exact-Hessian curvature, model evidence, and weak identification

The analytic Hessian is exported directly as `posterior_hessian()`, with four
methods for its second-order term — `t2_method = "loop"`, `"contract_once"`,
`"hvp_solution"`, and `"adjoint_solution"`. The last two never form the
state-space second derivative; `"adjoint_solution"` is exact and
finite-difference-free (it reuses the solution adjoint). Its inverse is the
Laplace posterior covariance, and `laplace_log_marglik()` turns the mode plus
that Hessian into a Laplace estimate of the log marginal likelihood (a cheap
model-comparison score that does not need SMC):

`posterior_hessian()` differentiates the state-space likelihood itself, so it
takes the model pieces rather than the log-posterior closure: the parsed and
compiled model, the decision rule and full parameter vector at the mode, the
names of the estimated parameters, and the data.  `laplace_log_marglik()` reads
the exact Hessian off a `run_mode_finding(use_exact_hessian = TRUE)` result:

```{r hessian-curvature, eval = FALSE, purl = FALSE}
params_mode <- apply_theta_to_params(model, mode$theta_mode)
dr_mode     <- solve_perturbation(model, compiled,
                                  solve_steady(compiled, params_mode)$values,
                                  params_mode, verbose = FALSE)

H <- posterior_hessian(model, compiled, dr_mode, params_mode,
                       param_names = priors$name,
                       obs_vars = obs_vars, data = Y,
                       include_prior = TRUE, prior_spec = priors,
                       t2_method = "adjoint_solution")

## Laplace evidence: from a mode result that carries the exact Hessian
solved   <- solve_model(model)
mode_res <- run_mode_finding(solved, data = Y, obs_vars = obs_vars,
                             use_exact_hessian = TRUE, verbose = FALSE)
lm <- laplace_log_marglik(mode_res)
```

On weakly-identified or ill-conditioned posteriors, `check_hessian_conditioning()`
flags near-flat directions before you trust the mode covariance,
`fd_safe_hessian()` returns a positive-definite fallback, and `profile_ci()`
builds profile-likelihood confidence intervals that remain valid when the
quadratic (Wald) approximation does not. `run_estimation_passport()` bundles
these into a one-call pre-flight report for an estimation problem.

---

## 4  RWMH sampling

```{r rwmh, purl = FALSE}
# Scale the Hessian-based proposal
n_par      <- length(mode$theta_mode)
Sigma_prop <- mode$V_mode * (2.38^2 / n_par)

chains_rwmh <- mcmc(
  log_post_fn = log_post,
  theta0      = mode$theta_mode,
  Sigma_prop  = Sigma_prop,
  n_draws     = 100000L,
  n_warmup    = 25000L
)
print(chains_rwmh)
summary(chains_rwmh)
```

The returned `dynhr_chains` object has a `$chain` slot (\eqn{n_{\text{draws}} \times n_p}
post-warmup matrix).

### Multi-chain RWMH

Pass `n_chains > 1` to `run_full_estimation()` for automatic multi-chain
dispatch with Gelman--Rubin convergence diagnostics.

---

## 5  SMC (no mode required)

SMC is a particle filter over the parameter space.  It does not require
a mode estimate and provides a free log-marginal-likelihood estimate:

```{r smc, purl = FALSE}
chains_smc <- smc(
  log_post_fn = log_post,
  prior_spec  = priors,
  n_particles = 5000L,
  verbose     = TRUE
)
print(chains_smc)
cat("log p(Y|M):", chains_smc$log_marginal_lik, "\n")
```

When the likelihood itself is a noisy-but-unbiased particle-filter estimate
(the tempered particle filter, `likelihood = "tpf"`, or the SV-on-shocks
Rao-Blackwellised filter, `"sv_rbpf"`), `dynhr_smc2()` runs SMC^2: an outer
theta-tempering SMC around an inner particle filter, using pseudo-marginal
(PMMH-style) mutation moves so the final-stage marginal is still the exact
posterior. See `?dynhr_smc2` and `vignette("sbc-matrix")` for which
sampler/likelihood pairs are certified.

---

## 6  NUTS

NUTS adapts its step size and mass matrix automatically:

```{r nuts, purl = FALSE}
chains_nuts <- nuts(
  log_post_fn = log_post,
  theta0      = mode$theta_mode,
  n_draws     = 4000L,
  n_warmup    = 2000L
)
print(chains_nuts)
```

Check divergences with `chains_nuts$n_divergent`.  A well-tuned run
should have zero divergences.

---

## 7  One-call pipeline

All of the above in one call.  This chunk actually runs (about 7 seconds) on
the packaged `nk_demo` model and data:

```{r onecall, eval = TRUE, results = "hide", purl = FALSE}
result <- run_full_estimation(
  mod_file    = mod,                # system.file(... "nk_demo.mod" ...) from §1
  data        = Y,
  obs_vars    = obs_vars,
  output_dir  = tempdir(),
  sampler     = "smc",              # no mode required; gives log p(Y | M)
  n_particles = 500L,
  seed        = 1L,
  verbose     = FALSE
)
```

```{r onecall-summary, eval = TRUE, purl = FALSE}
summary(result$chains)
result$chains$log_marginal_lik
```

A production run would use many more particles, or the random-walk sampler
with multiple chains:

```{r onecall-rwmh, purl = FALSE}
result <- run_full_estimation(
  mod_file    = mod,
  data        = Y,
  obs_vars    = obs_vars,
  output_dir  = "output/",
  sampler     = "rwmh",
  n_draws     = 100000L,
  n_warmup    = 25000L,
  n_chains    = 4L,
  seed        = 42L,
  run_diag    = TRUE,
  verbose     = TRUE
)
print(result)
```

`output/dynhr_est_mode.rds` and `output/dynhr_est_chains.rds` are
written automatically.

### Gradient samplers, the parameter transform, and metrics

`method = "NUTS"`, `"MALA"`, `"CHEES"` and `"HMC"` are gradient-based. By default
(`transform_params = TRUE`) they sample on an unconstrained reparameterisation of
the bounded parameter vector, built from the priors. This is the recommended
default: in the constrained geometry of parameters near a unit root every
gradient sampler suffers step-size collapse, which the transform resolves while
leaving the target distribution unchanged. Set `transform_params = FALSE` to
sample in raw parameter space.

`MALA` and `NUTS` accept a `metric`: `"diagonal"` (default, identity mass),
`"hessian"` (a constant dense metric from the mode covariance — the best choice
for ill-conditioned posteriors), or `"monge"`. **The `"monge"` metric is
experimental and disabled by default**: \(G = I + \alpha^2 g g^\top\) inflates
along the gradient, so the proposal step collapses on sharp / near-unit-root
posteriors (in testing it explored a tight direction to only ~1% of its variance
where `"hessian"` recovered it fully). It errors unless you set
`dynhr_set_options(allow_monge_metric = TRUE)`, and warns when used — prefer
`"hessian"`.

---

## 8  Convergence and diagnostics

After any sampler:

```{r diag, purl = FALSE}
# Run the diagnostic battery (D0–D41)
diags <- run_diagnostics(
  model    = model,
  compiled = compiled,
  draws    = result$chains$chain,
  data     = Y
)

# Write a structured report
write_report(diags, file = "output/report.md")
```

For a quick per-parameter chain summary without the full battery, call
`chain_diagnostics()` directly on a draws matrix — it returns split-R-hat,
bulk/tail effective sample size, and Monte-Carlo standard errors:

```{r chain-diag, eval = FALSE, purl = FALSE}
cd <- chain_diagnostics(result$chains$chain)   # n_draws x n_par (or a chain list)
print(cd)                                       # rhat, ess_bulk, ess_tail, mcse
```

The diagnostic suite covers static equation-system rank (D0), pre-estimation
identification (D1–D4, D19–D26, D37), MCMC quality (D5–D7), post-estimation fit
(D8–D18), and advanced topics such as OBC identification and scenarios (D27,
D31–D32), regime-switching identification (D28), data-driven constraints (D29),
arbitrary-precision rank checks (D30), the "deep parameter" checks D33–D36, and
Kalman-innovation whiteness (D41).  Three further diagnostics — D38
(`diag_sloppiness()`), D39 (`diag_stability_map()`) and D40
(`diag_near_unit_root()`) — are called directly rather than through the
orchestrator.  See `vignette("diagnostics")` for details.

Key convergence checks:

| Metric | Good | Concern |
|---|---|---|
| R-hat | \(<1.05\) | \(>1.10\) |
| Bulk ESS | \(>400\) per chain | \(<100\) |
| Acceptance rate (RWMH) | 20–35% | \(<10\%\) or \(>50\%\) |
| Divergences (NUTS) | 0 | \(>0\) |

---

## 9  Optimal policy (optional)

dynhr ships a comprehensive optimal-policy toolkit that goes well beyond
the simple simulation-based approach:

### 9.1  Ramsey optimal commitment

Two complementary engines are available:

| Method | Function | Description |
|--------|----------|-------------|
| **Augmented system** | `ramsey_model()` | Derives FOCs symbolically, double state space (Bodenstein–Guerrieri) |
| **(n, n+1) approximation** | `ramsey_nn1()` | Modified objective without doubling the state (Gross & Hansen 2021) |

`ramsey_model()` is the gold standard; `ramsey_nn1()` is lighter weight
and scales naturally to OBC problems.

Both can be called from `run_full_estimation(..., run_ramsey = TRUE)`:

```{r ramsey, eval = FALSE, purl = FALSE}
result_ramsey <- ramsey_model(model, compiled, params,
                               planner_objective = "log(c)",
                               order = 1L, verbose = TRUE)
print(result_ramsey)
```

### 9.2  Optimal simple rules (OSR)

`osr()` optimises the coefficients of a pre-specified policy rule to
minimise the unconditional variance of target variables, using CMA-ES:

```{r osr, eval = FALSE, purl = FALSE}
osr_result <- osr(model, free_params = c(phi_pi = 1.5, phi_y = 0.5),
                   loss_vars = c("pi", "y_gap"),
                   loss_weights = c(pi = 1, y_gap = 0.5))
```

### 9.3  Discretionary policy

`discretionary_policy()` computes the Markov-perfect equilibrium under
discretion via Riccati fixed-point iteration:

```{r disc, eval = FALSE, purl = FALSE}
disc_result <- discretionary_policy(model,
                                     policy_instrument = "i",
                                     loss_vars = c("pi", "y_gap"))
```

### 9.4  Nash–Ramsey games

- `nash_ramsey_cooperative()` — multiple policymakers with a weighted-sum objective
- `nash_ramsey_openloop()` — open-loop Nash equilibrium (welfare matrix)

### 9.5  Ramsey with OBC

- `ramsey_obc_pf()` — perfect-foresight Ramsey under binding constraints
- `ramsey_obc_pwlinear()` — piecewise-linear Ramsey commitment
- `ramsey_obc_welfare_cost()` — compare welfare across OBC regimes

### 9.6  Regime-dependent Ramsey

- `ramsey_regime_deterministic()` — deterministic regime path with per-regime objectives
- `ramsey_regime_independent()` — Markov-switching Ramsey with ergodic welfare

### 9.7  Sufficient-statistics optimal policy

`opp_sufficient_stats()` implements the Barnichon & Mesters (2023)
approach: given IRFs to a policy instrument, compute the optimal
instrument deviation that minimises a quadratic loss subject to the
IRF-based transmission:

```{r opp, eval = FALSE, purl = FALSE}
opp_result <- opp_sufficient_stats(model,
                                    target_vars = c("pi", "y_gap"),
                                    loss_weights = c(pi = 1, y_gap = 0.5))
```

### 9.8  Welfare toolkit

- `welfare_compute()` — simulation-based welfare evaluation
- `welfare_ce_diff()` — consumption-equivalent welfare difference
- `welfare_cost_of_rule()` — cost of a given policy rule
- `welfare_decompose()` — welfare decomposition into steady-state, variance, and higher-order components
- `conditional_welfare()` — conditional welfare under alternative policies
- `evaluate_planner_objective()` — evaluate the planner's objective at a given solution

`run_diagnostics()` will consume Ramsey output in D18 welfare plausibility when
available; otherwise D18 remains informational.

---

## 10  Saving and loading chains

```{r save, purl = FALSE}
saveRDS(result$chains, "output/chains.rds")
chains <- readRDS("output/chains.rds")

# Convert to coda::mcmc.list for standard diagnostics
chain_mat <- chains$chain
# Use coda::as.mcmc(chain_mat) if coda is installed
```

### Checkpointing and resuming a long run

`saveRDS()` preserves the draws but not the sampler: restarting from it
begins a fresh chain at a fresh adaptation state. For runs long enough to
outlive a session or a scheduler slot, `mcmc()` checkpoints instead:

```{r checkpoint, eval = FALSE, purl = FALSE}
# First leg: writes a state pack every `flush_every` draws
res1 <- mcmc(log_post, theta0, Sigma_prop,
             n_draws = 5000, n_warmup = 2000,
             checkpoint_dir = "output/chain1",
             flush_every    = 1000)

# Later, in a new session: continue where it stopped.
# `n_warmup` stays as it was; grow `n_draws` to the new TOTAL.
res2 <- mcmc(log_post, theta0, Sigma_prop,
             n_draws = 20000, n_warmup = 2000,
             checkpoint_dir = "output/chain1",
             resume = TRUE)
```

The state pack carries the position, the log-posterior, the adaptation
state and `.Random.seed`, so a resumed chain is statistically identical to
the uninterrupted run. It is checksummed and written atomically, and
`mcmc_chain_restore()` refuses a tampered pack or one saved mid-adaptation
(pass `allow_mid_adaptation = TRUE` only if you know why you want that).
`mcmc_chain_state()`, `mcmc_chain_save()`, `mcmc_chain_restore()` and
`mcmc_chain_extend()` are the sampler-agnostic primitives underneath, for
building the same behaviour into a bespoke sampler.

---

## 11  Delayed acceptance

When a model has both a cheap approximate likelihood and an expensive exact
one, `screen_fn` runs the cheap one first and only evaluates the expensive
one on proposals that survive:

```{r delayed, eval = FALSE, purl = FALSE}
cheap <- make_posterior(model, data = Y, prior_spec = priors,
                        obs_vars = obs_vars, compiled = compiled)

exact <- make_log_posterior(model, data = Y, prior_spec = priors,
                            obs_vars = obs_vars, compiled = compiled,
                            likelihood = "tpf", me_variance = 0.01)

res <- mcmc(exact, theta0, Sigma_prop,
            n_draws = 10000, n_warmup = 2000,
            screen_fn = cheap)
```

The two-stage acceptance ratio corrects for the screening, so the invariant
distribution is still the exact posterior. The saving is real only when the
screen rejects most proposals and is much cheaper than the target; a screen
that disagrees with the target wastes both evaluations.

---

## 12  Method of moments (GMM / SMM)

`method_of_moments()` estimates by matching model-implied moments to sample
moments rather than by evaluating a likelihood — useful when the likelihood
is unavailable, unreliable, or simply not the estimator you want to defend:

```{r mom, eval = FALSE, purl = FALSE}
mm <- method_of_moments(model, data = Y, obs_vars = obs_vars,
                        moments   = list(orders = 1:2, lags = 0:4),
                        method    = "gmm",       # or "smm"
                        weighting = "optimal",   # identity/newey_west/diagonal
                        two_step  = TRUE)
```

`method = "gmm"` uses analytic model moments; `"smm"` simulates them
(`n_sim` replications of the sample, common random numbers across
parameter draws). `weighting = "optimal"` runs the standard two-step
procedure, with `"newey_west"` available for serially correlated moment
conditions. The moment Jacobian is analytic where dynhr has one and falls
back to finite differences otherwise (`jacobian = "auto"`).

---

## See also

- `?run_full_estimation` -- complete argument reference
- `?prior_spec` -- prior distribution families supported
- `?make_posterior` -- posterior evaluator options
- `?find_mode` -- optimiser methods (CMA-ES, Nelder-Mead, JADE)
- `?run_mode_finding`, `?run_posterior_estimation` -- stepwise estimation API
- `?make_log_posterior_cumulant` -- cumulant-based posterior
- `?make_log_posterior_global_pf`, `?global_pf_sbc` -- projection + particle filter
- `?kalman_filter` (`obs_aggregation`), `?mf_augment_state_space` -- mixed frequency
- `?mcmc` (`checkpoint_dir`, `resume`, `screen_fn`) -- checkpointing, delayed acceptance
- `?mcmc_chain_state`, `?mcmc_chain_save`, `?mcmc_chain_restore` -- state packs
- `?method_of_moments` -- GMM / SMM
- `?forecast_backtest` -- out-of-sample scoring (see `vignette("diagnostics")`)
- `?ramsey_model`, `?ramsey_nn1` -- Ramsey optimal policy
- `?osr`, `?discretionary_policy` -- OSR and discretionary policy
- `?nash_ramsey_cooperative`, `?nash_ramsey_openloop` -- Nash games
- `?ramsey_obc_pf`, `?ramsey_obc_pwlinear` -- Ramsey with OBC
- `?ramsey_regime_deterministic`, `?ramsey_regime_independent` -- regime-dependent
- `?opp_sufficient_stats` -- sufficient-statistics OPP
- `?welfare_compute`, `?welfare_ce_diff` -- welfare toolkit
- `vignette("diagnostics")` -- the D0–D41 diagnostic battery

---

## References

Barnichon, R., & Mesters, G. (2023). A sufficient statistics approach for macro policy evaluation. *Journal of Political Economy*, 131(1), 159-197.

Bodenstein, M., & Guerrieri, L. (2011). The welfare-optimal degree of central bank transparency. *Journal of Money, Credit and Banking*, 43(5), 857-889.

Chopin, N. (2002). A sequential particle filter method for static models. *Biometrika*, 89(3), 539-552.

Gelman, A., Carlin, J. B., Stern, H. S., & Rubin, D. B. (2013). *Bayesian Data Analysis* (3rd ed.). Chapman & Hall/CRC.

Gross, T., & Hansen, J. (2021). Optimal policy design in nonlinear DSGE models: An n-order accurate approximation. *Working Paper*.

Hansen, N. (2016). The CMA evolution strategy: A tutorial. *arXiv:1604.00772*.

Herbst, E. P., & Schorfheide, F. (2015). *Bayesian Estimation of DSGE Models*. Princeton University Press.

Hoffman, M. D., & Gelman, A. (2014). The No-U-Turn sampler: Adaptively setting path lengths in Hamiltonian Monte Carlo. *Journal of Machine Learning Research*, 15(1), 1593-1623.

Mariano, R. S., & Murasawa, Y. (2003). A new coincident index of business cycles based on monthly and quarterly series. *Journal of Applied Econometrics*, 18(4), 427-443.

Mutschler, W. (2015). Identification of DSGE models — The effect of higher-order approximation and pruning. *Journal of Economic Dynamics and Control*, 56, 1-33.

Schmitt-Grohé, S., & Uribe, M. (2004). Solving dynamic general equilibrium models using a second-order approximation to the policy function. *Journal of Economic Dynamics and Control*, 28(4), 755-775.

Söderlind, P. (1999). Solution and estimation of RE macromodels with optimal policy. *European Economic Review*, 43(4-6), 813-823.
