---
title: "Bayesian estimation with dynhr"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Bayesian estimation with dynhr}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE, purl = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>",
  eval     = FALSE   # estimation is slow; show code without running
)
## These chunks are deliberately illustrative -- they refer to YOUR data file
## and YOUR estimation output, so they cannot run standalone. `eval = FALSE`
## alone is not enough: R CMD check also RUNS the tangled code, which
## knitr::purl() emits whatever `eval` says. Hence `purl = FALSE` in every
## chunk HEADER -- it cannot be set here, because during tangling knitr never
## evaluates this chunk.
```

## Overview

This vignette walks through Bayesian estimation of a DSGE model in dynhr:

1. Specify priors from the `.mod` file
2. Build a log-posterior evaluator (Gaussian or cumulant-based)
3. Find the posterior mode
4. Draw from the posterior (RWMH, SMC, or NUTS)
5. Inspect convergence and run diagnostics (D0–D41)
6. (Optional) Evaluate Ramsey, OSR, discretionary, or Nash welfare

For a one-call pipeline skip to §7; for specs, run records, reruns and
reproducibility checks see §13.

The running example is `nk_demo.mod` — a three-equation New Keynesian
model — together with `nk_demo_data.csv`, 200 quarters of simulated
output-growth observations.  Both ship with the package, so every chunk below
runs from a clean install with no data of your own.

---

## 1  Prior specification

Priors live in the `estimated_params` block of the `.mod` file:

```
// inst/extdata/models/nk_demo.mod
estimated_params;
// name, init, lower, upper, distribution, p1 (mean), p2 (std)
  kappa, 0.10, 0.001, 0.95, beta_pdf, 0.10, 0.05;   // Phillips slope
  phi_pi, 1.50, 1.010, 4.00, normal_pdf, 1.50, 0.25;   // Taylor: inflation
  phi_y, 0.25, 0.000, 2.00, normal_pdf, 0.25, 0.10;   // Taylor: output gap
  rho_r, 0.75, 0.010, 0.99, beta_pdf, 0.75, 0.10;   // rate smoothing
  rho_g, 0.85, 0.010, 0.99, beta_pdf, 0.85, 0.08;   // demand persistence
  rho_u, 0.50, 0.010, 0.99, beta_pdf, 0.50, 0.15;   // cost-push persistence
  stderr e_g,   inv_gamma_pdf, 0.30,  2.00;
  stderr e_u,   inv_gamma_pdf, 0.15,  2.00;
  stderr e_m,   inv_gamma_pdf, 0.20,  2.00;
end;
```

The third and fourth columns are hard bounds; `phi_pi`'s lower bound
of 1.01 keeps the sampler inside the determinacy region.

Extract the prior specification in R:

```{r priors, eval = TRUE, purl = FALSE}
library(dynhr)
mod    <- system.file("extdata", "models", "nk_demo.mod", package = "dynhr")
model  <- parse_mod(mod, verbose = FALSE)
priors <- prior_spec(model)   # extracts estimated_params block
print(priors)
```

`priors` is a data.frame with columns `name`, `distribution`, `p1`, `p2`,
`lower`, `upper`, `mean`, `std`, `p3`, `p4` and `init`.

`p1`/`p2` are Dynare's hyperparameters: the prior **mean** and **standard
deviation** for every shape. For `uniform_pdf` that means
`name, uniform_pdf, m, s;` is uniform on `[m - sqrt(3) s, m + sqrt(3) s]`,
exactly as in Dynare; a uniform given by its bounds is written with empty
`p1`/`p2` and the bounds in the third and fourth prior parameters,
`name, uniform_pdf, , , lb, ub;`. In the extracted spec a uniform carries its
mean/sd in `p1`/`p2` and its support in `p3`/`p4` (Dynare's layout), and a
row whose `p1`/`p2` look like bounds read as mean/sd triggers a
`dynhr_warning_uniform_parameterisation` warning.

---

## 2  Build the log-posterior

```{r posterior, eval = TRUE, purl = FALSE}
compiled <- compile_model(model, verbose = FALSE)

# Observed data (T × n_obs matrix; column names must match obs vars in model)
obs_vars <- c("ygr", "infl", "intr")
Y <- read.csv(system.file("extdata", "models", "nk_demo_data.csv",
                          package = "dynhr"))
Y <- as.matrix(Y[, obs_vars])

# Build the closure
log_post <- make_posterior(model, data = Y, prior_spec = priors,
                           obs_vars    = obs_vars,
                           compiled    = compiled,
                           me_variance = 0)

# Test at prior means
theta0 <- setNames(priors$mean, priors$name)
val    <- log_post(theta0)
cat("logpost:", val$logpost, " loglik:", val$loglik, " logprior:", val$logprior)
```

`make_posterior()` caches the compiled model structure; evaluating the
returned closure is fast (one Kalman filter pass per call).

#### Likelihood initialization for (near-)unit-root models

The Kalman filter initialization is controlled by `lik_init`, which defaults
to `"auto"` and almost never needs to be set by hand. `"auto"` picks the
cheapest *valid* initialization: it uses the stationary Lyapunov
initialization whenever that yields a valid (finite, positive semi-definite)
initial covariance, and falls back to the exact-diffuse filter only for a
genuine unit or explosive root (e.g. a random-walk trend), where the
stationary variance does not exist.

This matters for any model whose feasible region is *near* a unit root (debt,
net foreign assets, trends). A root just inside the unit circle is still
stationary, so `"auto"` keeps the fast stationary path (~10 ms/eval) rather
than the per-step exact-diffuse phase (~500–800 ms/eval) — and returns the
*exact* stationary likelihood there, not a diffuse approximation. Because
global mode-finders (CMA-ES) and MCMC sample the feasibility boundary
heavily, this is frequently the difference between a tractable and an
intractable estimation. If you genuinely want the exact-diffuse treatment for
every near-unit root, pass it explicitly:
`make_posterior(..., lik_init = "diffuse")` (`run_full_estimation()` takes the
same `lik_init` argument).

### Cumulant-based likelihood

Instead of the standard Gaussian (Kalman) likelihood, pass
`likelihood = "cumulant"` to match first through fourth moments
(mean, variance, skewness, kurtosis) of the pruned state space
against the data (Mutschler 2015).  This is particularly useful
when the data exhibit non-Gaussian features:

```{r cumulant, eval = FALSE, purl = FALSE}
log_post_cumulant <- make_posterior(model, data = Y,
  prior_spec   = priors,
  obs_vars     = obs_vars,
  compiled     = compiled,
  likelihood   = "cumulant",
  cumulant_orders = 1:4     # which cumulants to match
)
```

The underlying `make_log_posterior_cumulant()` is also callable directly,
and the individual cumulant functions `compute_third_cumulant()` and
`compute_fourth_cumulant()` can be used standalone for analysis.

### Nonlinear likelihood: the global (projection) particle filter

Every likelihood above evaluates a *perturbation* solution. When the
nonlinearity is the object of interest rather than a nuisance, the
`"global_pf"` likelihood pairs a projection solve with a bootstrap particle
filter, so the model is never linearised:

```{r globalpf, eval = FALSE, purl = FALSE}
log_post_gpf <- make_log_posterior(model, data = Y,
  prior_spec  = priors,
  obs_vars    = obs_vars,
  compiled    = compiled,
  likelihood  = "global_pf",
  n_particles = 1000L,
  poly_degree = 3L,     # Chebyshev degree of the policy approximation
  n_quad      = 5L,     # quadrature nodes for the expectation
  me_variance = 0.01    # particle filters need observation noise
)
```

The estimate is unbiased for the marginal likelihood, so `pmmh()` over this
closure is a valid pseudo-marginal sampler — provided you keep the default
`seed = NULL`, which draws a fresh particle cloud per evaluation. A fixed
seed makes the filter deterministic and breaks pseudo-marginal validity.
`global_pf_sbc()` runs the rank-uniformity certification for this posterior.

The underlying solver, `solve_global()`, is described in
`vignette("solving")`, and its accuracy diagnostics (`euler_errors()`,
`den_haan_marcet()`) are in the same place.

### Mixed-frequency observations

`obs_aggregation` declares that an observable is the *k*-period temporal
aggregate of a higher-frequency model variable, so a monthly model can be
estimated on quarterly data without leaving the monthly frequency:

```{r mixedfreq, eval = FALSE, purl = FALSE}
# The aggregation spec lives on the model; the Gaussian (Kalman) likelihood
# reads it from model$obs_aggregation. obs_vars must include "gdp_q".
model$obs_aggregation <- list(
  gdp_q = list(of = "gdp_m", type = "flow_sum", k = 3L)
)

log_post_mf <- make_posterior(model, data = Y,
                              prior_spec = priors,
                              obs_vars   = obs_vars,
                              compiled   = compiled)
```

`kalman_filter()` also takes the spec directly as its `obs_aggregation`
argument. The other likelihoods refuse a model that carries one rather than
silently scoring the aggregate as a contemporaneous observable.

The aggregated observable is `NA` in the periods where it is not observed;
the filter's existing missing-data path handles those unchanged. Four
aggregators are supported, all of them fixed linear filters of the latent:

| `type` | weights | use |
|---|---|---|
| `flow_sum` | `(1, …, 1)` | *k*-period sum of a flow |
| `flow_mean` | `(1/k, …, 1/k)` | *k*-period average |
| `stock_end` | `(1)` | end-of-period stock |
| `triangle` | `(1, 2, …, k, …, 2, 1)/k` | Mariano–Murasawa growth-rate approximation |

Implementation is fixed-weight state augmentation (Harvey 1989 §6.3;
Mariano & Murasawa 2003): the *m*−1 lags of the latent are carried as extra
states, so `ZZ` and `TT` stay constant and the filter's hot path is
untouched. The cost is *m*−1 extra states per aggregated observable. A model
with no `obs_aggregation` returns a byte-identical log-likelihood.
`mf_augment_state_space()` exposes the augmentation directly, and
`mf_expand_observations()` places a low-frequency series onto the
high-frequency grid.

---

## 3  Find the posterior mode

```{r mode, purl = FALSE}
mode <- find_mode(
  log_post_fn = log_post,
  theta_init  = theta0,
  prior_spec  = priors,
  n_iter      = 10000L,
  method      = "newrat",     # default: csminwel quasi-Newton (= Dynare mode_compute 4)
  verbose     = TRUE
)
cat("Mode logpost:", mode$logpost, "\n")
cat("Mode estimate:\n")
print(round(mode$theta_mode, 4))
```

The `mode` list contains `theta_mode`, `logpost`, `convergence`,
`iterations` and `method`. `find_mode()` optimises a bare closure only; for
the mode together with a proposal covariance (`Sigma_prop`, the scaled
inverse Hessian at the mode, and `V_mode`), the exact gradient and a run
record, use `run_mode_finding()`:

```{r mode-full, purl = FALSE}
mode_res <- run_mode_finding(solve_model(model), data = Y,
                             obs_vars = obs_vars, verbose = FALSE)
mode_res$theta_mode
mode_res$Sigma_prop      # RWMH proposal covariance, used in §4
```

#### Choosing the optimizer

The default `method = "newrat"` is the csminwel quasi-Newton optimizer —
the same algorithm as Dynare's `mode_compute = 4`. It is gradient-based with
an infeasibility-backtracking line search, which makes it both the most
robust and the cheapest choice on near-unit-root models: it typically
converges in hundreds of evaluations, where the gradient-free CMA-ES search
needs thousands and tends to stall on the feasibility boundary.
`run_mode_finding()` builds the exact analytic gradient automatically for
Gaussian, cumulant and Whittle likelihoods (`grad_method = "auto"`); on a
bare closure `find_mode()` uses csminwel's numerical gradient. If the
posterior is multimodal or the starting point is poor, use
`method = "cmaes_newrat"` (CMA-ES global search to escape poor starts, then
newrat polish) or the previous default `method = "combined"` (CMA-ES then
L-BFGS-B).

#### Hessian performance levers

The proposal-covariance Hessian (the inverse of which is `V_mode`) is the
main cost of `run_mode_finding()` after the optimizer itself. When the mode
stage has an exact gradient (Gaussian, cumulant or Whittle likelihood, no
OBC, `grad_method` not `"hybrid"`), that Hessian is a central difference of
the gradient: \eqn{2 n_{par}} gradient calls (about 0.7 s for Smets–Wouters'
36 parameters and 5 s for a 68-parameter model, installed build). Otherwise
it falls back to the \eqn{2 n_{par}(n_{par}+1)}-evaluation log-posterior
stencil. Two knobs control it:

- **`parallel = TRUE`** (default `FALSE`) spreads those calls over a mirai
  daemon pool for a roughly \eqn{n}-fold speed-up on \eqn{n} cores. It is
  opt-in (not auto-enabled) because the daemon pool has setup/teardown cost
  and is undesirable in CI or embedded runs — but for a large parameter
  vector it is the recommended setting.

- **`use_exact_hessian = TRUE`** (default `FALSE`) swaps the difference
  Hessian for the analytic `posterior_hessian`. This is *not* a universal
  speed-up: the analytic Hessian costs
  \eqn{O(n_{par}^2 \cdot n_{state}^3)} and is the faster route only for a
  small parameter vector / long sample. For a large state and many parameters
  (e.g. \eqn{n_{par} \gtrsim 60}, \eqn{n_{state} \gtrsim 70}) the difference
  Hessian wins, which is why it is the default — reach for `parallel = TRUE`
  rather than `use_exact_hessian = TRUE` on big models. (Separately,
  `mode_options = list(use_analytic_hess = FALSE)` skips the analytic
  Hessian that otherwise seeds the newrat optimizer's initial curvature,
  which is itself expensive on a large model.)

`parallel` and `use_exact_hessian` can be set per call; `use_exact_hessian`
also globally via `dynhr_set_options()`.

#### Exact-Hessian curvature, model evidence, and weak identification

The analytic Hessian is exported directly as `posterior_hessian()`, with four
methods for its second-order term — `t2_method = "loop"`, `"contract_once"`,
`"hvp_solution"`, and `"adjoint_solution"`. The last two never form the
state-space second derivative; `"adjoint_solution"` is exact and
finite-difference-free (it reuses the solution adjoint). Its inverse is the
Laplace posterior covariance, and `laplace_log_marglik()` turns the mode plus
that Hessian into a Laplace estimate of the log marginal likelihood (a cheap
model-comparison score that does not need SMC):

`posterior_hessian()` differentiates the state-space likelihood itself, so it
takes the model pieces rather than the log-posterior closure: the parsed and
compiled model, the decision rule and full parameter vector at the mode, the
names of the estimated parameters, and the data.  `laplace_log_marglik()` reads
the exact Hessian off a `run_mode_finding(use_exact_hessian = TRUE)` result:

```{r hessian-curvature, eval = FALSE, purl = FALSE}
params_mode <- apply_theta_to_params(model, mode$theta_mode)
dr_mode     <- solve_perturbation(model, compiled,
                                  solve_steady(compiled, params_mode)$values,
                                  params_mode, verbose = FALSE)

H <- posterior_hessian(model, compiled, dr_mode, params_mode,
                       param_names = priors$name,
                       obs_vars = obs_vars, data = Y,
                       include_prior = TRUE, prior_spec = priors,
                       t2_method = "adjoint_solution")

## Laplace evidence: from a mode result that carries the exact Hessian
solved   <- solve_model(model)
mode_res <- run_mode_finding(solved, data = Y, obs_vars = obs_vars,
                             use_exact_hessian = TRUE, verbose = FALSE)
lm <- laplace_log_marglik(mode_res)
```

On weakly-identified or ill-conditioned posteriors, `check_hessian_conditioning()`
flags near-flat directions before you trust the mode covariance,
`fd_safe_hessian()` returns a positive-definite fallback, and `profile_ci()`
builds profile-likelihood confidence intervals that remain valid when the
quadratic (Wald) approximation does not. `run_estimation_passport()` bundles
these into a one-call pre-flight report for an estimation problem.

---

## 4  RWMH sampling

```{r rwmh, purl = FALSE}
# The proposal from run_mode_finding() (§3), already scaled by 2.38^2 / n_par
Sigma_prop <- mode_res$Sigma_prop

chains_rwmh <- dynhr_mcmc(
  log_post_fn = log_post,
  theta0      = mode_res$theta_mode,
  Sigma_prop  = Sigma_prop,
  n_draws     = 100000L,
  n_warmup    = 25000L
)
print(chains_rwmh)
summary(chains_rwmh)
```

The returned `dynhr_chains` object has a `$chain` slot (\eqn{n_{\text{draws}} \times n_p}
post-warmup matrix).

### Multi-chain RWMH

Pass `n_chains > 1` to `run_full_estimation()` for automatic multi-chain
dispatch with Gelman--Rubin convergence diagnostics.

---

## 5  SMC (no mode required)

SMC is a particle filter over the parameter space.  It does not require
a mode estimate and provides a free log-marginal-likelihood estimate:

```{r smc, purl = FALSE}
chains_smc <- smc(
  log_post_fn = log_post,
  prior_spec  = priors,
  n_particles = 5000L,
  verbose     = TRUE
)
print(chains_smc)
cat("log p(Y|M):", chains_smc$log_marginal_lik, "\n")
```

When the likelihood itself is a noisy-but-unbiased particle-filter estimate
(the tempered particle filter, `likelihood = "tpf"`, or the SV-on-shocks
Rao-Blackwellised filter, `"sv_rbpf"`), `dynhr_smc2()` runs SMC^2: an outer
theta-tempering SMC around an inner particle filter, using pseudo-marginal
(PMMH-style) mutation moves so the final-stage marginal is still the exact
posterior. See `?dynhr_smc2` and `vignette("sbc-matrix")` for which
sampler/likelihood pairs are certified.

---

## 6  NUTS

NUTS adapts its step size and mass matrix automatically. It needs the
gradient of the log-posterior: pass the exact one from
`make_posterior_grad()`, whose default `grad_method = "auto"` resolves to the
finite-difference-free `"adjoint_solution"` for a Gaussian likelihood. Called
without `grad_fn`, `nuts()` falls back to forward finite differences of
`log_post`, which costs one log-posterior evaluation per parameter per
gradient and is only approximate:

```{r nuts, purl = FALSE}
grad <- make_posterior_grad(model, data = Y, prior_spec = priors,
                            obs_vars = obs_vars, compiled = compiled)
chains_nuts <- nuts(
  log_post_fn = log_post,
  theta0      = mode$theta_mode,
  grad_fn     = grad,
  n_draws     = 4000L,
  n_warmup    = 2000L
)
print(chains_nuts)
```

Check divergences with `chains_nuts$n_divergent`.  A well-tuned run
should have zero divergences. `nuts()` samples in raw parameter space; the
runner (`run_posterior_estimation()`, `run_full_estimation()`,
`run_estimation()`) adds the unconstrained reparameterisation and, by
default, the exact gradient for you (`analytic_grad = TRUE`, see §7).

---

## 7  One-call pipeline

All of the above in one call.  This chunk actually runs (about 7 seconds) on
the packaged `nk_demo` model and data:

```{r onecall, eval = TRUE, results = "hide", purl = FALSE}
result <- run_full_estimation(
  mod_file    = mod,                # system.file(... "nk_demo.mod" ...) from §1
  data        = Y,
  obs_vars    = obs_vars,
  output_dir  = tempdir(),
  sampler     = "smc",              # no mode required; gives log p(Y | M)
  n_particles = 500L,
  seed        = 1L,
  verbose     = FALSE
)
```

```{r onecall-summary, eval = TRUE, purl = FALSE}
summary(result$chains)
result$chains$log_marginal_lik
```

The draws also convert to the formats of the wider R ecosystem:
`posterior::as_draws()` (chains kept apart; a weighted SMC cloud keeps its
weights), `coda::as.mcmc.list()`, `as.data.frame()`, `coef()` (posterior
mean) and `vcov()` — so `posterior::summarise_draws()` and the
`bayesplot::mcmc_*()` plots take dynhr output directly. SMC particles are
not a Markov chain, so ask for location/scale summaries only (R-hat and ESS
are for MCMC chains):

```{r onecall-posterior, eval = requireNamespace("posterior", quietly = TRUE), purl = FALSE}
draws <- posterior::as_draws(result$chains)
posterior::summarise_draws(draws, "mean", "sd", "median")
coef(result$chains)   # posterior mean
```

A production run would use many more particles, or the random-walk sampler
with multiple chains:

```{r onecall-rwmh, purl = FALSE}
result <- run_full_estimation(
  mod_file    = mod,
  data        = Y,
  obs_vars    = obs_vars,
  output_dir  = "output/",
  sampler     = "rwmh",
  n_draws     = 100000L,
  n_warmup    = 25000L,
  n_chains    = 4L,
  seed        = 42L,
  run_diag    = TRUE,
  verbose     = TRUE
)
print(result)
```

`output/dynhr_est_mode.rds` and `output/dynhr_est_chains.rds` are
written automatically.

### Gradient samplers, the parameter transform, and metrics

`method = "NUTS"`, `"MALA"`, `"CHEES"` and `"HMC"` are gradient-based. By
default (`analytic_grad = TRUE`, also the `sampler_spec()` default) the
runners give them the exact gradient: `grad_method = "auto"` (the default of
the `grad_method` option) resolves to `"adjoint_solution"` for the Gaussian
likelihood and to an analytic method for the Whittle, cumulant and pruned
likelihoods (see `?make_posterior_grad`). A likelihood without an analytic
gradient (the OBC filters, `"tpf"`, `"pskf"`, ...) keeps the numerical one
without a warning; `analytic_grad = FALSE` always finite-differences the
log-posterior. By default
(`transform_params = TRUE`) they sample on an unconstrained reparameterisation of
the bounded parameter vector, built from the priors. This is the recommended
default, and it leaves the target distribution unchanged. In the constrained
geometry of parameters near a unit root the gradient samplers run into the
bounds. On a 68-parameter New Zealand DSGE (20 simulated observables,
T = 132), NUTS with the default adapted diagonal metric (300 warmup + 300 draws) gave:

- without the transform: 246 of 300 transitions divergent, minimum bulk ESS 1.3;
- with the transform: 0 divergences, minimum bulk ESS 57.

Set `transform_params = FALSE` to sample in raw parameter space.

On that model the choice of sampler also matters. The table compares each
sampler's worst-mixing parameter (minimum bulk ESS). All runs start at the posterior
mode. The gradient samplers use the transform. RWMH runs in raw parameter
space with a diagonal proposal scaled from the curvature at the mode, after
5,000 burn-in draws.

| Sampler | Draws | Min bulk ESS | Min ESS per minute |
|---|---|---|---|
| RWMH | 20000 | 2.2 | 1.83 |
| ChEES | 300 | 132 | 1.58 |
| NUTS, `"lowrank"` | 300 | 88 | 0.51 |
| NUTS, `"diagonal"` | 300 | 57 | 0.50 |
| HMC | 300 | 43 | 0.32 |
| NUTS, `"fisher_diag"` | 300 | 15 | 0.13 |

RWMH's per-minute figure comes from cheap draws of a chain that has not
converged: 20,000 draws gave a minimum ESS of 2.2. ChEES gave the best
minimum ESS. Those timings used the gradient that was then the default,
at about 0.26 s per gradient. The default `grad_method = "auto"` is now
exact and about 5–7× cheaper on this model. The gradient samplers'
ESS per minute should improve by roughly that factor, which would put ChEES
and NUTS well ahead of RWMH (not re-measured). On large models, do not choose RWMH
because its draws are cheap. Compare ESS on the worst-mixing parameter.

`NUTS`, `HMC` and `MALA` accept a `metric`. The metric is the mass matrix
\(M\), and the momentum is drawn as \(p \sim N(0, M)\). dynhr follows Stan's
convention: what is adapted or supplied is the *inverse* mass \(M^{-1}\), and
it should approximate the posterior covariance, so each direction is rescaled
to unit posterior scale.

Before 0.9.3.50 the adapted diagonal set the mass itself to the variance.
That squared the conditioning instead of removing it. Draws were still valid
at convergence, but finite NUTS/HMC/ChEES runs could mix badly.

Adapted during warmup:

- `"diagonal"` (default). NUTS starts from the inverse-Hessian diagonal at
  the mode, and HMC from the identity. During warmup both set a diagonal
  \(M^{-1}\) to the warmup variance of each parameter. NUTS updates it in
  expanding windows; HMC updates it once, at 70% of warmup. MALA has no
  adaptation and uses the identity.
- `"warmup_dense"` (NUTS, HMC). Diagonal windows first, then one dense
  Ledoit–Wolf-shrunk \(M^{-1}\) from the final window's draws.
- `"fisher_diag"` (NUTS, opt-in). A diagonal
  \(M^{-1} = \sqrt{\mathrm{var}(\theta)/\mathrm{var}(\nabla \log p)}\) from each
  window's draws *and* gradients. This minimises the sample Fisher divergence
  to a standard normal (Seyboldt, Carlson & Carpenter 2026).
  - It is exact after one window when parameters are independent Gaussians,
    so it helps most with short warmups: about 1.4× ESS per gradient in dynhr's
    tests.
  - On strongly correlated posteriors it can do worse than `"diagonal"`.
  - It costs one extra gradient per warmup iteration.
- `"lowrank"` (NUTS, opt-in). A low-rank-plus-diagonal \(M^{-1}\) estimated
  from draws and gradients (the same paper's Algorithm 1), using Lao's (2026)
  warmup schedule. It captures a few dominant correlated directions at
  \(O(dk)\) cost per step.

Fixed through warmup and sampling:

- `"hessian"` (NUTS, MALA). A constant dense metric,
  \(M^{-1} = \Sigma_{\text{prop}}\), the mode covariance. This is the best
  choice for ill-conditioned, correlated posteriors.
- `"whittle_fim"` (NUTS). A dense metric from the Whittle Fisher information
  at the mode. It is experimental and disabled by default:
  `dynhr_set_options(allow_whittle_fim_metric = TRUE)`.
- `"monge"` (MALA). A position-dependent metric. **It is experimental and
  disabled by default.** \(G = I + \alpha^2 g g^\top\) inflates along the
  gradient, so the proposal step collapses on sharp or near-unit-root
  posteriors. In testing it explored a tight direction to only ~1% of its
  variance, where `"hessian"` recovered it fully. It errors unless you set
  `dynhr_set_options(allow_monge_metric = TRUE)`, and warns when used. Prefer
  `"hessian"`.

The parallel multi-chain NUTS path (`parallel = TRUE`) takes the adapted
metrics (`"diagonal"`, `"warmup_dense"`, `"fisher_diag"`, `"lowrank"`). A
fixed metric (`"hessian"`, `"whittle_fim"`) is not sent to the daemons,
which run `"diagonal"` instead; `validate_spec()` warns
(`dynhr_warning_metric_ignored`).

`HMC` runs a fixed number of leapfrog steps `L`. It also jitters the step size
by ±20% each iteration (`step_jitter`), so a fixed trajectory length cannot
lock onto a periodic orbit of the target.

---

## 8  Convergence and diagnostics

After any sampler:

```{r diag, purl = FALSE}
# Run the diagnostic battery (D0–D41)
diags <- run_diagnostics(
  model    = model,
  compiled = compiled,
  draws    = result$chains$chain,
  data     = Y
)

# Write a structured report
write_report(diags, file = "output/report.md")
```

For a quick per-parameter chain summary without the full battery, call
`chain_diagnostics()` directly on a draws matrix — it returns split-R-hat,
bulk/tail effective sample size, and Monte-Carlo standard errors:

```{r chain-diag, eval = FALSE, purl = FALSE}
cd <- chain_diagnostics(result$chains$chain)   # n_draws x n_par (or a chain list)
print(cd)                                       # rhat, ess_bulk, ess_tail, mcse
```

The diagnostic suite covers static equation-system rank (D0), pre-estimation
identification (D1–D4, D19–D26, D37), MCMC quality (D5–D7), post-estimation fit
(D8–D18), and advanced topics such as OBC identification and scenarios (D27,
D31–D32), regime-switching identification (D28), data-driven constraints (D29),
arbitrary-precision rank checks (D30), the "deep parameter" checks D33–D36, and
Kalman-innovation whiteness (D41).  Three further diagnostics — D38
(`diag_sloppiness()`), D39 (`diag_stability_map()`) and D40
(`diag_near_unit_root()`) — are called directly rather than through the
orchestrator.  See `vignette("diagnostics")` for details.

Key convergence checks:

| Metric | Good | Concern |
|---|---|---|
| R-hat | \(\le 1.01\) | \(>1.01\) (WARN), \(\ge 1.05\) (FAIL) |
| Bulk **and** tail ESS | both \(\ge 100 \times\) `max(n_chains, 2)` | either below (WARN) |
| Acceptance rate (RWMH) | 20–35% | \(<10\%\) or \(>50\%\) |
| Divergences (NUTS) | 0 | \(>0\) |

---

## 9  Optimal policy (optional)

dynhr ships a comprehensive optimal-policy toolkit that goes well beyond
the simple simulation-based approach:

### 9.1  Ramsey optimal commitment

Two complementary engines are available:

| Method | Function | Description |
|--------|----------|-------------|
| **Augmented system** | `ramsey_model()` | Derives FOCs symbolically, double state space (Bodenstein–Guerrieri) |
| **(n, n+1) approximation** | `ramsey_nn1()` | Modified objective without doubling the state (Gross & Hansen 2021) |

`ramsey_model()` is the gold standard; `ramsey_nn1()` is lighter weight
and scales naturally to OBC problems.

Both can be called from `run_full_estimation(..., run_ramsey = TRUE)`:

```{r ramsey, eval = FALSE, purl = FALSE}
result_ramsey <- ramsey_model(model, compiled, params,
                               planner_objective = "log(c)",
                               order = 1L, verbose = TRUE)
print(result_ramsey)
```

### 9.2  Optimal simple rules (OSR)

`osr()` optimises the coefficients of a pre-specified policy rule to
minimise the unconditional variance of target variables, using CMA-ES:

```{r osr, eval = FALSE, purl = FALSE}
osr_result <- osr(model, free_params = c(phi_pi = 1.5, phi_y = 0.5),
                   loss_vars = c("pi", "y_gap"),
                   loss_weights = c(pi = 1, y_gap = 0.5),
                   lower = c(phi_pi = 1.01, phi_y = 0),
                   upper = c(phi_pi = 5,    phi_y = 2))
```

As in Dynare, the coefficients are unbounded unless you give bounds (here,
or in an `osr_params_bounds` block): a rule responding to current inflation
can otherwise run off towards infinity. A `.mod` file with `osr_params`,
`optim_weights` (including cross terms `var1, var2 w;`) and
`osr_params_bounds` drives `osr(model)` directly.

### 9.3  Discretionary policy

`discretionary_policy()` computes the Markov-perfect equilibrium under
discretion via Riccati fixed-point iteration:

```{r disc, eval = FALSE, purl = FALSE}
disc_result <- discretionary_policy(model,
                                     policy_instrument = "i",
                                     loss_vars = c("pi", "y_gap"))
```

### 9.4  Nash–Ramsey games

- `nash_ramsey_cooperative()` — multiple policymakers with a weighted-sum objective
- `nash_ramsey_openloop()` — open-loop Nash equilibrium (welfare matrix)

### 9.5  Ramsey with OBC

- `ramsey_obc_pf()` — perfect-foresight Ramsey under binding constraints
- `ramsey_obc_pwlinear()` — piecewise-linear Ramsey commitment
- `ramsey_obc_welfare_cost()` — compare welfare across OBC regimes

### 9.6  Regime-dependent Ramsey

- `ramsey_regime_deterministic()` — deterministic regime path with per-regime objectives
- `ramsey_regime_independent()` — Markov-switching Ramsey with ergodic welfare

### 9.7  Sufficient-statistics optimal policy

`opp_sufficient_stats()` implements the Barnichon & Mesters (2023)
approach: given IRFs to a policy instrument, compute the optimal
instrument deviation that minimises a quadratic loss subject to the
IRF-based transmission:

```{r opp, eval = FALSE, purl = FALSE}
opp_result <- opp_sufficient_stats(model,
                                    target_vars = c("pi", "y_gap"),
                                    loss_weights = c(pi = 1, y_gap = 0.5))
```

### 9.8  Welfare toolkit

- `welfare_compute()` — simulation-based welfare evaluation
- `welfare_ce_diff()` — consumption-equivalent welfare difference
- `welfare_cost_of_rule()` — cost of a given policy rule
- `welfare_decompose()` — welfare decomposition into steady-state, variance, and higher-order components
- `conditional_welfare()` — conditional welfare under alternative policies
- `evaluate_planner_objective()` — evaluate the planner's objective at a given solution

`run_diagnostics()` will consume Ramsey output in D18 welfare plausibility when
available; otherwise D18 remains informational.

---

## 10  Saving and loading chains

```{r save, purl = FALSE}
saveRDS(result$chains, "output/chains.rds")
chains <- readRDS("output/chains.rds")

# Convert to coda::mcmc.list for standard diagnostics
chain_mat <- chains$chain
# Use coda::as.mcmc(chain_mat) if coda is installed
```

### Checkpointing and resuming a long run

`saveRDS()` preserves the draws but not the sampler: restarting from it
begins a fresh chain at a fresh adaptation state. For runs long enough to
outlive a session or a scheduler slot, `dynhr_mcmc()` checkpoints instead:

```{r checkpoint, eval = FALSE, purl = FALSE}
# First leg: writes a state pack every `flush_every` draws
res1 <- dynhr_mcmc(log_post, theta0, Sigma_prop,
                   n_draws = 5000, n_warmup = 2000,
                   checkpoint_dir = "output/chain1",
                   flush_every    = 1000)

# Later, in a new session: continue where it stopped.
# `n_warmup` stays as it was; grow `n_draws` to the new TOTAL.
res2 <- dynhr_mcmc(log_post, theta0, Sigma_prop,
                   n_draws = 20000, n_warmup = 2000,
                   checkpoint_dir = "output/chain1",
                   resume = TRUE)
```

The state pack carries the position, the log-posterior, the adaptation
state and `.Random.seed`, so a resumed chain is statistically identical to
the uninterrupted run. It is checksummed and written atomically, and
`mcmc_chain_restore()` refuses a tampered pack or one saved mid-adaptation
(pass `allow_mid_adaptation = TRUE` only if you know why you want that).
`mcmc_chain_state()`, `mcmc_chain_save()`, `mcmc_chain_restore()` and
`mcmc_chain_extend()` are the sampler-agnostic primitives underneath, for
building the same behaviour into a bespoke sampler.

---

## 11  Delayed acceptance

When a model has both a cheap approximate likelihood and an expensive exact
one, `screen_fn` runs the cheap one first and only evaluates the expensive
one on proposals that survive:

```{r delayed, eval = FALSE, purl = FALSE}
cheap <- make_posterior(model, data = Y, prior_spec = priors,
                        obs_vars = obs_vars, compiled = compiled)

exact <- make_log_posterior(model, data = Y, prior_spec = priors,
                            obs_vars = obs_vars, compiled = compiled,
                            likelihood = "tpf", me_variance = 0.01)

res <- dynhr_mcmc(exact, theta0, Sigma_prop,
                  n_draws = 10000, n_warmup = 2000,
                  screen_fn = cheap)
```

The two-stage acceptance ratio corrects for the screening, so the invariant
distribution is still the exact posterior. The saving is real only when the
screen rejects most proposals and is much cheaper than the target; a screen
that disagrees with the target wastes both evaluations.

---

## 12  Method of moments (GMM / SMM)

`method_of_moments()` estimates by matching model-implied moments to sample
moments rather than by evaluating a likelihood — useful when the likelihood
is unavailable, unreliable, or simply not the estimator you want to defend:

```{r mom, eval = FALSE, purl = FALSE}
mm <- method_of_moments(model, data = Y, obs_vars = obs_vars,
                        moments   = list(orders = 1:2, lags = 0:4),
                        method    = "gmm",       # or "smm"
                        weighting = "optimal",   # identity/newey_west/diagonal
                        two_step  = TRUE)
```

`method = "gmm"` uses analytic model moments; `"smm"` simulates them
(`n_sim` replications of the sample, common random numbers across
parameter draws). `weighting = "optimal"` runs the standard two-step
procedure, with `"newey_west"` available for serially correlated moment
conditions. The moment Jacobian is analytic where dynhr has one and falls
back to finite differences otherwise (`jacobian = "auto"`).

---

## 13  Specs, records and reruns

Every estimation entry point — `run_mode_finding()`,
`run_posterior_estimation()`, `run_full_estimation()` — builds an
**estimation spec** from its arguments and hands it to one runner,
`run_estimation()`. The spec is a single versioned object that fully
determines a run: the model, the data, the likelihood, the mode stage, the
sampler(s), the compute settings (parallelism, seed, checkpointing), the
post-estimation outputs, and every package option that can change a result.

### Building and editing a spec

```{r spec-build, purl = FALSE}
spec <- dynhr_estimation_spec(
  mod, data = Y, obs_vars = obs_vars,
  likelihood = likelihood_spec("gaussian", me_variance = 0),
  mode       = mode_spec(method = "newrat", n_iter = 2000L),
  sampler    = sampler_spec("nuts", n_draws = 2000L, n_warmup = 1000L,
                            n_chains = 4L, metric = "lowrank"),
  compute    = compute_spec(parallel = TRUE, seed = 1L),
  outputs    = outputs_spec(diagnostics = TRUE))
spec                                   # compact, one line per non-default field

spec8 <- update(spec, sampler = list(n_chains = 8L))   # edit only what changes
diff_specs(spec, spec8)                # sampler$n_chains: 4 -> 8
```

Each sub-spec validates its own fields when it is built — a field the chosen
sampler does not use is an error, not silently ignored — and
`validate_spec()` holds every check that spans components (OBC with a
Kalman-only likelihood, a checkpoint with a sampler that cannot stream, and
so on). Options that change results (`transform_params`, `grad_method`,
`power_posterior`, `seed_base`, ...) are **typed fields** of the sub-spec that
uses them. They default to the current `dynhr_set_options()` value when the
spec is built; later changes of the global option do not reach an existing
spec. While the run executes, the spec's values are the options in force,
including on parallel daemons, and the session's options are restored
afterwards.

`as_estimation_spec()` converts what you already have into a spec: the
arguments of the three entry points, a run record, an estimation result, or
a `.mod` file whose `estimation(...)` options become defaults.

### Spec files

```{r spec-files, purl = FALSE}
write_spec(spec, "nk_demo_spec.yaml")      # human-readable, commented
spec2 <- read_spec("nk_demo_spec.yaml")
identical(spec2$hashes$spec, spec$hashes$spec)   # TRUE: exact round trip
fit <- run_estimation(spec2)
```

YAML (`yaml`, a suggested package) is the primary text format and JSON the
secondary; both round-trip exactly, doubles included. They reference the
model and data by path and hash: `read_spec()` refuses a file whose content
changed since the spec was written (`force = TRUE` overrides). RDS stores
everything by value and is the canonical format; closures (a custom
system-prior density, for example) can only go there.

### Run records and `dynhr_rerun()`

Every result carries `$run_record`: the spec it ran (schema 2), the
resolved arguments, the option snapshot, the RNG state at entry and the
provenance (dynhr version and `GIT_COMMIT`, R, platform, BLAS/LAPACK,
cores). A record survives `saveRDS()` / `readRDS()` and replays with
`dynhr_rerun()`:

```{r rerun, purl = FALSE}
saveRDS(fit$run_record, "fit_record.rds")
again  <- dynhr_rerun(readRDS("fit_record.rds"))       # same build: bit-identical
longer <- dynhr_rerun(fit, sampler = list(n_draws = 10000L))  # edit, then replay
```

### Checkpoints and resuming

With `compute_spec(checkpoint_dir = ...)` (or the `checkpoint_dir` argument
of `run_posterior_estimation()` / `run_full_estimation()`) the chains stream
to disk and `resume = TRUE` continues them. The checkpoint stores the
spec's per-component content hashes and the build and environment it was
written under. A resume is **refused** when the target changed (model, data,
likelihood, mode or sampler settings — the continued chain would sample a
different posterior, so the pooled chain would be invalid), or when a
*registered result change* between the two dynhr versions touches a
component the chain uses. dynhr keeps a registry of every change that alters
results, tagged with the components it touches (the Whittle likelihood, the
OBC filter, the PSKF, the NUTS mass adaptation, the analytic gradient, ...),
and each spec declares the components it uses. `on_mismatch = "warn"`
resumes anyway; the result is then marked permanently
(`$provenance$integrity`, the run record and a `print()` banner). Any other
code or environment difference only warns: the chain stays a valid MCMC
chain, but it is not bit-identical to an uninterrupted run.

### What to expect across builds and machines, and `dynhr_verify()`

A record is compared with the current session in three classes: **A** the
target and algorithm (the spec's content hashes), **B** the code (version,
commit, and the result-change registry), **C** the numerical environment
(R, platform, BLAS/LAPACK, cores).

| Same A? | B (code) | C (environment) | What to expect |
|---|---|---|---|
| yes | same | same | bit-identical |
| yes | differs, no registered change touches the run | any | deterministic quantities (mode, log-posterior at a point) agree to ~1e-8 relative; MCMC draws are equivalent in distribution (means within MCSE), not bit-identical |
| yes | same | differs | as above; the number of cores alone does not change seeded draws |
| yes | a registered change touches the run | any | results are expected to differ; the registry says why |
| no | — | — | a different run |

| Operation | A differs | B: registered change touches the run | B otherwise, or C differs |
|---|---|---|---|
| `dynhr_rerun()` | n/a (the replay uses the record's A) | warning naming the changes and the version to install | message stating the expectation |
| resume a checkpoint | refused | refused | warning |

`dynhr_verify()` checks a finished result against this table under the
current build and environment. It rebuilds the log-posterior from the
record, re-evaluates it at the recorded mode and at a sample of recorded
draws, re-optimises briefly from the recorded mode (all at 1e-8 relative),
replays the sampler for a few hundred draws and compares the posterior means
within Monte Carlo error, and reports PASS or FAIL for each class:

```{r verify, purl = FALSE}
v <- dynhr_verify(fit)
v                  # verdict, per-class status, each check and its tolerance
v$checks
```

---

## See also

- `?run_full_estimation` -- complete argument reference
- `?prior_spec` -- prior distribution families supported
- `?make_posterior` -- posterior evaluator options
- `?find_mode` -- optimiser methods (CMA-ES, Nelder-Mead, JADE)
- `?run_mode_finding`, `?run_posterior_estimation` -- stepwise estimation API
- `?run_estimation`, `?dynhr_estimation_spec`, `?write_spec` -- estimation specs (§13)
- `?dynhr_rerun`, `?dynhr_verify` -- run records, replays and reproducibility checks
- `?make_log_posterior_cumulant` -- cumulant-based posterior
- `?make_log_posterior_global_pf`, `?global_pf_sbc` -- projection + particle filter
- `?kalman_filter` (`obs_aggregation`), `?mf_augment_state_space` -- mixed frequency
- `?dynhr_mcmc` (`checkpoint_dir`, `resume`, `screen_fn`) -- checkpointing, delayed acceptance
- `?mcmc_chain_state`, `?mcmc_chain_save`, `?mcmc_chain_restore` -- state packs
- `?method_of_moments` -- GMM / SMM
- `?forecast_backtest` -- out-of-sample scoring (see `vignette("diagnostics")`)
- `?ramsey_model`, `?ramsey_nn1` -- Ramsey optimal policy
- `?osr`, `?discretionary_policy` -- OSR and discretionary policy
- `?nash_ramsey_cooperative`, `?nash_ramsey_openloop` -- Nash games
- `?ramsey_obc_pf`, `?ramsey_obc_pwlinear` -- Ramsey with OBC
- `?ramsey_regime_deterministic`, `?ramsey_regime_independent` -- regime-dependent
- `?opp_sufficient_stats` -- sufficient-statistics OPP
- `?welfare_compute`, `?welfare_ce_diff` -- welfare toolkit
- `vignette("diagnostics")` -- the D0–D41 diagnostic battery

---

## References

Barnichon, R., & Mesters, G. (2023). A sufficient statistics approach for macro policy evaluation. *Journal of Political Economy*, 131(1), 159-197.

Bodenstein, M., & Guerrieri, L. (2011). The welfare-optimal degree of central bank transparency. *Journal of Money, Credit and Banking*, 43(5), 857-889.

Chopin, N. (2002). A sequential particle filter method for static models. *Biometrika*, 89(3), 539-552.

Gelman, A., Carlin, J. B., Stern, H. S., & Rubin, D. B. (2013). *Bayesian Data Analysis* (3rd ed.). Chapman & Hall/CRC.

Gross, T., & Hansen, J. (2021). Optimal policy design in nonlinear DSGE models: An n-order accurate approximation. *Working Paper*.

Hansen, N. (2016). The CMA evolution strategy: A tutorial. *arXiv:1604.00772*.

Herbst, E. P., & Schorfheide, F. (2015). *Bayesian Estimation of DSGE Models*. Princeton University Press.

Hoffman, M. D., & Gelman, A. (2014). The No-U-Turn sampler: Adaptively setting path lengths in Hamiltonian Monte Carlo. *Journal of Machine Learning Research*, 15(1), 1593-1623.

Mariano, R. S., & Murasawa, Y. (2003). A new coincident index of business cycles based on monthly and quarterly series. *Journal of Applied Econometrics*, 18(4), 427-443.

Mutschler, W. (2015). Identification of DSGE models — The effect of higher-order approximation and pruning. *Journal of Economic Dynamics and Control*, 56, 1-33.

Schmitt-Grohé, S., & Uribe, M. (2004). Solving dynamic general equilibrium models using a second-order approximation to the policy function. *Journal of Economic Dynamics and Control*, 28(4), 755-775.

Söderlind, P. (1999). Solution and estimation of RE macromodels with optimal policy. *European Economic Review*, 43(4-6), 813-823.
