---
title: "Bayesian estimation with dynhr"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Bayesian estimation with dynhr}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>",
  eval     = FALSE   # estimation is slow; show code without running
)
```

## Overview

This vignette walks through Bayesian estimation of a DSGE model in dynhr:

1. Specify priors from the `.mod` file
2. Build a log-posterior evaluator (Gaussian or cumulant-based)
3. Find the posterior mode
4. Draw from the posterior (RWMH, SMC, or NUTS)
5. Inspect convergence and run diagnostics (D1–D30)
6. (Optional) Evaluate Ramsey, OSR, discretionary, or Nash welfare

For a one-call pipeline skip to §6.

---

## 1  Prior specification

Priors live in the `estimated_params` block of the `.mod` file:

```
// example: nk_hs2016.mod
estimated_params;
  tau,    gamma_pdf,  2.00, 0.50;   // IES (inverse)
  kappa,  beta_pdf,   0.33, 0.10;   // NK slope
  psi1,   gamma_pdf,  1.50, 0.25;   // Taylor inflation
  rho_R,  beta_pdf,   0.75, 0.10;   // interest rate smoothing
  stderr eps_R, inv_gamma_pdf, 0.20, 0.10;
end;
```

Extract the prior specification in R:

```{r priors}
library(dynhr)
mod    <- system.file("extdata", "models", "nk_hs2016.mod", package = "dynhr")
model  <- parse_mod(mod, verbose = FALSE)
priors <- prior_spec(model)   # extracts estimated_params block
print(priors)
```

`priors` is a data.frame with columns `name`, `distribution`, `p1`, `p2`,
`lower`, `upper`, `mean`, `std`.

---

## 2  Build the log-posterior

```{r posterior}
compiled <- compile_model(model, verbose = FALSE)

# Observed data (T × n_obs matrix; column names must match obs vars in model)
Y <- read.csv("my_data.csv")
Y <- as.matrix(Y[, c("YGR", "INFL", "INT")])

# Build the closure
log_post <- make_posterior(model, data = Y, prior_spec = priors,
                           obs_vars    = c("YGR", "INFL", "INT"),
                           compiled    = compiled,
                           me_variance = 0)

# Test at prior means
theta0 <- setNames(priors$mean, priors$name)
val    <- log_post(theta0)
cat("logpost:", val$logpost, " loglik:", val$loglik, " logprior:", val$logprior)
```

`make_posterior()` caches the compiled model structure; evaluating the
returned closure is fast (one Kalman filter pass per call).

#### Likelihood initialization for (near-)unit-root models

The Kalman filter initialization is controlled by `lik_init`, which defaults
to `"auto"` and almost never needs to be set by hand. `"auto"` picks the
cheapest *valid* initialization: it uses the stationary Lyapunov
initialization whenever that yields a valid (finite, positive semi-definite)
initial covariance, and falls back to the exact-diffuse filter only for a
genuine unit or explosive root (e.g. a random-walk trend), where the
stationary variance does not exist.

This matters for any model whose feasible region is *near* a unit root (debt,
net foreign assets, trends). A root just inside the unit circle is still
stationary, so `"auto"` keeps the fast stationary path (~10 ms/eval) rather
than the per-step exact-diffuse phase (~500–800 ms/eval) — and returns the
*exact* stationary likelihood there, not a diffuse approximation. Because
global mode-finders (CMA-ES) and MCMC sample the feasibility boundary
heavily, this is frequently the difference between a tractable and an
intractable estimation. If you genuinely want the exact-diffuse treatment for
every near-unit root, pass `lik_init = "diffuse"` explicitly via
`posterior_options`.

### Cumulant-based likelihood

Instead of the standard Gaussian (Kalman) likelihood, pass
`likelihood = "cumulant"` to match first through fourth moments
(mean, variance, skewness, kurtosis) of the pruned state space
against the data (Mutschler 2015).  This is particularly useful
when the data exhibit non-Gaussian features:

```{r cumulant, eval = FALSE}
log_post_cumulant <- make_posterior(model, data = Y,
  prior_spec   = priors,
  obs_vars     = c("YGR", "INFL", "INT"),
  compiled     = compiled,
  likelihood   = "cumulant",
  orders       = 1:4        # which cumulants to match
)
```

The underlying `make_log_posterior_cumulant()` is also callable directly,
and the individual cumulant functions `compute_third_cumulant()` and
`compute_fourth_cumulant()` can be used standalone for analysis.

---

## 3  Find the posterior mode

```{r mode}
mode <- find_mode(
  log_post_fn = log_post,
  theta_init  = theta0,
  prior_spec  = priors,
  n_iter      = 10000L,
  method      = "newrat",     # default: csminwel quasi-Newton (= Dynare mode_compute 4)
  verbose     = TRUE
)
cat("Mode logpost:", mode$logpost, "\n")
cat("Mode estimate:\n")
print(round(mode$theta_mode, 4))
```

The `mode` list contains `theta_mode`, `logpost`, `hessian` (numerical),
`V_mode` (inverse-Hessian as proposal covariance), and `se_mode`.

#### Choosing the optimizer

The default `method = "newrat"` is the csminwel quasi-Newton optimizer —
the same algorithm as Dynare's `mode_compute = 4`. It is gradient-based
(using the analytic gradient, built automatically for Gaussian/cumulant/
Whittle likelihoods) with an infeasibility-backtracking line search, which
makes it both the most robust and the cheapest choice on near-unit-root
models: it typically converges in hundreds of evaluations, where the
gradient-free CMA-ES search needs thousands and tends to stall on the
feasibility boundary. If the posterior is multimodal or the starting point is
poor, use `method = "cmaes_newrat"` (CMA-ES global search to escape poor
starts, then newrat polish) or the previous default `method = "combined"`
(CMA-ES then L-BFGS-B).

#### Hessian performance levers

The proposal-covariance Hessian (the inverse of which is `V_mode`) is the
main cost after the optimizer itself. Two knobs control it, and the
performance-optimal choice depends on model size:

- **`parallel = TRUE`** (default `FALSE`) is the single biggest lever. The
  default computes the finite-difference Hessian serially
  (\eqn{O(n_{par}^2)} likelihood evaluations); `parallel = TRUE` spreads
  those over a mirai daemon pool for a roughly \eqn{n}-fold speed-up on
  \eqn{n} cores. It is opt-in (not auto-enabled) because the daemon pool has
  setup/teardown cost and is undesirable in CI or embedded runs — but for a
  large parameter vector it is the recommended setting.

- **`use_exact_hessian = TRUE`** (default `FALSE`) swaps the finite-difference
  Hessian for the analytic `posterior_hessian`. This is *not* a universal
  speed-up: the analytic Hessian costs
  \eqn{O(n_{par}^2 \cdot n_{state}^3)} and is the faster route only for a
  small parameter vector / long sample. For a large state and many parameters
  (e.g. \eqn{n_{par} \gtrsim 60}, \eqn{n_{state} \gtrsim 70}) the
  finite-difference Hessian wins, which is why it is the default — reach for
  `parallel = TRUE` rather than `use_exact_hessian = TRUE` on big models. (A
  related option, `use_analytic_hess`, default `TRUE`, uses the analytic
  Hessian only to seed the newrat optimizer's initial curvature; set
  `dynhr_set_options(use_analytic_hess = FALSE)` to skip that expensive seed
  on a large model.)

Both can be set per call or globally via `dynhr_set_options()`.

---

## 4  RWMH sampling

```{r rwmh}
# Scale the Hessian-based proposal
n_par      <- length(mode$theta_mode)
Sigma_prop <- mode$V_mode * (2.38^2 / n_par)

chains_rwmh <- mcmc(
  log_post_fn = log_post,
  theta0      = mode$theta_mode,
  Sigma_prop  = Sigma_prop,
  n_draws     = 100000L,
  n_warmup    = 25000L
)
print(chains_rwmh)
summary(chains_rwmh)
```

The returned `dynhr_chains` object has a `$chain` slot (\eqn{n_{\text{draws}} \times n_p}
post-warmup matrix).

### Multi-chain RWMH

Pass `n_chains > 1` to `run_full_estimation()` for automatic multi-chain
dispatch with Gelman--Rubin convergence diagnostics.

---

## 5  SMC (no mode required)

SMC is a particle filter over the parameter space.  It does not require
a mode estimate and provides a free log-marginal-likelihood estimate:

```{r smc}
chains_smc <- smc(
  log_post_fn = log_post,
  prior_spec  = priors,
  n_particles = 5000L,
  verbose     = TRUE
)
print(chains_smc)
cat("log p(Y|M):", chains_smc$log_marginal_lik, "\n")
```

---

## 6  NUTS

NUTS adapts its step size and mass matrix automatically:

```{r nuts}
chains_nuts <- nuts(
  log_post_fn = log_post,
  theta0      = mode$theta_mode,
  n_draws     = 4000L,
  n_warmup    = 2000L
)
print(chains_nuts)
```

Check divergences with `chains_nuts$n_divergent`.  A well-tuned run
should have zero divergences.

---

## 7  One-call pipeline

All of the above in \(<30\) lines:

```{r onecall}
result <- run_full_estimation(
  mod_file    = "nk_hs2016.mod",
  data        = Y,
  obs_vars    = c("YGR", "INFL", "INT"),
  output_dir  = "output/",
  sampler     = "rwmh",
  n_draws     = 100000L,
  n_warmup    = 25000L,
  n_chains    = 4L,
  seed        = 42L,
  run_diag    = TRUE,
  verbose     = TRUE
)
print(result)
```

`output/dynhr_est_mode.rds` and `output/dynhr_est_chains.rds` are
written automatically.

### Gradient samplers, the parameter transform, and metrics

`method = "NUTS"`, `"MALA"`, `"CHEES"` and `"HMC"` are gradient-based. By default
(`transform_params = TRUE`) they sample on an unconstrained reparameterisation of
the bounded parameter vector, built from the priors. This is the recommended
default: in the constrained geometry of parameters near a unit root every
gradient sampler suffers step-size collapse, which the transform resolves while
leaving the target distribution unchanged. Set `transform_params = FALSE` to
sample in raw parameter space.

`MALA` and `NUTS` accept a `metric`: `"diagonal"` (default, identity mass),
`"hessian"` (a constant dense metric from the mode covariance — the best choice
for ill-conditioned posteriors), or `"monge"`. **The `"monge"` metric is
experimental and disabled by default**: \(G = I + \alpha^2 g g^\top\) inflates
along the gradient, so the proposal step collapses on sharp / near-unit-root
posteriors (in testing it explored a tight direction to only ~1% of its variance
where `"hessian"` recovered it fully). It errors unless you set
`dynhr_set_options(allow_monge_metric = TRUE)`, and warns when used — prefer
`"hessian"`.

---

## 8  Convergence and diagnostics

After any sampler:

```{r diag}
# Run the 30-diagnostic battery (D1–D30)
diags <- run_diagnostics(
  model    = model,
  compiled = compiled,
  draws    = result$chains$chain,
  data     = Y
)

# Write a structured report
write_report(diags, file = "output/report.md")
```

The diagnostic suite covers pre-estimation identification (D1–D4, D19–D26),
MCMC quality (D5–D7), post-estimation fit (D8–D18), and advanced topics
such as OBC identification (D27), regime-switching identification (D28),
data-driven constraints (D29), and arbitrary-precision rank checks (D30).
See `vignette("diagnostics")` for details.

Key convergence checks:

| Metric | Good | Concern |
|---|---|---|
| R-hat | \(<1.05\) | \(>1.10\) |
| Bulk ESS | \(>400\) per chain | \(<100\) |
| Acceptance rate (RWMH) | 20–35% | \(<10\%\) or \(>50\%\) |
| Divergences (NUTS) | 0 | \(>0\) |

---

## 9  Optimal policy (optional)

dynhr ships a comprehensive optimal-policy toolkit that goes well beyond
the simple simulation-based approach:

### 9.1  Ramsey optimal commitment

Two complementary engines are available:

| Method | Function | Description |
|--------|----------|-------------|
| **Augmented system** | `ramsey_model()` | Derives FOCs symbolically, double state space (Bodenstein–Guerrieri) |
| **(n, n+1) approximation** | `ramsey_nn1()` | Modified objective without doubling the state (Gross & Hansen 2021) |

`ramsey_model()` is the gold standard; `ramsey_nn1()` is lighter weight
and scales naturally to OBC problems.

Both can be called from `run_full_estimation(..., run_ramsey = TRUE)`:

```{r ramsey, eval = FALSE}
result_ramsey <- ramsey_model(model, compiled, params,
                               planner_objective = "log(c)",
                               order = 1L, verbose = TRUE)
print(result_ramsey)
```

### 9.2  Optimal simple rules (OSR)

`osr()` optimises the coefficients of a pre-specified policy rule to
minimise the unconditional variance of target variables, using CMA-ES:

```{r osr, eval = FALSE}
osr_result <- osr(model, free_params = c(phi_pi = 1.5, phi_y = 0.5),
                   loss_vars = c("pi", "y_gap"),
                   loss_weights = c(pi = 1, y_gap = 0.5))
```

### 9.3  Discretionary policy

`discretionary_policy()` computes the Markov-perfect equilibrium under
discretion via Riccati fixed-point iteration:

```{r disc, eval = FALSE}
disc_result <- discretionary_policy(model,
                                     policy_instrument = "i",
                                     loss_vars = c("pi", "y_gap"))
```

### 9.4  Nash–Ramsey games

- `nash_ramsey_cooperative()` — multiple policymakers with a weighted-sum objective
- `nash_ramsey_openloop()` — open-loop Nash equilibrium (welfare matrix)

### 9.5  Ramsey with OBC

- `ramsey_obc_pf()` — perfect-foresight Ramsey under binding constraints
- `ramsey_obc_pwlinear()` — piecewise-linear Ramsey commitment
- `ramsey_obc_welfare_cost()` — compare welfare across OBC regimes

### 9.6  Regime-dependent Ramsey

- `ramsey_regime_deterministic()` — deterministic regime path with per-regime objectives
- `ramsey_regime_markov()` — Markov-switching Ramsey with ergodic welfare

### 9.7  Sufficient-statistics optimal policy

`opp_sufficient_stats()` implements the Barnichon & Mesters (2023)
approach: given IRFs to a policy instrument, compute the optimal
instrument deviation that minimises a quadratic loss subject to the
IRF-based transmission:

```{r opp, eval = FALSE}
opp_result <- opp_sufficient_stats(model,
                                    target_vars = c("pi", "y_gap"),
                                    loss_weights = c(pi = 1, y_gap = 0.5))
```

### 9.8  Welfare toolkit

- `welfare_compute()` — simulation-based welfare evaluation
- `welfare_ce_diff()` — consumption-equivalent welfare difference
- `welfare_cost_of_rule()` — cost of a given policy rule
- `welfare_decompose()` — welfare decomposition into steady-state, variance, and higher-order components
- `conditional_welfare()` — conditional welfare under alternative policies
- `evaluate_planner_objective()` — evaluate the planner's objective at a given solution

`run_diagnostics()` will consume Ramsey output in D18 welfare plausibility when
available; otherwise D18 remains informational.

---

## 10  Saving and loading chains

```{r save}
saveRDS(result$chains, "output/chains.rds")
chains <- readRDS("output/chains.rds")

# Convert to coda::mcmc.list for standard diagnostics
chain_mat <- chains$chain
# Use coda::as.mcmc(chain_mat) if coda is installed
```

---

## See also

- `?run_full_estimation` -- complete argument reference
- `?prior_spec` -- prior distribution families supported
- `?make_posterior` -- posterior evaluator options
- `?find_mode` -- optimiser methods (CMA-ES, Nelder-Mead, JADE)
- `?estimate_mode`, `?estimate_posterior` -- stepwise estimation API
- `?make_log_posterior_cumulant` -- cumulant-based posterior
- `?ramsey_model`, `?ramsey_nn1` -- Ramsey optimal policy
- `?osr`, `?discretionary_policy` -- OSR and discretionary policy
- `?nash_ramsey_cooperative`, `?nash_ramsey_openloop` -- Nash games
- `?ramsey_obc_pf`, `?ramsey_obc_pwlinear` -- Ramsey with OBC
- `?ramsey_regime_deterministic`, `?ramsey_regime_markov` -- regime-dependent
- `?opp_sufficient_stats` -- sufficient-statistics OPP
- `?welfare_compute`, `?welfare_ce_diff` -- welfare toolkit
- `vignette("diagnostics")` -- the D1–D30 diagnostic battery

---

## References

Barnichon, R., & Mesters, G. (2023). A sufficient statistics approach for macro policy evaluation. *Journal of Political Economy*, 131(1), 159-197.

Bodenstein, M., & Guerrieri, L. (2011). The welfare-optimal degree of central bank transparency. *Journal of Money, Credit and Banking*, 43(5), 857-889.

Chopin, N. (2002). A sequential particle filter method for static models. *Biometrika*, 89(3), 539-552.

Gelman, A., Carlin, J. B., Stern, H. S., & Rubin, D. B. (2013). *Bayesian Data Analysis* (3rd ed.). Chapman & Hall/CRC.

Gross, T., & Hansen, J. (2021). Optimal policy design in nonlinear DSGE models: An n-order accurate approximation. *Working Paper*.

Hansen, N. (2016). The CMA evolution strategy: A tutorial. *arXiv:1604.00772*.

Herbst, E. P., & Schorfheide, F. (2015). *Bayesian Estimation of DSGE Models*. Princeton University Press.

Hoffman, M. D., & Gelman, A. (2014). The No-U-Turn sampler: Adaptively setting path lengths in Hamiltonian Monte Carlo. *Journal of Machine Learning Research*, 15(1), 1593-1623.

Mutschler, W. (2015). Identification of DSGE models — The effect of higher-order approximation and pruning. *Journal of Economic Dynamics and Control*, 56, 1-33.

Schmitt-Grohé, S., & Uribe, M. (2004). Solving dynamic general equilibrium models using a second-order approximation to the policy function. *Journal of Economic Dynamics and Control*, 28(4), 755-775.

Söderlind, P. (1999). Solution and estimation of RE macromodels with optimal policy. *European Economic Review*, 43(4-6), 813-823.
