---
title: "Heterogeneous-agent (HANK) models with dynhr"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Heterogeneous-agent (HANK) models with dynhr}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE, purl = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>",
  eval     = FALSE   # runnable, but skipped on build for speed
)
## `eval = FALSE` only stops KNITTING. R CMD check's "running R code from
## vignettes" step executes the TANGLED code, which knitr::purl() emits
## regardless of `eval` -- that is what made this vignette a check ERROR.
## `purl = FALSE` must go in each chunk HEADER, not here: during tangling
## knitr never evaluates this chunk, so opts_chunk$set() cannot reach it.
## The calls below are gated instead by tests/testthat/test-vignette-code.R.
```

## Overview

dynhr solves, computes impulse responses for, and **Bayesian-estimates**
heterogeneous-agent models using the sequence-space Jacobian (Auclert,
Bardóczy, Rognlie & Straub 2021). Unlike the representative-agent workflow,
a HANK household block is *not* written in a `.mod` file — it is specified on
the R side from three ingredients:

1. an idiosyncratic income process (a discretized AR(1)),
2. an asset grid, and
3. an EGM household solver + Young's-method distribution.

The aggregate ("simple") blocks — a firm, market clearing — compose with the
household block through the sequence-space Jacobian to close general
equilibrium. This vignette walks through the canonical **Krusell-Smith** model.

### What the package provides, and what you decide

The division of labour is worth stating up front, because "dynhr does GE" is
true in a narrower sense than it sounds.

**The package owns the machinery.** Household blocks and their sequence-space
Jacobians at every asset tier; the DAG accumulation and the equilibrium solve
`dU = -H_U⁻¹ H_Z dZ` (`hank_model()`, `hank_model_irf()`); the map from a
solved model to a state space and a likelihood (`hank_state_space()`,
`hank_kalman_loglik()`, `hank_loglik_ar()`); and the estimation stack on top
of that. None of it is specific to any one economy — `hank_model()` takes an
arbitrary DAG.

**You own the model.** Which aggregate blocks exist, which paths are the
unknowns, which equations are the zero targets, what drives the system, and
the steady state it is all linearised around. Those are the arguments to
`hank_model()`, and the package cannot infer them: they *are* the economics.

Two demonstration closures are supplied — Krusell-Smith (`hank_ks_steady()`,
below) and a two-asset economy (`hank_twoasset_model()`) — but they are
examples of the modelling step, not a substitute for it. Notably there is **no
built-in three-asset closure**: a third asset has to say what it *is* (a
foreign claim, at what world price, with what counterparty), and that is a
modelling decision.

### Household tiers

The household block comes in tiers, differing in how many costly assets the
household chooses jointly. All of them expose the same three things — a
fake-news Jacobian, a brute-force numerical-differentiation oracle, and a DAG
block spec:

| tier | constructor | prices | DAG spec |
|---|---|---|---|
| one asset | `hank_het_block()` | `r`, `w` (+ `Tr`, `r_minus`) | `hank_het_block_spec()` |
| two assets | `hank_het2_block()` | `rb`, `ra`, `w` (+ `Tr`, `theta_coll`) | `hank_het2_block_spec()` |
| three assets | `hank_het3_block()` | `rd`, `rf`, `ra`, `w` (+ `px`) | `hank_het3_block_spec()` |

Every tier additionally accepts **endogenous transition probabilities**: build
the block with `Pi_fn`/`Pi_inputs` (e.g. from `hank_employment_income()`, whose
job-finding rate `f` and separation rate `s` move the income transition matrix)
and those become perturbable aggregate inputs alongside the prices, with their
own Jacobian columns. This is what lets an upstream search-and-matching block
drive household employment risk. The transition inputs default *on* at the
Jacobian layer and are *opt-in* at the DAG layer, since declaring a DAG input
commits you to producing it upstream.

## 1. The household block

```{r household, purl = FALSE}
library(dynhr)

## Idiosyncratic productivity: a 7-state Rouwenhorst chain, mean-normalised to 1.
inc <- hank_income_rouwenhorst(rho = 0.966, sigma = 0.5, n = 7)

## Asset grid: 500 points, clustered near the borrowing constraint at 0.
a_grid <- hank_asset_grid(amax = 200, n = 500, amin = 0)

## Solve the household problem (EGM) at fixed prices, then its distribution.
hh  <- hank_egm_solve(a_grid, y = inc$e, r = 0.01, beta = 0.98,
                      eis = 1, Pi = inc$Pi)
Lam <- hank_forward_operator(hh$a, a_grid, inc$Pi)
D   <- hank_stationary_dist(Lam)$d

## Aggregate assets and consumption:
A <- hank_aggregate(D, hh$a)
C <- hank_aggregate(D, hh$c)
```

The Euler-equation residual (`hank_euler_residual(hh)`) and the exact
aggregate-budget identity `C == r*A + Ybar` provide self-contained correctness
checks.

Two extensions to the income process live alongside this two-state (E/U)
version: `hank_employment_income3()` adds a third, not-in-labour-force state
(E/U/N) with its own job-finding/separation rates, and `hank_het_block(Tr_incidence =)`
accepts a full `n_e x n_a` matrix so a transfer's incidence can vary by
wealth as well as income state, not just be a flat per-state amount.

## 2. General equilibrium (Krusell-Smith)

`hank_ks_steady()` closes the model with a Cobb-Douglas firm and asset-market
clearing, solving for the interest rate at which household savings equal the
firm's capital demand (the Aiyagari fixed point).

```{r ge, purl = FALSE}
ks <- hank_ks_steady(a_grid, inc$Pi, inc$e, beta = 0.98, eis = 1,
                     alpha = 0.36, delta = 0.025)
ks$r; ks$K            # equilibrium interest rate and capital

## Linear GE impulse response to a persistent TFP shock:
Th  <- 300
dZ  <- 0.01 * 0.9^(0:(Th - 1))
irf <- hank_ks_linear_irf(ks, dZ)
hank_plot_irf(irf, vars = c("dY", "dK", "dC"))

## Fully nonlinear transition (quasi-Newton in sequence space):
nl <- hank_ks_nonlinear_irf(ks, Z_path = ks$Z * (1 + 0.05 * 0.9^(0:(Th - 1))))
```

## 3. Diagnostics

```{r diag, purl = FALSE}
## Sequence-space Jacobian of consumption to the interest rate, and its heatmap:
J <- hank_het_jacobian(ks$block, Th, inputs = "r", outputs = "C")
hank_plot_jacobian(J[["C"]][["r"]])

## Marginal propensity to consume by wealth:
m <- hank_mpc(ks$block)
m$aggregate                       # distribution-weighted aggregate MPC
plot(m$by_wealth$a, m$by_wealth$mpc, type = "l",
     xlab = "assets", ylab = "MPC")
```

`hank_het_jacobian()` above gives the response of an *aggregate*; the
distribution-Jacobian family (`hank_het_dist_jacobian()` and its two/three-asset
siblings `hank_het2_dist_jacobian()` / `hank_het3_dist_jacobian()`) instead
returns `dD_t / d input_s`, the response of the cross-sectional distribution
itself, which is what `hank_model_dist_irf()` propagates through GE.

## 4. Estimation

The linearized GE solution implies a moving-average representation of the
aggregate observables in the structural innovations; the exact Gaussian
likelihood follows from the implied autocovariance function (ABRS eq. 37) — no
state-space or Kalman filter is required.

```{r estimate, purl = FALSE}
Theta <- hank_ma_coefficients(ks, rho_z = 0.9, T_h = 300, observables = "Y")

## Simulate a sample and evaluate / profile the likelihood:
set.seed(1)
Y  <- hank_simulate_aggregate(Theta, sigma_eps = 0.01, T_data = 200)
ll <- hank_loglik_aggregate(Y, Theta, sigma_eps = 0.01)

## Profile over the shock standard deviation (recovers the true value):
grid <- seq(0.006, 0.016, by = 0.0005)
ll_prof <- sapply(grid, function(s) hank_loglik_aggregate(Y, Theta, sigma_eps = s))
grid[which.max(ll_prof)]
```

With more observables than structural shocks the aggregate covariance is
stochastically singular; supply a measurement-error variance (`me_var`) to keep
the likelihood well-defined.

## 5. A discount-heterogeneity mixture economy

The flagship application in dynhr 0.9.0 is a **mixture of household types that
differ in their discount factor** — the cstwMPC device for matching the
empirical wealth distribution. Each type faces the same aggregate prices and
interacts only through them, so the economy is a single Krusell-Smith GE with a
population `omega`-weighted over types.

```{r mixture-solve, purl = FALSE}
## Two types: patient (beta = 0.985) and impatient (beta = 0.96), with
## population weights omega. All types share the income process and asset grid.
betas <- c(0.985, 0.96)
omega <- c(0.6, 0.4)

mks <- hank_mixture_ks_steady(a_grid, inc$Pi, inc$e,
                              betas = betas, omega = omega, eis = 1,
                              alpha = 0.36, delta = 0.025)
mks$r; mks$K                     # shared equilibrium price and aggregate capital

## Assemble the sequence-space GE model and its impulse response to a TFP shock:
T_h   <- 300
model <- hank_mixture_ks_model(mks, T_h)

## NOTE the calling convention: the general sequence-space `hank_model_irf()`
## takes a NAMED list keyed on the model's own exogenous ("Z" here) and returns
## the model's bare variable names. `hank_ks_linear_irf()` above is the other
## convention -- a bare vector in, `d`-prefixed names out.
irf   <- hank_model_irf(model, dZ = list(Z = 0.01 * 0.9^(0:(T_h - 1))))
names(irf)                       # K, Z, r, w, A, C, asset_mkt
hank_plot_irf(irf, vars = c("K", "C", "r"))
```

## 6. Estimating the discount spread

The headline identification result baked into these tools: the discount **spread**
between types is pinned down by the stationary cross-sectional wealth **level**
(à la cstwMPC), *not* by the aggregate response to a price shock, which is nearly
inert in the spread. `hank_mixture_joint_logpost()` evaluates a joint posterior
over up to three information channels — aggregate **macro** dynamics (through the
Kalman filter), the cross-sectional wealth **level**, and the cross-sectional
**response** to a shock:

```{r mixture-post, purl = FALSE}
## `config` bundles everything shared across every posterior evaluation
## (asset grid, income process, GE horizon, observables, shock specs, me_var,
##  and the fixed mixture coordinates). See ?hank_mixture_joint_logpost.
config <- list(a_grid = a_grid, Pi = inc$Pi, e = inc$e, T_h = T_h,
               observables = "Y", betas = betas, omega = omega,
               shock_specs = list(Z = list(rho = 0.9)), me_var = 1e-4)

## theta here is the estimated coordinate(s) — e.g. the discount-factor spread.
lp <- function(theta)
  hank_mixture_joint_logpost(theta, Y = Y, m_lev_hat = m_lev_hat,
                             level_metric = level_metric, config = config)
```

**Choosing a posterior route.** The channels are strongly non-diagonal, and a
diagonal-metric random-walk sampler fails on this posterior class. The
recommended routes are the **exact grid** (cheap here because the mixture GE
solve is data-independent and precomputable) or a **Laplace approximation** —
never a diagonal RWM:

```{r mixture-route, purl = FALSE}
## (a) exact grid, with simulation-based-calibration certification:
emu <- hank_mixture_emulator(config)          # D-agnostic surrogate over theta
sbc <- hank_mixture_sbc(emu, n_rep = 200, sampler = "grid",
                        channels = c("macro", "level", "response"))
sbc$uniformity                                # per-channel SBC rank uniformity

## (b) Laplace approximation: mode + Laplace-covariance proposal (NOT diagonal RWM):
fit <- hank_mixture_laplace(theta0 = 0.02, config = config,
                            Y = Y, m_lev_hat = m_lev_hat,
                            level_metric = level_metric)
fit$mode; fit$cov
```

## 7. Welfare

`hank_welfare_posterior()` propagates posterior draws through a welfare
functional (e.g. the consumption-equivalent gain of a policy) to get a posterior
over welfare quantiles, distinguishing the interest-income and labour-income
incidence channels across the wealth distribution:

```{r mixture-welfare, purl = FALSE}
draws <- fit$mode + as.numeric(chol(fit$cov)) * rnorm(1000)   # illustrative draws
wp <- hank_welfare_posterior(draws,
                             welfare_fn = function(theta)
                               hank_cev(mks, dr = +0.001))     # CEV of +10bps r
wp$quantiles
```

## References

- Auclert, Bardóczy, Rognlie & Straub (2021), "Using the Sequence-Space
  Jacobian to Solve and Estimate Heterogeneous-Agent Models," *Econometrica*
  89(5), 2375–2408.
- Carroll (2006), the endogenous gridpoints method.
- Young (2010), non-stochastic simulation of the distribution.
