---
title: "Heterogeneous-agent (HANK) models with dynhr"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Heterogeneous-agent (HANK) models with dynhr}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>",
  eval     = FALSE   # runnable, but skipped on build for speed
)
```

## Overview

dynhr solves, computes impulse responses for, and **Bayesian-estimates**
heterogeneous-agent models using the sequence-space Jacobian (Auclert,
Bardóczy, Rognlie & Straub 2021). Unlike the representative-agent workflow,
a HANK household block is *not* written in a `.mod` file — it is specified on
the R side from three ingredients:

1. an idiosyncratic income process (a discretized AR(1)),
2. an asset grid, and
3. an EGM household solver + Young's-method distribution.

The aggregate ("simple") blocks — a firm, market clearing — compose with the
household block through the sequence-space Jacobian to close general
equilibrium. This vignette walks through the canonical **Krusell-Smith** model.

## 1. The household block

```{r household}
library(dynhr)

## Idiosyncratic productivity: a 7-state Rouwenhorst chain, mean-normalised to 1.
inc <- hank_income_rouwenhorst(rho = 0.966, sigma = 0.5, n = 7)

## Asset grid: 500 points, clustered near the borrowing constraint at 0.
a_grid <- hank_asset_grid(amax = 200, n = 500, amin = 0)

## Solve the household problem (EGM) at fixed prices, then its distribution.
hh  <- hank_egm_solve(a_grid, y = inc$e, r = 0.01, beta = 0.98,
                      eis = 1, Pi = inc$Pi)
Lam <- hank_forward_operator(hh$a, a_grid, inc$Pi)
D   <- hank_stationary_dist(Lam)$d

## Aggregate assets and consumption:
A <- hank_aggregate(D, hh$a)
C <- hank_aggregate(D, hh$c)
```

The Euler-equation residual (`hank_euler_residual(hh)`) and the exact
aggregate-budget identity `C == r*A + Ybar` provide self-contained correctness
checks.

## 2. General equilibrium (Krusell-Smith)

`hank_ks_steady()` closes the model with a Cobb-Douglas firm and asset-market
clearing, solving for the interest rate at which household savings equal the
firm's capital demand (the Aiyagari fixed point).

```{r ge}
ks <- hank_ks_steady(a_grid, inc$Pi, inc$e, beta = 0.98, eis = 1,
                     alpha = 0.36, delta = 0.025)
ks$r; ks$K            # equilibrium interest rate and capital

## Linear GE impulse response to a persistent TFP shock:
Th  <- 300
dZ  <- 0.01 * 0.9^(0:(Th - 1))
irf <- hank_ks_linear_irf(ks, dZ)
hank_plot_irf(irf, vars = c("dY", "dK", "dC"))

## Fully nonlinear transition (quasi-Newton in sequence space):
nl <- hank_ks_nonlinear_irf(ks, Z_path = ks$Z * (1 + 0.05 * 0.9^(0:(Th - 1))))
```

## 3. Diagnostics

```{r diag}
## Sequence-space Jacobian of consumption to the interest rate, and its heatmap:
J <- hank_het_jacobian(ks$block, Th, inputs = "r", outputs = "C")
hank_plot_jacobian(J[["C"]][["r"]])

## Marginal propensity to consume by wealth:
m <- hank_mpc(ks$block)
m$aggregate                       # distribution-weighted aggregate MPC
plot(m$by_wealth$a, m$by_wealth$mpc, type = "l",
     xlab = "assets", ylab = "MPC")
```

## 4. Estimation

The linearized GE solution implies a moving-average representation of the
aggregate observables in the structural innovations; the exact Gaussian
likelihood follows from the implied autocovariance function (ABRS eq. 37) — no
state-space or Kalman filter is required.

```{r estimate}
Theta <- hank_ma_coefficients(ks, rho_z = 0.9, T_h = 300, observables = "Y")

## Simulate a sample and evaluate / profile the likelihood:
set.seed(1)
Y  <- hank_simulate_aggregate(Theta, sigma_eps = 0.01, T_data = 200)
ll <- hank_loglik_aggregate(Y, Theta, sigma_eps = 0.01)

## Profile over the shock standard deviation (recovers the true value):
grid <- seq(0.006, 0.016, by = 0.0005)
ll_prof <- sapply(grid, function(s) hank_loglik_aggregate(Y, Theta, sigma_eps = s))
grid[which.max(ll_prof)]
```

With more observables than structural shocks the aggregate covariance is
stochastically singular; supply a measurement-error variance (`me_var`) to keep
the likelihood well-defined.

## 5. A discount-heterogeneity mixture economy

The flagship application in dynhr 0.9.0 is a **mixture of household types that
differ in their discount factor** — the cstwMPC device for matching the
empirical wealth distribution. Each type faces the same aggregate prices and
interacts only through them, so the economy is a single Krusell-Smith GE with a
population `omega`-weighted over types.

```{r mixture-solve}
## Two types: patient (beta = 0.985) and impatient (beta = 0.96), with
## population weights omega. All types share the income process and asset grid.
betas <- c(0.985, 0.96)
omega <- c(0.6, 0.4)

mks <- hank_mixture_ks_steady(a_grid, inc$Pi, inc$e,
                              betas = betas, omega = omega, eis = 1,
                              alpha = 0.36, delta = 0.025)
mks$r; mks$K                     # shared equilibrium price and aggregate capital

## Assemble the sequence-space GE model and its impulse response to a TFP shock:
T_h   <- 300
model <- hank_mixture_ks_model(mks, T_h)
dZ    <- 0.01 * 0.9^(0:(T_h - 1))
irf   <- hank_model_irf(model, dZ)
hank_plot_irf(irf, vars = c("dY", "dK", "dC"))
```

## 6. Estimating the discount spread

The headline identification result baked into these tools: the discount **spread**
between types is pinned down by the stationary cross-sectional wealth **level**
(à la cstwMPC), *not* by the aggregate response to a price shock, which is nearly
inert in the spread. `hank_mixture_joint_logpost()` evaluates a joint posterior
over up to three information channels — aggregate **macro** dynamics (through the
Kalman filter), the cross-sectional wealth **level**, and the cross-sectional
**response** to a shock:

```{r mixture-post}
## `config` bundles everything shared across every posterior evaluation
## (asset grid, income process, GE horizon, observables, shock specs, me_var,
##  and the fixed mixture coordinates). See ?hank_mixture_joint_logpost.
config <- list(a_grid = a_grid, Pi = inc$Pi, e = inc$e, T_h = T_h,
               observables = "Y", betas = betas, omega = omega,
               shock_specs = list(Z = list(rho = 0.9)), me_var = 1e-4)

## theta here is the estimated coordinate(s) — e.g. the discount-factor spread.
lp <- function(theta)
  hank_mixture_joint_logpost(theta, Y = Y, m_lev_hat = m_lev_hat,
                             level_metric = level_metric, config = config)
```

**Choosing a posterior route.** The channels are strongly non-diagonal, and a
diagonal-metric random-walk sampler fails on this posterior class. The
recommended routes are the **exact grid** (cheap here because the mixture GE
solve is data-independent and precomputable) or a **Laplace approximation** —
never a diagonal RWM:

```{r mixture-route}
## (a) exact grid, with simulation-based-calibration certification:
emu <- hank_mixture_emulator(config)          # D-agnostic surrogate over theta
sbc <- hank_mixture_sbc(emu, n_rep = 200, sampler = "grid",
                        channels = c("macro", "level", "response"))
sbc$uniformity                                # per-channel SBC rank uniformity

## (b) Laplace approximation: mode + Laplace-covariance proposal (NOT diagonal RWM):
fit <- hank_mixture_laplace(theta0 = 0.02, config = config,
                            Y = Y, m_lev_hat = m_lev_hat,
                            level_metric = level_metric)
fit$mode; fit$cov
```

## 7. Welfare

`hank_welfare_posterior()` propagates posterior draws through a welfare
functional (e.g. the consumption-equivalent gain of a policy) to get a posterior
over welfare quantiles, distinguishing the interest-income and labour-income
incidence channels across the wealth distribution:

```{r mixture-welfare}
draws <- fit$mode + as.numeric(chol(fit$cov)) * rnorm(1000)   # illustrative draws
wp <- hank_welfare_posterior(draws,
                             welfare_fn = function(theta)
                               hank_cev(mks, dr = +0.001))     # CEV of +10bps r
wp$quantiles
```

## References

- Auclert, Bardóczy, Rognlie & Straub (2021), "Using the Sequence-Space
  Jacobian to Solve and Estimate Heterogeneous-Agent Models," *Econometrica*
  89(5), 2375–2408.
- Carroll (2006), the endogenous gridpoints method.
- Young (2010), non-stochastic simulation of the distribution.
