---
title: "Converting Dynare .mod files for dynhr"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Converting Dynare .mod files for dynhr}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>",
  eval     = TRUE
)
```

## Overview

dynhr parses standard Dynare `.mod` files with `parse_mod()`, so most
existing Dynare models can be loaded with no modifications.  On top of
the standard Dynare grammar, dynhr recognises a set of **optional
extensions** that live inside `//` comment blocks (so they are ignored
by Dynare itself) and drive the diagnostic suite, narrative analysis,
OBC specification, and optimal-policy engine.

This vignette documents every dynhr-specific annotation you can add
to a `.mod` file and shows how to convert an existing Dynare model to
make full use of the dynhr toolkit.

```{r library}
library(dynhr)
```

---

## 1  Basic parsing

The entry point is `parse_mod()`, which accepts any standard Dynare
`.mod` file:

```{r basic-parse}
mod_path <- system.file("extdata", "models", "rbc.mod", package = "dynhr")
model    <- parse_mod(mod_path, verbose = FALSE)
print(model)
```

The returned `dynhr_mod` object contains parsed equations, variable
declarations, parameter calibrations, `estimated_params` blocks, and
any `@dynhr:` metadata blocks (extracted into `model$metadata`).

Use `check_mod_metadata()` for a quick summary of what dynhr recognised:

```{r check-meta}
check_mod_metadata(mod_path)
```

---

## 2  Dynare → dynhr field-name map

Several Dynare output names differ from their dynhr equivalents.  The
table below covers the most common translation traps:

| Dynare name | dynhr name | Where |
|-------------|------------|-------|
| `oo_.dr.ghs2` | `dr$ghss` | Order-2 σ² uncertainty-correction vector (the "constant correction" to the mean from stochastic volatility) |
| `oo_.var` | `moms$var_cov` | Unconditional variance-covariance from `compute_moments()` |
| `oo_.mean` | *(not yet available)* | Order-2 ergodic mean; use `ys + 0.5 * dr$ghss` manually |
| `oo_.dr.ghx` | `dr$ghx` | State transition matrix (same name) |
| `oo_.dr.ghu` | `dr$ghu` | Shock-impact matrix (same name) |

**Key points:**

- `dr$ghss` is the dynhr field; `oo_.dr.ghs2` is the Dynare field — they
  contain the same quantity.  The `2` suffix is Dynare's internal convention
  for "second-order sigma term"; dynhr uses `ghss` to follow the standard
  mathematical notation (g_{σσ}).
- `compute_moments()` returns a list with field `var_cov`, not
  `variance_covariance`.  Other fields: `std_dev`, `correlation`,
  `autocorr`, `var_decomp`, `var_decomp_pct`.

### 2.1  `stoch_simul(loglinear)` — log-deviation scaling

When a Dynare model is solved with `stoch_simul(loglinear)`, Dynare
internally log-linearises the model around its steady state and returns
decision rules in **log-deviation** units: each entry \eqn{G_{ij}} in
Dynare's `oo_.dr.ghx` measures the response of \eqn{\log y_i} to a unit
change in \eqn{\log x_j}.

dynhr always returns decision rules in **level-deviation** units (deviations
from the deterministic steady state \eqn{y^*}), regardless of whether the
`.mod` file uses `loglinear`.  To compare dynhr's \eqn{G} against Dynare's
loglinear \eqn{G_{\log}}, apply the steady-state scaling:
\deqn{G_{\log}[i,j] = G[i,j] \cdot \frac{y^*_j}{y^*_i}}

In R, with `dr` from `solve_perturbation()` and `ys = dr$ys`:

```r
# Convert dynhr level-deviation ghx to Dynare loglinear log-deviation ghx
G_log <- dr$ghx * outer(1 / ys[dr$endo_names], ys[dr$endo_names[dr$state_idx]])
```

The same scaling applies to `ghu`, `ghxx`, `ghxu`, `ghuu`, and any IRF
matrix: multiply each entry by \eqn{y^*_j / y^*_i} for the appropriate
input/output pair.  Note that `dr$ys` for a `model(linear)` block returns
all ones (the log-level at the zero log-deviation point), not the original
model steady state; the relevant \eqn{y^*} is the *original* (pre-log)
steady state, recoverable from `ss$params` after `solve_steady()`.

**Worked example:** `GarciaCicco_et_al_2010` (`stoch_simul(loglinear)`)
after applying the scaling above: \eqn{\Delta \log(y^*) = 4.4 \times
10^{-15}}, \eqn{\max|\Delta G| = 5.8 \times 10^{-10}},
\eqn{\max|\Delta \text{IRF}| = 2.7 \times 10^{-11}},
\eqn{\max|\Delta \text{std}| = 6.1 \times 10^{-12}} versus Dynare 7.1
--- machine precision throughout.  Without the scaling the apparent difference
is \eqn{O(y^*)} and looks like a solver failure, making this the most common
comparison trap for loglinear models.

---

## 3  Standard Dynare blocks supported

| Block | dynhr support | Notes |
|-------|---------------|-------|
| `var`, `varexo`, `parameters` | ✅ Full | Including `var(log)` and `%(long_name='...')` annotations |
| `model;` ... `end;` | ✅ Full | Linear and nonlinear; lead/lag detection |
| `steady_state_model;` | ✅ Full | Analytical steady state |
| `shocks;` | ✅ Full | Shock variances |
| `estimated_params;` | ✅ Full | Prior distributions (beta, gamma, inv_gamma, normal, uniform) |
| `estimated_params_init;` | ✅ Full | Overrides for initial values |
| `varobs;` | ✅ Full | Observable variable declaration |
| `planner_objective(...);` | ✅ Full | Used by `ramsey_model()` and `ramsey_nn1()` |
| `occbin_constraints;` | ✅ Full | OBC constraint declarations |
| `model(linear);` | ✅ Full | Required for OBC piecewise-linear methods |
| `steady;`, `check;`, `stoch_simul;` | ✅ Parsed | Executed by `solve_steady()`, `solve_perturbation()`, `stoch_simul()` |

---

## 3a  Preserving Dynare syntax when converting: what you do NOT need to inline

A common conversion mistake is flattening Dynare-native constructs
before loading the model.  **dynhr supports all of the following
natively** — there is no need to pre-process the file:

### 3a.1  Model-local `#` intermediates

Dynare lets you define intermediate expressions inside the `model`
block with `#name = expr;`.  These are called *model-local variables*
and are distinct from endogenous or exogenous variables.  dynhr
preserves them in `model$local_variables` and substitutes them when
computing the Jacobian.

```
model;
  #theta = (1 - gamma) / (1 - 1/psi);   // Epstein-Zin aggregator exponent
  V = ((1 - beta) * c^((1 - gamma)/theta) + beta * s^(1/theta))^(theta/(1 - gamma));
  ...
end;
```

You do **not** need to substitute `(1 - gamma)/(1 - 1/psi)` everywhere
`theta` appears — just keep the `#theta` line as-is.

Verified in `caldara_rp.mod` and `nk_hs2016.mod`.

### 3a.2  `steady_state_model` blocks

An analytical steady state expressed as a sequence of assignments is
parsed directly.  dynhr's `solve_steady()` calls
`eval_steady_state_model()` which evaluates the assignments
top-to-bottom and returns the SS vector.

```
steady_state_model;
  dA   = exp(gam);
  m    = mst;
  k    = ((1 - gst*bet*(1-del)) / (alp*gst^alp*bet))^(1/(alp-1));
  ...
end;
```

You do **not** need to convert this into a hand-coded R function.
Keep the `steady_state_model` block in the `.mod` file.

Verified in `fs2000.mod` (18 assignments) and `caldara_rp.mod`.

### 3a.3  `@#` macro directives

dynhr expands Dynare's macro language **before** block parsing.  All
of the following are supported:

- `@#define NAME = VALUE` or `@#define NAME = ["a","b","c"]`
- `@#for V in [list]`, `@#for V in A:B`, `@#for V in defined_array`
- `@#endfor`
- `@#if COND`, `@#else`, `@#endif`
- `@#ifdef NAME`, `@#ifndef NAME`
- `@#include "file.inc"` (resolved relative to the `.mod` file's directory)
- `@{expr}` interpolation inside any line

You do **not** need to pre-expand macros before loading.  Run
`parse_mod()` directly on the unexpanded source.

Note: `@#elseif` is **not** supported (it errors loudly) — rewrite as
nested `@#if`/`@#else`.

### 3a.4  Common conversion gotchas

| Situation | What to do |
|-----------|-----------|
| `#theta = ...` inside model | Leave it — dynhr handles it |
| `steady_state_model;` block | Leave it — dynhr calls `eval_steady_state_model()` |
| `@#for`, `@#define`, `@#include` | Leave them — dynhr expands macros first |
| Parameters computed in `*_steadystate.m` | `parse_mod()` warns; use `inject_params(model, vals)` to supply them |
| `x(+1)` lead, `x(-1)` lag | Standard Dynare timing, fully supported |
| `x(+2)`, `eps(-1)` multi-period | dynhr auto-introduces auxiliary variables |
| `predetermined_variables k;` | Supported — dynhr shifts timing by -1 in ASTs |
| `model(linear)` | Supported — passed through in `model$model_options` |
| `var(log)` annotation | Supported — variable name stripped of `(log)` |

See `vignette("mod-syntax")` for the complete syntax reference with
verified examples.

---

## 4  dynhr-specific metadata blocks

dynhr recognises several `// @dynhr:<name>` comment blocks that are
**silently ignored by Dynare** (they are `//` comments).  They are
extracted by `extract_mod_metadata()` and stored in `model$metadata`.

### 4.1  `@dynhr:expectations`

Declares quantitative checks that `run_diagnostics()` evaluates
automatically.  Five check types are supported:

| Type | Fields | What it checks |
|------|--------|----------------|
| `data_mean` | `variable`, `min`, `max` | Mean of a data column is in range |
| `data_sd` | `variable`, `min`, `max` | Std dev of a data column is in range |
| `data_ratio` | `numerator`, `denominator`, `min`, `max` | Ratio of two column means is in range |
| `param_range` | `variable`, `min`, `max` | A parameter value is in range |
| `irf_sign` | `shock`, `variable`, `horizon`, `sign` | IRF has expected sign at horizon |

Example from the NK-HS2016 model:

```
// @dynhr:expectations
// infl_mean:  type="data_mean",   variable="INFL",  min=0.0,  max=6.0,
//             description="Annualised inflation should be positive and moderate"
// rho_R_hi:   type="param_range", variable="rho_R", min=0.5,  max=0.99,
//             description="Interest rate smoothing should be moderate to high"
// @dynhr:end
```

Pass data and parameters to `diag_expectations()` to evaluate them, or
let `run_diagnostics()` handle it automatically.

### 4.2  `@dynhr:benchmarks`

Declares expected IRF sign patterns that D8 (IRF plausibility) checks
automatically.  Each benchmark has the form:

```
// shock_name: shock=<VAR>, variable=<VAR>, direction=<positive|negative>,
//             description="<text>"
```

Example from the `03j` NZ production model:

```
// @dynhr:benchmarks
// monetary_y:    shock=eps_r,     variable=y,  direction=negative,
//                description="Contractionary MP reduces output"
// technology_y:  shock=eps_a,     variable=y,  direction=positive,
//                description="Positive TFP shock raises output"
// commodity_y:   shock=eps_cp,    variable=y,  direction=positive,
//                description="Commodity price boom raises output"
// @dynhr:end
```

### 4.3  `@dynhr:variable_labels`

Provides human-readable labels for endogenous variables, used in
diagnostic plots and reports:

```
// @dynhr:variable_labels
// y:   Output gap
// pi:  CPI inflation
// r:   Nominal interest rate
// q:   Real exchange rate
// @dynhr:end
```

You can also use Dynare's standard `%(long_name='...')` annotation on
`var` declarations — dynhr merges both sources, preferring the
`@dynhr:variable_labels` block.

### 4.4  `@dynhr:shock_categories`

Groups shocks into named categories for narrative analysis (D17) and
for aggregated shock decompositions:

```
// @dynhr:shock_categories
// monetary:      eps_r
// technology:    eps_a
// demand:        eps_pref_
// foreign:       eps_ys, eps_ps, eps_rs
// commodity:     eps_cp
// @dynhr:end
```

Each category maps to one or more shock names.  A shock can belong to
at most one category.  Uncategorised shocks are treated as their own
category.

### 4.5  `@dynhr:narratives`

Declares historical episodes for narrative identification (D17).  Each
episode specifies a date range, a shock category, a variable, and the
expected sign of the shock contribution:

```
// @dynhr:narratives
// gfc_demand:      date_range=2008-Q4:2009-Q2, shock=foreign,
//                  variable=y, sign=negative, min_sd=1.5,
//                  description="GFC global demand collapse"
// covid_demand:    date_range=2020-Q1:2020-Q2, shock=demand,
//                  variable=y, sign=negative, min_sd=2.0,
//                  description="COVID lockdown demand collapse"
// @dynhr:end
```

D17 checks whether, during each episode's date range, the smoothed
shocks from the specified category dominate the historical
decomposition with the expected sign at least `min_sd` standard
deviations above zero.

---

## 5  OccBin / OBC annotations

### 5.1  MCP equation tags

For models with occasionally binding constraints (ZLB, borrowing
limits, etc.), dynher recognises Dynare's `[mcp = '...']` equation
tags.  The syntax follows Dynare's convention:

```
model;
  [mcp = 'r > -0.00401606']
  r = rho_r * r(-1) + (1 - rho_r) * (phi_pi * pi + phi_y * y_gap) + m;
end;
```

The tag format is `mcp = 'VAR OP BOUND'` where `VAR` is the constrained
variable, `OP` is `>` (lower bound) or `<` (upper bound), and `BOUND` is
the numeric bound in model units.

Use `obc_parse_tags()` to extract OBC specs, then pass them to
`solve_obc_lcp()` or `occbin_solve_path()`:

```{r obc-tags, eval = FALSE}
# Extract OBC spec from equation tags
obc_specs <- obc_parse_tags(model)

# Solve with LCP complementarity approach
lcp_result <- solve_obc_lcp(model, compiled, steady$values, params,
                             tags = obc_specs, verbose = FALSE)
```

### 5.2  `occbin_constraints` block

Dynare-style `occbin_constraints` blocks are also parsed:

```
occbin_constraints;
  name 'ZLB'; bind log_R - log(steady_state(R)) < -1e-6;
end;
```

Use `obc_collect_specs()` to merge both tag-based and block-based
constraints:

```{r obc-collect, eval = FALSE}
specs <- obc_collect_specs(model)
```

### 5.3  `model(linear)` requirement

OBC piecewise-linear methods require `model(linear);` in the model
block.  dynhr enforces this with `obc_assert_linear()`.

---

## 6  `planner_objective(...)` block

For Ramsey optimal policy, add a `planner_objective(...)` block to
the `.mod` file (standard Dynare syntax):

```
planner_objective(log(c) + 0.5 * log(l));
```

This is consumed by `ramsey_model()` (augmented-system approach) and
`ramsey_nn1()` (Gross–Hansen (n, n+1) approximation):

```{r ramsey-from-mod, eval = FALSE}
# Both engines automatically detect the planner_objective block
ramsey_result <- ramsey_model(model, compiled, params)
nn1_result    <- ramsey_nn1(model, compiled, params, n = 2L)
```

If the `.mod` file lacks a `planner_objective` block, you can pass the
objective expression directly:

```{r ramsey-explicit, eval = FALSE}
ramsey_result <- ramsey_model(model, compiled, params,
                               planner_objective = "log(c) + 0.5 * log(l)")
```

---

## 7  Variable annotations

Dynare's `%(...)` variable annotations are captured by `parse_mod()`:

```
var y %(long_name='Output gap')
    pi %(long_name='CPI inflation')
    r  %(long_name='Nominal interest rate');
```

These are stored in `model$metadata$annotations` and merged into
`model$metadata$var_labels`, used by diagnostic plots and reports.

---

## 8  Model registry and metadata checking

dynhr provides several tools for inspecting model metadata:

### 8.1  `check_mod_metadata()`

Prints a human-readable summary of all recognised metadata:

```{r check-meta-2}
check_mod_metadata(mod_path)
```

### 8.2  `model_registry()`

Lists all models found in the package's `inst/extdata/models/`
directory, along with their metadata summary:

```{r registry}
reg <- model_registry()
print(reg)
```

### 8.3  `print_model_registry()`

Formatted table of the model registry:

```{r print-reg}
print_model_registry(reg)
```

### 8.4  `extract_mod_metadata()`

Low-level metadata extraction — returns the structured metadata list
without requiring full model parsing:

```{r extract-meta}
meta <- extract_mod_metadata(mod_path)
str(meta, max.level = 2)
```

---

## 9  Injecting external parameters (HA / .mat models)

Winberry-class heterogeneous-agent models and other models with external
steady-state scripts often store dozens of Chebyshev or quadrature parameters
in MATLAB `.mat` files.  These parameters are absent from the `.mod` file
(which only declares them in a `parameters` block), so `parse_mod()` will
warn that they are `NA`.  The two-stage workflow is:

```r
## 1. Parse the .mod file — many params will be NA
m <- parse_mod("winberry_dynhr.mod", verbose = FALSE)

## 2a. Load params from a MATLAB .mat file (requires R.matlab)
#      install.packages("R.matlab")   # once
pars <- read_mat_params("grids.mat")         # named numeric vector

## 2b. Alternative: load from JSON exported from Octave
#  pars <- unlist(jsonlite::fromJSON("grids.json"))

## 3. Inject into the model — warns on extra names, silently ignores them
##    (unlike set_param_values(), which errors on any unknown name)
m <- inject_params(m, pars)
## inject_params: filled 25 previously-NA parameter(s) out of 312 injected.

## 4. Compile, seed the steady state from the external solver, perturbate
comp <- compile_model(m)
ss   <- solve_steady(comp, m$param_values, y0 = external_ss_vector)
dr   <- solve_perturbation(comp, ss, order = 1)
```

**Why `inject_params()` and not `set_param_values()`?**

`set_param_values()` errors if any name in the supplied vector is not a
declared model parameter.  When the `.mat` file contains the full MATLAB
workspace (grid arrays, helper scalars, etc.), the list will contain many
names that are not in the `.mod` `parameters` block.  `inject_params()`
handles this gracefully: it warns about unknown names and drops them, then
fills the matching entries in `model$param_values`.

See `?inject_params` and `?read_mat_params` for full documentation.

---

## 9b  Going the other way: `write_mod()`

Conversion is not one-directional. `write_mod()` serialises a parsed model
back to Dynare `.mod` source, so a model that was built or edited
programmatically in R can be handed to Dynare, diffed against the original,
or checked into version control in its canonical form:

```{r write-mod, eval = FALSE}
model <- parse_mod(mod_path)

# ... modify the model object programmatically ...

write_mod(model, file = "output/model_roundtrip.mod")

# Or capture the source as a character vector without writing:
src <- write_mod(model)
```

A round trip through `parse_mod()` -> `write_mod()` -> `parse_mod()` is the
cheapest available check that dynhr read your file the way you meant it: if
the re-parsed model differs, the disagreement is in the round trip, not in
your head. `header = FALSE` omits the generated-by banner when you want a
byte-comparable diff.

### Initial conditions: `histval` and `smoother2histval`

`histval` blocks are parsed into `model$histval` (lag-indexed), so a
deterministic simulation can start from a supplied history rather than the
steady state. `smoother2histval()` builds that history from a completed
smoother run — the standard way to start a forecast or a counterfactual
from where the data actually left the economy:

```{r histval, eval = FALSE}
sm <- kalman_smoother(Y, dr, model, params, obs_vars = obs)

# Take the smoothed state at the last observed period as the initial history
hv <- smoother2histval(sm, period = nrow(Y), lags = 1L)
```

### Shock groups

`shock_groups` blocks are parsed into `model$shock_groups` and consumed by
`historical_decomposition()` and `realtime_decomposition()`, so the grouping
used in a chart lives with the model rather than in the plotting call. See
`vignette("diagnostics")`.

---

## 10  Completing a conversion checklist

When porting an existing Dynare `.mod` file to dynhr, follow these
steps for full functionality:

| Step | What to add | Why |
|------|-------------|-----|
| 1 | Basic `.mod` file | `parse_mod()` reads it as-is |
| 2 | `@dynhr:expectations` | Quantitative sanity checks in D0–D41 |
| 3 | `@dynhr:benchmarks` | IRF plausibility check (D8) |
| 4 | `@dynhr:variable_labels` | Better plot labels in diagnostics |
| 5 | `@dynhr:shock_categories` | Narrative analysis (D17) |
| 6 | `@dynhr:narratives` | Narrative identification (D17) |
| 7 | `[mcp = '...']` equation tags | OBC solving (OccBin, LCP, PKF) |
| 8 | `planner_objective(...);` | Ramsey optimal policy |
| 9 | `%(long_name='...')` annotations | Richer variable labels |
| 10 | `estimated_params;` block | Bayesian estimation pipeline |

The `03j_no_omega_cp_est.mod` model in `inst/extdata/models/` is a
complete example that uses all metadata block types.  Inspect it for
real-world patterns:

```{r full-example, eval = FALSE}
file.show(system.file("extdata", "models", "03j_no_omega_cp_est.mod",
                      package = "dynhr"))
```

---

## References

Boehl, G. (2020). Efficient solution and computation of models with occasionally binding constraints. *Deutsche Bundesbank Discussion Paper*, 38/2020.

Cottle, R. W., Pang, J.-S., & Stone, R. E. (2009). *The Linear Complementarity Problem*. SIAM.

Guerrieri, L., & Iacoviello, M. (2015). OccBin: A toolkit for solving dynamic models with occasionally binding constraints easily. *Journal of Monetary Economics*, 70, 22-38.

Iskrev, N. (2010). Local identification in DSGE models. *Journal of Monetary Economics*, 57(2), 189-202.

Komunjer, I., & Ng, S. (2011). Dynamic identification of dynamic stochastic general equilibrium models. *Econometrica*, 79(6), 1995-2032.

Mutschler, W. (2015). Identification of DSGE models — The effect of higher-order approximation and pruning. *Journal of Economic Dynamics and Control*, 56, 1-33.

Schmitt-Grohé, S., & Uribe, M. (2004). Solving dynamic general equilibrium models using a second-order approximation to the policy function. *Journal of Economic Dynamics and Control*, 28(4), 755-775.

---

```{r session}
sessionInfo()
```
