---
title: "dynhr .mod syntax reference"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{dynhr .mod syntax reference}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>",
  eval     = TRUE
)
```

## Overview

dynhr reads standard Dynare `.mod` files with `parse_mod()`.  This
vignette enumerates the complete supported syntax, with short verified
examples for each feature, and ends with a clear "Not currently
supported" section.

Every `.mod` snippet shown below has been confirmed to parse without
error under the current package build.

```{r library}
library(dynhr)
```

---

## 1  Declaration blocks

### 1.1  `var` — endogenous variables

```{r var-decl}
m <- parse_mod("
  var y c k;
  varexo e;
  parameters rho; rho = 0.9;
  model(linear);
    y = rho * y(-1) + e;
    c = y;
    k = y;
  end;
")
m$var_names
```

The `var` keyword also accepts:

- **`var(log)`** — marks a variable as log-transformed (Dynare
  `loglinear` convention); the name is still parsed without the `(log)`:

```{r var-log}
m_log <- suppressWarnings(parse_mod("
  var c(log) y;
  varexo e;
  parameters b; b = 0.99;
  model; c = b*c(+1) + e; y = c; end;
"))
m_log$var_names   # c and y, without the (log) suffix
```

- **LaTeX labels** — `var c $c$ y;` attaches a display label; ignored
  by the parser but preserved for downstream renderers.

- **Long-name annotations** — `var y %(long_name='Output');` stores the
  label in `model$metadata`.

### 1.2  `varexo` — exogenous shocks

```{r varexo-decl}
m2 <- parse_mod("
  var y;
  varexo eps_a eps_m;
  parameters rho; rho = 0.8;
  model(linear); y = rho*y(-1) + eps_a + eps_m; end;
")
m2$varexo_names
```

### 1.3  `varexo_det` — deterministic exogenous

`varexo_det` variables are declared identically to `varexo` but are
treated as deterministic.  They appear in `model$varexo_det_names`.

### 1.4  `parameters` — calibrated and estimated parameters

```{r param-decl}
m3 <- suppressWarnings(parse_mod("
  var y;
  varexo e;
  parameters alpha beta;
  alpha = 0.36;
  beta  = 0.99;
  model(linear); y = alpha*y(-1) + e; end;
"))
m3$param_names
m3$param_values
```

Top-level assignments (`alpha = 0.36;`) calibrate parameters.
Assignments may use expressions that reference earlier-declared
parameters: `sigma = 1 / (1 - alpha);`.

### 1.5  `predetermined_variables`

Variables declared `predetermined` are beginning-of-period stocks.
dynhr shifts their timing by -1 in the equation ASTs so that writing
`k` in the model means `k(-1)` in standard Dynare timing.

```{r predet}
m_pre <- suppressWarnings(parse_mod("
  var k y;
  varexo e;
  parameters delta; delta = 0.025;
  predetermined_variables k;
  model;
    y = k^0.36;
    k = (1 - delta) * k + e;
  end;
"))
m_pre$predetermined_vars
```

### 1.6  `varobs` — observed variables

```{r varobs}
m_obs <- suppressWarnings(parse_mod("
  var y c;
  varexo e;
  parameters b; b = 0.99;
  varobs y;
  model(linear); c = b*c(+1) + e; y = c; end;
"))
m_obs$varobs_names
```

---

## 2  The `model` block

### 2.1  Basic nonlinear equations

Equations inside `model; ... end;` are written in standard Dynare
notation.  Each equation ends with a semicolon.

```{r model-basic}
m_nl <- suppressWarnings(parse_mod("
  var c y k;
  varexo e;
  parameters alpha beta delta;
  alpha = 0.36; beta = 0.99; delta = 0.025;
  model;
    1/c = beta * (1/c(+1)) * (alpha * k^(alpha-1) + 1 - delta);
    y = k(-1)^alpha;
    k = (1 - delta) * k(-1) + y - c;
  end;
"))
length(m_nl$equations)
```

### 2.2  `model(linear)` — linearised models

When the model block is declared with `model(linear)`, equations are
treated as already log-linearised (deviations from steady state).
dynhr passes this option through to `model$model_options`:

```{r model-linear}
m_lin <- parse_mod("
  var y pi i;
  varexo eps_r;
  parameters phi kappa rho; phi = 1.5; kappa = 0.1; rho = 0.8;
  model(linear);
    y  = y(+1) - (i - pi(+1));
    pi = kappa * y + pi(+1);
    i  = rho * i(-1) + (1 - rho) * phi * pi + eps_r;
  end;
")
m_lin$model_options
```

### 2.3  Timing convention — leads and lags

- `x(+1)` — one-period lead (next period's expectation)
- `x(-1)` — one-period lag (last period's value)
- `x(+2)`, `x(-3)`, etc. — multi-period leads/lags (dynhr
  automatically introduces auxiliary variables for `|lag| > 1` on
  endogenous variables and for any nonzero timing on exogenous
  variables)

```{r timing}
m_lead <- suppressWarnings(parse_mod("
  var c y;
  varexo e;
  parameters b; b = 0.99;
  model;
    c = b * c(+1) + e;
    y = c;
  end;
"))
# Variable classification reflects the lead on c
m_lead$variable_classification$forward
```

### 2.4  Model-local `#` variables

Inside the `model` block, `#name = expr;` defines a model-local
intermediate variable that is substituted into the equations before
the Jacobian is computed.  These are stored in `model$local_variables`
and are **not** declared as endogenous or exogenous variables.

```{r model-local}
m_local <- parse_mod("
  var c y;
  varexo e;
  parameters alpha; alpha = 0.36;
  model;
    #theta = alpha / (1 - alpha);
    c = theta * y;
    y = c + e;
  end;
")
names(m_local$local_variables)    # 'theta'
length(m_local$var_names)         # 2 (c, y) — theta is NOT endogenous
```

This is one of the most commonly overlooked features: you do **not**
need to inline `#` intermediates before loading into dynhr.  They are
supported natively.

### 2.5  Equation tags

Equations can carry a tag in square brackets immediately before the
equation:

```{r eq-tags}
m_tagged <- parse_mod("
  var y pi i;
  varexo eps_r;
  parameters phi kappa rho; phi = 1.5; kappa = 0.1; rho = 0.8;
  model(linear);
    [name=IS]     y  = y(+1) - (i - pi(+1));
    [name=NKPC]   pi = kappa * y + pi(+1);
    [name=Taylor] i  = rho * i(-1) + (1 - rho) * phi * pi + eps_r;
  end;
")
sapply(m_tagged$equations, function(e) e$tag)
```

Tags are also used for OBC / MCP constraints (see Section 9).

### 2.6  Supported operators and functions

Standard arithmetic: `+`, `-`, `*`, `/`, `^` (power).

The following math functions are recognised in equations:

```{r known-fns, echo = FALSE}
fns <- c(
  "exp, log, ln, log2, log10",
  "sqrt, cbrt, abs, sign",
  "sin, cos, tan, asin, acos, atan",
  "sinh, cosh, tanh",
  "max, min",
  "normcdf, normpdf, erf, erfc",
  "diff, adl",
  "STEADY_STATE() / steady_state()",
  "EXPECTATION()"
)
cat(paste(fns, collapse = "\n"))
```

### 2.7  `STEADY_STATE()` operator

`STEADY_STATE(x)` refers to the steady-state value of `x` and is
recognised in nonlinear models:

```{r steady-state-op}
m_ssop <- suppressWarnings(parse_mod("
  var y c;
  varexo e;
  parameters beta; beta = 0.99;
  model;
    c = beta * c(+1) * (1 + STEADY_STATE(y) - y);
    y = c + e;
  end;
"))
length(m_ssop$equations)
```

---

## 3  `steady_state_model` block

An analytical steady state can be provided as a sequence of
assignments inside `steady_state_model; ... end;`.  Assignments are
evaluated top-to-bottom, so earlier values are available to later
ones.

```{r ssm}
m_ssm <- parse_mod("
  var y c;
  varexo e;
  parameters bet; bet = 0.99;
  model;
    c = bet * c(+1) + e;
    y = c;
  end;
  steady_state_model;
    c = 0;
    y = 0;
  end;
")
length(m_ssm$steady_state_model)   # 2 assignments
```

Canonical fixtures:

- `fs2000.mod` — money-in-utility model with 18-line analytical SS
- `caldara_rp.mod` — Epstein-Zin model with SS that also sets a
  parameter (`nu`)
- `kpr1988.mod` — King-Plosser-Rebelo RBC

When a parameter is assigned inside `steady_state_model`, dynhr warns
that the `.mod` calibration value is a placeholder (the SS value wins
at solve time).

---

## 4  `initval` and `endval` blocks

Initial and terminal values for perfect-foresight simulations or as
starting points for the steady-state solver:

```{r initval}
m_iv <- suppressWarnings(parse_mod("
  var y c;
  varexo e;
  parameters rho; rho = 0.99;
  model;
    c = rho * c(+1) + e;
    y = c;
  end;
  initval;
    c = 1.0;
    y = 1.0;
  end;
"))
m_iv$initval
```

`endval` follows the same syntax and stores into `model$endval`.

---

## 5  `shocks` block

### 5.1  Standard deviations and variances

```{r shocks-stderr}
m_shk <- parse_mod("
  var x y;
  varexo eps_x eps_y;
  parameters a b; a = 0.8; b = 0.5;
  model(linear);
    x = a * x(-1) + eps_x;
    y = b * y(-1) + eps_y;
  end;
  shocks;
    var eps_x; stderr 0.1;
    var eps_y; variance 0.04;
  end;
")
m_shk$shocks$variances[, c("name", "stderr", "variance")]
```

### 5.2  Covariances and correlations

```{r shocks-corr}
m_corr <- parse_mod("
  var x y;
  varexo eps_x eps_y;
  parameters a b; a = 0.8; b = 0.5;
  model(linear);
    x = a * x(-1) + eps_x;
    y = b * y(-1) + eps_y;
  end;
  shocks;
    var eps_x; stderr 0.1;
    var eps_y; stderr 0.2;
    corr eps_x, eps_y = 0.3;
  end;
")
m_corr$shocks$correlations[, c("var1", "var2", "corr")]
```

### 5.3  Multiple `shocks` blocks

dynhr merges multiple `shocks` blocks (later block wins per shock).
`shocks(overwrite)` discards the prior specification entirely before
applying its own entries.  A warning is emitted when two blocks define
conflicting values for the same shock.

---

## 6  `estimated_params` block

Prior distributions for Bayesian estimation follow standard Dynare
syntax.  Five prior families are supported: `beta`, `gamma`,
`inv_gamma`, `normal`, `uniform`.

```{r est-params, eval = FALSE}
# (Not evaluated: requires a more complete model with data)
estimated_params;
  rho,  beta_pdf,     0.5, 0.2;
  sig,  inv_gamma_pdf, 0.1, 2;
  stderr eps_a, inv_gamma_pdf, 0.01, 4;
end;
```

The parsed table is in `model$estimated_params`.

`estimated_params_init` (with optional `use_calibration`) is also
supported and stored in `model$estimated_params_init`.

---

## 7  `@#` macro language

dynhr expands Dynare's macro preprocessor **before** parsing.  The
following directives are supported:

### 7.1  `@#define`

```{r macro-define}
m_def <- parse_mod("
  @#define N = 3
  var x1 x2 x3;
  varexo e;
  parameters rho; rho = 0.9;
  model(linear);
    @#for j in 1:N
    x@{j} = rho * x@{j}(-1) + e;
    @#endfor
  end;
")
m_def$var_names
```

`@#define` also accepts string values and array literals:

```{r macro-array}
m_arr <- parse_mod('
  @#define COUNTRIES = ["H","F"]
  @#for X in COUNTRIES
  var y_@{X};
  @#endfor
  varexo e;
  parameters r; r = 0.7;
  model(linear);
    @#for X in COUNTRIES
    y_@{X} = r * y_@{X}(-1) + e;
    @#endfor
  end;
')
sort(m_arr$var_names)
```

### 7.2  `@#for` / `@#endfor`

Iterates over an explicit list `[a, b, c]`, an integer range `A:B`,
or a `@#define`-d array variable.  Supports `@{expr}` interpolation
including arithmetic on the loop index (e.g. `x@{j-1}`).  Nested
loops are supported.

### 7.3  `@#if` / `@#else` / `@#endif`

Conditional inclusion based on a macro-language expression:

```{r macro-if}
txt_flag <- function(flag) paste(
  sprintf("@#define USE_AR = %d", flag),
  "var x; varexo e; parameters r; r = 0.9;",
  "model(linear);",
  "@#if USE_AR",
  "x = r*x(-1) + e;",
  "@#else",
  "x = e;",
  "@#endif",
  "end;",
  sep = "\n"
)
m_on  <- parse_mod(txt_flag(1L))
m_off <- parse_mod(txt_flag(0L))
dynhr:::ast_to_string(m_on$equations[[1]]$rhs)   # AR(1)
dynhr:::ast_to_string(m_off$equations[[1]]$rhs)  # just e
```

### 7.4  `@#ifdef` / `@#ifndef`

Tests whether a macro variable is defined (or not defined):

```{r macro-ifdef}
m_ifd <- parse_mod('
  @#define MYVAR = 1
  @#ifdef MYVAR
  var y;
  @#endif
  @#ifndef MYVAR
  var z;
  @#endif
  varexo e; parameters p; p = 0.5;
  model; y = p*y(-1) + e; end;
')
m_ifd$var_names   # only y, not z
```

### 7.5  `@#include`

Textually splices another file into the current source.  The path is
resolved relative to the including file's directory.  Nested includes
(up to depth 32) are supported.  Definitions made inside the included
file are visible after the `@#include`.

```{r macro-include, eval = FALSE}
# In parent.mod:
@#include "helper_equations.inc"
```

### 7.6  `@{expr}` interpolation

Inside any non-directive line (including declaration blocks and
equation bodies), `@{expr}` is replaced by the evaluated macro
expression.  Arithmetic on the loop index is supported:
`p@{j-1}_fwrd`.  Quoted string elements have their quotes stripped
when spliced into identifiers.

### 7.7  What is NOT supported in the macro language

- **`@#elseif`** — rewrite as nested `@#if`/`@#else`.
- **`@#echo` / `@#error`** — silently dropped (no structural effect).
- Any other unrecognised `@#` directive causes an **immediate error**
  rather than silent omission.

---

## 8  Comments

Three comment styles are supported, all stripped before parsing:

- `// ...` — C++ single-line comment
- `/* ... */` — C-style block comment (may span lines)
- `% ...` — MATLAB/Dynare percent comment

---

## 9  Other supported blocks and commands

### 9.1  `occbin_constraints` block

```
occbin_constraints;
  name 'ZLB'; bind log_R - log(steady_state(R)) < -1e-6;
end;
```

Parsed into `model$occbin_constraints`.

### 9.2  MCP equation tags

```
model;
  [mcp = 'r > -0.004']
  r = rho_r * r(-1) + (1 - rho_r) * phi_pi * pi + eps_r;
end;
```

Used by `solve_obc_lcp()` and related solvers.

### 9.3  `planner_objective`

```
planner_objective(log(c) + 0.5 * log(l));
```

Stored in `model$planner_objective`.  Used by `ramsey_model()`.

### 9.4  Commands

The following command keywords are extracted (stored in
`model$commands` as a list with `name`, `options`, and `var_list`):

`stoch_simul`, `estimation`, `steady`, `check`, `model_diagnostics`,
`model_info`, `forecast`, `calib_smoother`, `shock_decomposition`,
`conditional_forecast`, `osr`, `ramsey_model`, `ramsey_policy`,
`planner_objective`, `discretionary_policy`,
`perfect_foresight_setup`, `perfect_foresight_solver`,
`extended_path`, `simul`.

### 9.5  `verbatim` blocks

`verbatim; ... end;` blocks contain raw R code that is evaluated into
an environment whose scalar bindings seed parameter values.  This lets
you compute parameters programmatically (e.g. from matrix steady-state
computations) without exposing the intermediate objects as declared
parameters.

### 9.6  dynhr-specific metadata blocks

`// @dynhr:expectations`, `@dynhr:benchmarks`,
`@dynhr:variable_labels`, `@dynhr:shock_categories`,
`@dynhr:narratives`, and `@dynhr:end` comment blocks are extracted by
`extract_mod_metadata()` and drive the 30-diagnostic battery.  They
are plain `//` comments so Dynare ignores them.  See
`vignette("mod-conversion")` for full syntax.

---

## 10  Not currently supported

The following Dynare constructs are **not** parsed or executed by
dynhr.  Using them will not crash `parse_mod()` — the unrecognised
block or option is silently ignored — unless noted otherwise.

| Construct | Status |
|-----------|--------|
| `model_local_variable` block (distinct from `#` inside model) | Not parsed |
| `external_function` declarations | Not parsed |
| `load_mh_file`, `mh_recover` in estimation | Not executed |
| `shock_groups` | Not parsed |
| `heteroskedastic_shocks` scaling (dynhr extension) | Parsed; dynhr-only |
| `filter_tunes` block (dynhr extension) | Parsed; dynhr-only |
| `@#elseif` in macros | Errors loudly — rewrite as nested `@#if` |
| Dynare `use_dll` compilation | Ignored |
| `datafile =` in `estimation(...)` | Option parsed, file loaded by estimation engine |
| `global_initialization` block | Not parsed |

**External `*_steadystate.m` files**: Dynare models that compute
parameter values in an external `<name>_steadystate.m` file will
produce a warning listing the unset parameters.  Use
`inject_params(model, named_vec)` to supply them from R.

**`@#elseif`** causes an explicit error rather than silent behaviour —
this is intentional so that a missing branch is not silently wrong.

---

## 11  Canonical fixture models

The package ships with several reference `.mod` files under
`inst/extdata/models/` that exercise the full supported syntax:

| File | Features |
|------|----------|
| `rbc.mod` | Basic nonlinear RBC, `shocks`, `steady`, `stoch_simul` |
| `nk_canonical.mod` | 3-equation NK, `model(linear)`, `initval` |
| `nk_hs2016.mod` | Linear NK with `#` local vars, `estimated_params` |
| `caldara_rp.mod` | Recursive preferences, `#theta` local var, `steady_state_model` |
| `fs2000.mod` | Money-in-utility, `steady_state_model`, `estimated_params`, aux expansion |
| `kpr1988.mod` | King-Plosser-Rebelo, `steady_state_model` |
| `aikman_3obc.mod` | OBC model, `model(linear)`, `occbin_constraints`, MCP tags |
| `nz_base_nonlinear.mod` | Large nonlinear open-economy, `#` local vars |

```{r fixture-path}
system.file("extdata", "models", "rbc.mod", package = "dynhr")
```

---

## See also

- `?parse_mod` — function documentation with argument details
- `vignette("mod-conversion")` — converting Dynare models and using
  dynhr-specific annotations
- `vignette("solving")` — perturbation solver after parsing
- `vignette("estimation")` — Bayesian estimation workflow
