---
title: "dynhr .mod syntax reference"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{dynhr .mod syntax reference}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>",
  eval     = TRUE
)
```

## Overview

dynhr reads standard Dynare `.mod` files with `parse_mod()`.  This
vignette enumerates the complete supported syntax, with short verified
examples for each feature, and ends with a clear "Not currently
supported" section.

Every `.mod` snippet shown below has been confirmed to parse without
error under the current package build.

```{r library}
library(dynhr)
```

---

## 1  Declaration blocks

### 1.1  `var` — endogenous variables

```{r var-decl}
m <- parse_mod("
  var y c k;
  varexo e;
  parameters rho; rho = 0.9;
  model(linear);
    y = rho * y(-1) + e;
    c = y;
    k = y;
  end;
")
m$var_names
```

The `var` keyword also accepts:

- **`var(log)`** — Dynare 6+ semantics.  `var(log) c;` declares `c`
  and also creates the endogenous variable `LOG_c`: every `c` in the
  model block (any timing, `#` locals included) is replaced by
  `exp(LOG_c)` and the equation `c = exp(LOG_c);` is appended, so the
  perturbation is taken in `LOG_c` while `c` stays available.  A
  `steady_state_model` that assigns `c` gets `LOG_c = log(c);` appended,
  and `initval`/`endval`/`histval` values of `c` carry over as logs.
  `model$log_vars` maps each declared name to its `LOG_` variable:

```{r var-log}
m_log <- parse_mod("
  var(log) c;
  var y;
  varexo e;
  parameters b; b = 0.5;
  model; c = exp(b*log(c(-1)) + e); y = c; end;
  steady_state_model; c = 1; y = 1; end;
")
m_log$var_names   # c, y, then LOG_c
m_log$log_vars
```

  At first order the level rows of the decision rule are the same as
  without `var(log)`; `LOG_c` carries the log-linear rule, and at
  higher orders the approximation is taken in logs.  A declared name
  `LOG_c` clashes and is an error.

- **LaTeX labels** — `var c $c$ y;` attaches a display label; ignored
  by the parser but preserved for downstream renderers.

- **Long-name annotations** — `var y %(long_name='Output');` stores the
  label in `model$metadata`.

### 1.2  `varexo` — exogenous shocks

```{r varexo-decl}
m2 <- parse_mod("
  var y;
  varexo eps_a eps_m;
  parameters rho; rho = 0.8;
  model(linear); y = rho*y(-1) + eps_a + eps_m; end;
")
m2$varexo_names
```

### 1.3  `varexo_det` — deterministic exogenous

`varexo_det` variables are declared identically to `varexo` but are
treated as deterministic.  They appear in `model$varexo_det_names`.

### 1.4  `parameters` — calibrated and estimated parameters

```{r param-decl}
m3 <- suppressWarnings(parse_mod("
  var y;
  varexo e;
  parameters alpha beta;
  alpha = 0.36;
  beta  = 0.99;
  model(linear); y = alpha*y(-1) + e; end;
"))
m3$param_names
m3$param_values
```

Top-level assignments (`alpha = 0.36;`) calibrate parameters.
Assignments may use expressions that reference earlier-declared
parameters: `sigma = 1 / (1 - alpha);`.

### 1.5  `predetermined_variables`

Variables declared `predetermined` are beginning-of-period stocks.
dynhr shifts their timing by -1 so that writing `k` in the model means
`k(-1)` in standard Dynare timing.  As in Dynare's preprocessor, the
shift is applied to the model-block text (`#` locals included) before
`EXPECTATION` and the auxiliary lead/lag variables are built.

```{r predet}
m_pre <- suppressWarnings(parse_mod("
  var k y;
  varexo e;
  parameters delta; delta = 0.025;
  predetermined_variables k;
  model;
    y = k^0.36;
    k = (1 - delta) * k + e;
  end;
"))
m_pre$predetermined_vars
```

### 1.6  `varobs` — observed variables

```{r varobs}
m_obs <- suppressWarnings(parse_mod("
  var y c;
  varexo e;
  parameters b; b = 0.99;
  varobs y;
  model(linear); c = b*c(+1) + e; y = c; end;
"))
m_obs$varobs_names
```

---

## 2  The `model` block

### 2.1  Basic nonlinear equations

Equations inside `model; ... end;` are written in standard Dynare
notation.  Each equation ends with a semicolon.

```{r model-basic}
m_nl <- suppressWarnings(parse_mod("
  var c y k;
  varexo e;
  parameters alpha beta delta;
  alpha = 0.36; beta = 0.99; delta = 0.025;
  model;
    1/c = beta * (1/c(+1)) * (alpha * k^(alpha-1) + 1 - delta);
    y = k(-1)^alpha;
    k = (1 - delta) * k(-1) + y - c;
  end;
"))
length(m_nl$equations)
```

### 2.2  `model(linear)` — linearised models

When the model block is declared with `model(linear)`, equations are
treated as already log-linearised (deviations from steady state).
dynhr passes this option through to `model$model_options`:

```{r model-linear}
m_lin <- parse_mod("
  var y pi i;
  varexo eps_r;
  parameters phi kappa rho; phi = 1.5; kappa = 0.1; rho = 0.8;
  model(linear);
    y  = y(+1) - (i - pi(+1));
    pi = kappa * y + pi(+1);
    i  = rho * i(-1) + (1 - rho) * phi * pi + eps_r;
  end;
")
m_lin$model_options
```

### 2.3  Timing convention — leads and lags

- `x(+1)` — one-period lead (next period's expectation)
- `x(-1)` — one-period lag (last period's value)
- `x(+2)`, `x(-3)`, etc. — multi-period leads/lags (dynhr
  automatically introduces auxiliary variables for `|lag| > 1` on
  endogenous variables and for any nonzero timing on exogenous
  variables)

```{r timing}
m_lead <- suppressWarnings(parse_mod("
  var c y;
  varexo e;
  parameters b; b = 0.99;
  model;
    c = b * c(+1) + e;
    y = c;
  end;
"))
# Variable classification reflects the lead on c
m_lead$variable_classification$forward
```

### 2.4  Model-local `#` variables

Inside the `model` block, `#name = expr;` defines a model-local
intermediate variable that is substituted into the equations before
the Jacobian is computed.  These are stored in `model$local_variables`
and are **not** declared as endogenous or exogenous variables.

```{r model-local}
m_local <- parse_mod("
  var c y;
  varexo e;
  parameters alpha; alpha = 0.36;
  model;
    #theta = alpha / (1 - alpha);
    c = theta * y;
    y = c + e;
  end;
")
names(m_local$local_variables)    # 'theta'
length(m_local$var_names)         # 2 (c, y) — theta is NOT endogenous
```

This is one of the most commonly overlooked features: you do **not**
need to inline `#` intermediates before loading into dynhr.  They are
supported natively.

As in Dynare 7, a local may be used with a lead or a lag.  `r_real(-1)`
stands for the definition of `r_real` with every variable shifted back
one period (`i(-1) - pi` below), and leads or lags beyond one period
that this creates get auxiliary variables like any other:

```{r model-local-timing}
m_lt <- parse_mod("
  var y i pi;
  varexo e;
  parameters phi; phi = 1.5;
  model;
    # r_real = i - pi(+1);
    y = 0.9*y(-1) - 0.1*r_real(-1) + e;
    pi = 0.99*pi(+1) + 0.1*y;
    i = phi*pi;
  end;
")
dynhr:::ast_to_string(m_lt$equations[[1]]$rhs)
```

### 2.5  Equation tags

Equations can carry a tag in square brackets immediately before the
equation:

```{r eq-tags}
m_tagged <- parse_mod("
  var y pi i;
  varexo eps_r;
  parameters phi kappa rho; phi = 1.5; kappa = 0.1; rho = 0.8;
  model(linear);
    [name=IS]     y  = y(+1) - (i - pi(+1));
    [name=NKPC]   pi = kappa * y + pi(+1);
    [name=Taylor] i  = rho * i(-1) + (1 - rho) * phi * pi + eps_r;
  end;
")
sapply(m_tagged$equations, function(e) e$tag)
```

Tags are also used for OBC / MCP constraints (see Section 9).

### 2.6  Supported operators and functions

Standard arithmetic: `+`, `-`, `*`, `/`, `^` (power).

The following math functions are recognised in equations:

```{r known-fns, echo = FALSE}
fns <- c(
  "exp, log, ln, log2, log10",
  "sqrt, cbrt, abs, sign",
  "sin, cos, tan, asin, acos, atan",
  "sinh, cosh, tanh",
  "max, min",
  "normcdf, normpdf, erf, erfc",
  "diff(expr) -- expr - expr(-1), expanded inline (adl() is not a Dynare 7 operator and is an error)",
  "STEADY_STATE() / steady_state()",
  "EXPECTATION(k)(expr) -- expectation of expr on the period t+k information set (Dynare semantics, via an auxiliary variable)"
)
cat(paste(fns, collapse = "\n"))
```

### 2.7  `STEADY_STATE()` operator

`STEADY_STATE(x)` refers to the steady-state value of `x` and is
recognised in nonlinear models:

```{r steady-state-op}
m_ssop <- suppressWarnings(parse_mod("
  var y c;
  varexo e;
  parameters beta; beta = 0.99;
  model;
    c = beta * c(+1) * (1 + STEADY_STATE(y) - y);
    y = c + e;
  end;
"))
length(m_ssop$equations)
```

---

## 3  `steady_state_model` block

An analytical steady state can be provided as a sequence of
assignments inside `steady_state_model; ... end;`.  Assignments are
evaluated top-to-bottom, so earlier values are available to later
ones.

```{r ssm}
m_ssm <- parse_mod("
  var y c;
  varexo e;
  parameters bet; bet = 0.99;
  model;
    c = bet * c(+1) + e;
    y = c;
  end;
  steady_state_model;
    c = 0;
    y = 0;
  end;
")
length(m_ssm$steady_state_model)   # 2 assignments
```

Verified against these development fixtures (not shipped — see
§11):

- `fs2000.mod` — money-in-utility model with 18-line analytical SS
- `caldara_rp.mod` — Epstein-Zin model with SS that also sets a
  parameter (`nu`)
- `kpr1988.mod` — King-Plosser-Rebelo RBC

When a parameter is assigned inside `steady_state_model`, dynhr warns
that the `.mod` calibration value is a placeholder (the SS value wins
at solve time).

---

## 4  `initval` and `endval` blocks

Initial and terminal values for perfect-foresight simulations or as
starting points for the steady-state solver:

```{r initval}
m_iv <- suppressWarnings(parse_mod("
  var y c;
  varexo e;
  parameters rho; rho = 0.99;
  model;
    c = rho * c(+1) + e;
    y = c;
  end;
  initval;
    c = 1.0;
    y = 1.0;
  end;
"))
m_iv$initval
```

`endval` follows the same syntax and stores into `model$endval`.

---

## 5  `shocks` block

### 5.1  Standard deviations and variances

```{r shocks-stderr}
m_shk <- parse_mod("
  var x y;
  varexo eps_x eps_y;
  parameters a b; a = 0.8; b = 0.5;
  model(linear);
    x = a * x(-1) + eps_x;
    y = b * y(-1) + eps_y;
  end;
  shocks;
    var eps_x; stderr 0.1;
    var eps_y; variance 0.04;
  end;
")
m_shk$shocks$variances[, c("name", "stderr", "variance")]
```

### 5.2  Covariances and correlations

```{r shocks-corr}
m_corr <- parse_mod("
  var x y;
  varexo eps_x eps_y;
  parameters a b; a = 0.8; b = 0.5;
  model(linear);
    x = a * x(-1) + eps_x;
    y = b * y(-1) + eps_y;
  end;
  shocks;
    var eps_x; stderr 0.1;
    var eps_y; stderr 0.2;
    corr eps_x, eps_y = 0.3;
  end;
")
m_corr$shocks$correlations[, c("var1", "var2", "corr")]
```

### 5.3  Multiple `shocks` blocks

dynhr merges multiple `shocks` blocks (later block wins per shock).
`shocks(overwrite)` discards the prior specification entirely before
applying its own entries.  A warning is emitted when two blocks define
conflicting values for the same shock.

### 5.4  Deterministic scenarios: `periods`/`values` and `shock_paths`

A `shocks` block stanza `var x; periods 1:3; values 0.1;` stores a
deterministic path in `model$det_shocks` (one row per period).  Dynare
7's `shock_paths` block is parsed into the same structure, with the
values at the terminal period `end` going to `model$endval`.  Values
are expressions evaluated period by period, in stanza order, exactly
as Dynare does: `self.x(-1)` is the value just set for the previous
period, `self.z` another shock's current value, `initval.x` the
`initval` value, and parameters are used by name.  A range ending at
`end` needs the simulation length from
`perfect_foresight_setup(periods = T)`; dates need its
`first_simulation_period`.

```{r shock-paths}
m_sp <- parse_mod("
  var y;
  varexo x;
  parameters rho; rho = 0.5;
  model; y = rho*y(-1) + x + 0.01*y(+1); end;
  initval; x = 0.2; end;
  shock_paths;
    var x;
    periods 1, 2:4, 6:end;
    values initval.x*1.05, self.x(-1)*1.05, 0.3;
  end;
  perfect_foresight_setup(periods = 8);
")
m_sp$det_shocks
m_sp$endval
```

Not implemented (each an error, never a silent approximation):
database scopes (`db.foo`), the `prev.` / `learnt_in(...)` scopes and
`learnt_in` other than 1 (expectation errors), and combining
`shock_paths` with an `endval` block or deterministic `shocks` periods,
which Dynare forbids.  `write_mod()` writes the scenario back as a
`shock_paths` block with explicit periods and values.

#### Controlled paths: `perfect_foresight_controlled_paths`

Dynare 7's `perfect_foresight_controlled_paths` block (and the same
`exogenize` / `endogenize` stanzas inside a `shock_paths` block) imposes
the path of an endogenous variable over given periods and solves for
the path of a chosen shock instead, so the stacked system stays square.
Each stanza is `exogenize VAR; periods ...; values ...; endogenize
SHOCK;`; periods are integers, ranges or dates (not `end`), values are
numbers or parameter expressions (plus `initval.` inside `shock_paths`,
never `self.`).  The parser stores one row per controlled period in
`model$controlled_paths`, and `perfect_foresight_solve()` uses it by
default: in those periods the Newton unknown and Jacobian column of the
endogenous variable are swapped for the shock's, as in Dynare's
`controlled_paths_substitute_stacked_jacobian.m`.  The solved shocks are
returned in `$exo_path`.  A variable exogenized twice, or a shock
endogenized twice, in one period is an error, as is endogenizing a shock
in a period where it also has a deterministic value; only `learnt_in = 1`
is implemented.

```{r controlled-paths}
m_cp <- parse_mod("
  var y c;
  varexo e u;
  parameters rho a; rho = 0.5; a = 0.3;
  model;
    y = rho*y(-1) + e;
    c = a*c(+1) + y + u;
  end;
  perfect_foresight_controlled_paths;
    exogenize y;
    periods 1:2, 3;
    values 0.5, 0.2;
    endogenize e;
  end;
  perfect_foresight_setup(periods = 20);
")
m_cp$controlled_paths
cm_cp <- compile_model(m_cp, verbose = FALSE)
z <- c(y = 0, c = 0)
pf <- perfect_foresight_solve(cm_cp, y0 = z, y_terminal = z, n_periods = 20)
pf$Y[1:4, ]
pf$exo_path[1:4, ]      # e = y - rho * y(-1) in periods 1-3
```

---

## 6  `estimated_params` block

Prior distributions for Bayesian estimation follow standard Dynare
syntax.  Five prior families are supported: `beta`, `gamma`,
`inv_gamma`, `normal`, `uniform`.

```{r est-params, eval = FALSE}
# (Not evaluated: requires a more complete model with data)
estimated_params;
  rho,  beta_pdf,     0.5, 0.2;
  sig,  inv_gamma_pdf, 0.1, 2;
  stderr eps_a, inv_gamma_pdf, 0.01, 4;
end;
```

The parsed table is in `model$estimated_params`.

For every shape, `P1` and `P2` (the two numbers after the shape) are the
prior **mean** and **standard deviation**, as in Dynare. This includes
`uniform_pdf`: `x, uniform_pdf, 0.5, 0.2;` is a uniform with mean 0.5 and
standard deviation 0.2, i.e. on `[0.5 - sqrt(3)*0.2, 0.5 + sqrt(3)*0.2]`
(Dynare's `uniform_specification()`). To give a uniform's **bounds**, leave
`P1`/`P2` empty and put the bounds in `P3`/`P4`:

```
estimated_params;
  x, uniform_pdf, , , 0, 1;             // uniform on [0, 1]
  y, 0.5, 0, 1, uniform_pdf, , , 0, 1;  // same, long form with INITVAL, LB, UB
end;
```

`uniform_pdf, 0, 1` is therefore **not** the unit interval (it is mean 0, sd 1,
i.e. `[-1.73, 1.73]`). `extract_prior_spec()` warns
(`dynhr_warning_uniform_parameterisation`) when a mean/sd uniform row looks
like it was meant as bounds -- `P2 > P1 >= 0`, a calibrated/initial value
inside `[P1, P2]` but outside the implied support, or a support below 0 for a
`stderr` or outside `[-1, 1]` for a `corr`. Rows written with `P3`/`P4` never
warn.

`estimated_params_init` (with optional `use_calibration`) is also
supported and stored in `model$estimated_params_init`.

---

## 7  `@#` macro language

dynhr expands Dynare's macro preprocessor **before** parsing.  The
following directives are supported:

### 7.1  `@#define`

```{r macro-define}
m_def <- parse_mod("
  @#define N = 3
  var x1 x2 x3;
  varexo e;
  parameters rho; rho = 0.9;
  model(linear);
    @#for j in 1:N
    x@{j} = rho * x@{j}(-1) + e;
    @#endfor
  end;
")
m_def$var_names
```

`@#define` also accepts string values and array literals:

```{r macro-array}
m_arr <- parse_mod('
  @#define COUNTRIES = ["H","F"]
  @#for X in COUNTRIES
  var y_@{X};
  @#endfor
  varexo e;
  parameters r; r = 0.7;
  model(linear);
    @#for X in COUNTRIES
    y_@{X} = r * y_@{X}(-1) + e;
    @#endfor
  end;
')
sort(m_arr$var_names)
```

### 7.2  `@#for` / `@#endfor`

Iterates over an explicit list `[a, b, c]`, an integer range `A:B`,
or a `@#define`-d array variable.  Supports `@{expr}` interpolation
including arithmetic on the loop index (e.g. `x@{j-1}`).  Nested
loops are supported.

As in Dynare 6+, `@#for` also unpacks tuples (`@#for (i, c) in
[(1, "H"), (2, "F")]`), takes a `when` filter (`@#for k in 1:6 when
mod(k, 3) == 0`) and a stepped range (`1:2:7`).

### 7.3  `@#if` / `@#elseif` / `@#else` / `@#endif`

Conditional inclusion based on a macro-language expression.  Any
number of `@#elseif COND` clauses may follow `@#if`, `@#ifdef` or
`@#ifndef`; the first true branch is taken:

```{r macro-if}
txt_flag <- function(flag) paste(
  sprintf("@#define USE_AR = %d", flag),
  "var x; varexo e; parameters r; r = 0.9;",
  "model(linear);",
  "@#if USE_AR",
  "x = r*x(-1) + e;",
  "@#else",
  "x = e;",
  "@#endif",
  "end;",
  sep = "\n"
)
m_on  <- parse_mod(txt_flag(1L))
m_off <- parse_mod(txt_flag(0L))
dynhr:::ast_to_string(m_on$equations[[1]]$rhs)   # AR(1)
dynhr:::ast_to_string(m_off$equations[[1]]$rhs)  # just e
```

```{r macro-elseif}
txt_mode <- function(mode) paste(
  sprintf('@#define mode = "%s"', mode),
  '@#define y = "y_" + mode',          # string concatenation
  "var @{y}; varexo e; parameters r; r = 0.5;",
  "model;",
  '@#if mode == "ar"',
  "@{y} = r*@{y}(-1) + e;",
  '@#elseif mode == "ma"',
  "@{y} = e + r*e(-1);",
  "@#else",
  "@{y} = e;",
  "@#endif",
  "end;",
  sep = "\n"
)
parse_mod(txt_mode("ma"))$var_names[1]   # y_ma
```

### 7.4  `@#ifdef` / `@#ifndef`

Tests whether a macro variable is defined (or not defined):

```{r macro-ifdef}
m_ifd <- parse_mod('
  @#define MYVAR = 1
  @#ifdef MYVAR
  var y;
  @#endif
  @#ifndef MYVAR
  var z;
  @#endif
  varexo e; parameters p; p = 0.5;
  model; y = p*y(-1) + e; end;
')
m_ifd$var_names   # only y, not z
```

### 7.5  `@#include`

Textually splices another file into the current source.  The path is
resolved relative to the including file's directory.  Nested includes
(up to depth 32) are supported.  Definitions made inside the included
file are visible after the `@#include`.

```{r macro-include, eval = FALSE}
# In parent.mod:
@#include "helper_equations.inc"
```

The file name may also be a string-valued macro expression
(`@#include FILE`), and `@#includepath "dir"` adds a directory that is
searched after the including file's own.

### 7.6  `@{expr}` interpolation

Inside any non-directive line (including declaration blocks and
equation bodies), `@{expr}` is replaced by the evaluated macro
expression.  Arithmetic on the loop index is supported:
`p@{j-1}_fwrd`.  Quoted string elements have their quotes stripped
when spliced into identifiers.  Booleans are spliced as `true`/`false`,
arrays as `[a, b]` and tuples as `(a, b)`, as Dynare prints them.

### 7.7  Macro expressions

Macro expressions follow the Dynare 6/7 macro language (checked
against the Dynare 7.1 preprocessor's `savemacro` output):

- values: reals, strings, booleans (`true`/`false`), arrays `[a, b]`,
  tuples `(a, b)`, and functions `@#define f(x) = 2*x + 1`;
  `@#define NAME` with no value is `true`;
- operators: `+ - * / ^`, comparisons, `&& || !`, `in`, ranges `A:B`
  and `A:S:B`; `+` concatenates strings and arrays, `|` / `&` / `-` are
  union / intersection / difference of arrays, `*` and `^` build
  cartesian products;
- comprehensions `[x^2 for x in 1:5 when mod(x, 2) == 1]`, casts
  `(bool)`, `(real)`, `(string)`, `(tuple)`, `(array)`, and indexing
  `v[2]`, `v[[1, 3]]`, `v[2:3]` (also on strings);
- functions `length`, `isempty`, `isboolean`, `isreal`, `isstring`,
  `istuple`, `isarray`, `defined`, `exp`, `ln`, `log`, `log10`, the
  trigonometric functions, `sqrt`, `cbrt`, `sign`, `floor`, `ceil`,
  `trunc`, `round`, `erf`, `erfc`, `gamma`, `lgamma`, `sum`, `mod`,
  `max`, `min`, `normpdf`, `normcdf`.

Directive lines may end in a `//` comment.  `@#error` aborts when it is
reached in a taken branch; `@#echo` and `@#echomacrovars` are ignored.
The expressions are evaluated by dynhr's own interpreter, never as R
code.  One deliberate difference: a non-integer real is spliced with the
fewest digits that read back as the same double (`1/3` gives
`0.3333333333333333`), where Dynare prints 15 significant digits.

Any unrecognised `@#` directive causes an **immediate error** rather
than silent omission.

---

## 8  Comments

Three comment styles are supported, all stripped before parsing:

- `// ...` — C++ single-line comment
- `/* ... */` — C-style block comment (may span lines)
- `% ...` — MATLAB/Dynare percent comment

---

## 9  Other supported blocks and commands

### 9.1  `occbin_constraints` block

```
occbin_constraints;
  name 'ZLB'; bind log_R - log(steady_state(R)) < -1e-6;
end;
```

Parsed into `model$occbin_constraints`.

### 9.2  Complementarity conditions and MCP equation tags

Dynare 7 writes a complementarity condition after its equation,
separated by the perpendicular symbol `⟂` (U+27C2) or the ASCII `_|_`.
Both bounds may be given at once, and a bound may be any function of
parameters:

```{r complementarity}
m_cc <- parse_mod("
  var mu i y;
  varexo e;
  parameters alpha; alpha = 0.5;
  model;
    mu = 0 _|_ 0 < i < 1 + 2*alpha;
    i = 1.5*y + mu;
    y = 0.9*y(-1) + e;
  end;
")
m_cc$equations[[1]]$complementarity
sapply(mcp_parse_tags(m_cc), function(s) c(s$op, s$bound))
```

The constrained side must be a bare endogenous variable and the bounds
may not contain any variable (as in Dynare's preprocessor); `x > L`,
`L < x`, `x < U`, `U > x`, `L < x < U` and `U > x > L` are accepted,
with `>=` / `<=` meaning the same.  A two-sided condition gives one
lower and one upper MCP spec on the same equation, and bounds that are
functions of parameters are re-evaluated at the parameter vector the
solver is called with.  `write_mod()` writes the condition back with
`_|_`.

The older tag form is still accepted (Dynare 7 deprecates it):

```
model;
  [mcp = 'r > -0.004']
  r = rho_r * r(-1) + (1 - rho_r) * phi_pi * pi + eps_r;
end;
```

Used by `solve_obc_lcp()` and related solvers.

### 9.3  `planner_objective`

```
planner_objective(log(c) + 0.5 * log(l));
```

Stored in `model$planner_objective`.  Used by `ramsey_model()`.

### 9.4  Commands

The following command keywords are extracted (stored in
`model$commands` as a list with `name`, `options`, and `var_list`):

`stoch_simul`, `estimation`, `steady`, `check`, `model_diagnostics`,
`model_info`, `forecast`, `calib_smoother`, `shock_decomposition`,
`conditional_forecast`, `osr`, `ramsey_model`, `ramsey_policy`,
`planner_objective`, `discretionary_policy`,
`perfect_foresight_setup`, `perfect_foresight_solver`,
`extended_path`, `simul`.

Options are split on top-level commas only, so a list value is read
whole: `discretionary_policy(instruments=(i,tau), planner_discount=0.99)`
gives `options$instruments == c("i", "tau")`, `irf_shocks=(e,u)` a
character vector, `bandpass_filter=[6 32]` or
`conditional_variance_decomposition=(1:4)` a numeric vector; a MATLAB
cell such as `optim=('MaxIter',200)` is kept as its text.
`discretionary_policy(model)` takes its instruments and discount from
the `discretionary_policy` command, and `osr()` its planner discount
from the `osr` command.  With a `planner_objective` and no
`optim_weights` block, `osr()` minimises Dynare 7's welfare loss
`-E[W] = -(U + 0.5 * sum(U_yy * Var(y))) / (1 - planner_discount)`
(see `?osr`).

### 9.4b  `ramsey_constraints` block

```
ramsey_model(instruments=(i), planner_discount=0.99);
ramsey_constraints;
  i > 0;
end;
```

Bounds a variable of the Ramsey problem by a parameter expression
(`>`/`>=` lower, `<`/`<=` upper).  As in Dynare 7, each bound is
complementary to the Ramsey first-order condition with respect to its
variable and is used in perfect foresight: `ramsey_obc_pf()` imposes it
(in a binding period the FOC is replaced by `i = 0`), reproducing
`perfect_foresight_solver(lmmcp)`.  Stochastic Ramsey solutions ignore
it, as Dynare's do.  Stored in `model$ramsey_constraints`.

### 9.5  `observation_trends` block

Adds a deterministic linear trend to the measurement equation of an
observed variable, exactly as Dynare does:

```
varobs dy pinfobs;
observation_trends;
  dy (ctrend/100);
end;
estimation(datafile = data, first_obs = 71);
```

With this block the filter and smoother compare observation `t` (the
`t`-th row of the data you pass) with
`ys[dy] + (ctrend/100) * (first_obs + t - 1)`: the trend index of the
first data row is `estimation(first_obs = ...)` when present, 1
otherwise, so pass the data starting at row `first_obs` as Dynare
does.  The slope may be any expression in declared parameters and is
re-evaluated at every parameter vector, so an estimated slope moves the
likelihood.  An entry for a variable that is not in `varobs`, a slope
that uses anything but parameters, or a date-valued `first_obs` is an
error.  Stored in `model$observation_trends` (`$trends`, `$first_obs`)
and honoured by `kalman_filter()`, `kalman_smoother()` and the Gaussian
`make_log_posterior()`.

```{r obs-trends}
m_ot <- parse_mod("
  var y; varexo e; parameters rho g; rho = 0.5; g = 0.02;
  model; y = rho*y(-1) + e; end;
  shocks; var e; stderr 1; end;
  varobs y;
  observation_trends; y (g); end;
")
m_ot$observation_trends
```

**`deterministic_trends`** has the same entry syntax but may name any
declared endogenous variable.  It is stored in
`model$deterministic_trends`; each entry for an observed variable (every
entry when there is no `varobs`) is a trend of that observable exactly as
if it were in `observation_trends` (it is merged into
`model$observation_trends`).  A variable in both blocks is an error.
Dynare 7.1 itself stores the block in `options_.deterministic_trend_coeffs`,
which none of its routines reads, and its driver fails on the statement.

### 9.5a  Nonstationary models: `trend_var` and `var(deflator=)`

A model may be written in trending variables, as in Dynare:

- `trend_var(growth_factor = G) A;` declares a trend variable with
  `A_t = G_t * A_{t-1}`; `log_trend_var(log_growth_factor = g) LA;` an
  additive one, `LA_t = g_t + LA_{t-1}`.  `G` may use parameters and
  endogenous variables (not other trend variables).
- `var(deflator = D) Y;` declares `Y` as trending with the
  multiplicative trend `D` (an expression in trend variables, endogenous
  variables and parameters, e.g. `A` or `A^2`); `var(log_deflator = D) Y;`
  with the additive trend `D`.

`parse_mod()` rewrites the model into its stationary form the way
Dynare's preprocessor does (right after the `predetermined_variables`
re-timing): `Y(k)` becomes `Y(k)*D(k)` (`Y(k)+D(k)`), a trend variable at a
lead or lag becomes `A*G(+1)*...*G(+k)` or `A/(G*G(-1)*...*G(-k+1))`
(sums for an additive trend), and every trend variable then becomes 1
(additive: 0).  So **every declared name is the detrended variable**:
the steady state, `steady_state_model`, `initval` and the decision rule
are those of the stationary model, exactly as in Dynare.  The rewrite is
valid only on a balanced growth path; an equation that still depends on
the level of a trend after detrending is an error of class
`dynhr_error_unbalanced_growth`.  The declarations are kept in
`model$nonstationary`; `write_mod()` writes the detrended model and does
not re-declare the trends.

```{r nonstationary}
m_ns <- parse_mod("
  parameters alpha beta delta gA;
  alpha = 0.33; beta = 0.99; delta = 0.025; gA = 1.005;
  trend_var(growth_factor = gA) A;
  var(deflator = A) Y C K;
  var z; varexo e;
  model;
    Y = exp(z)*K(-1)^alpha*A^(1-alpha);
    K = (1-delta)*K(-1) + Y - C;
    1/C = beta/C(+1)*(alpha*Y(+1)/K + 1 - delta);
    z = 0.9*z(-1) + e;
  end;
")
## the detrended capital accumulation: K = (1-delta)*K(-1)/gA + Y - C
dynhr:::ast_to_string(m_ns$equations[[2]]$rhs)
```

### 9.6  `verbatim` blocks

`verbatim; ... end;` blocks contain raw R code that is evaluated into
an environment whose scalar bindings seed parameter values.  This lets
you compute parameters programmatically (e.g. from matrix steady-state
computations) without exposing the intermediate objects as declared
parameters.

### 9.7  dynhr-specific metadata blocks

`// @dynhr:expectations`, `@dynhr:benchmarks`,
`@dynhr:variable_labels`, `@dynhr:shock_categories`,
`@dynhr:narratives`, and `@dynhr:end` comment blocks are extracted by
`extract_mod_metadata()` and drive the D0–D41 diagnostic battery.  They
are plain `//` comments so Dynare ignores them.  See
`vignette("mod-conversion")` for full syntax.

---

## 10  Not currently supported

The following Dynare constructs are **not**, or only partly, parsed
or executed by dynhr.  Using them does not crash `parse_mod()` unless
noted otherwise.  (`deterministic_trends`, `trend_var` and
`var(deflator=...)`, once stripped with a warning, are implemented: see
§9.5 and §9.5a.)

| Construct | Status |
|-----------|--------|
| `model_local_variable` block (distinct from `#` inside model) | Not parsed |
| `external_function` declarations | Not parsed |
| `load_mh_file`, `mh_recover` in estimation | Not executed |
| `heteroskedastic_shocks` | `scales` parsed, with Dynare's period semantics: an integer period indexes the dataset (sample period = `p - first_obs + 1`), a date (`2020Q1`, Dynare 7) is resolved against a date-valued `estimation(first_obs=)`; a date with no date index is a `dynhr_error_mod_date_unresolved` error. The `values` statement is not supported |
| `shock_paths` `database` scopes, `learnt_in` other than 1 | Errors; see §5.4 |
| `filter_tunes` block (dynhr extension) | Parsed; dynhr-only |
| Dynare `use_dll` compilation | Ignored |
| `datafile =` in `estimation(...)` | Option parsed, file loaded by estimation engine |
| `global_initialization` block | Not parsed |
| `trend_var` / `log_trend_var` with a growth factor that uses a trend variable (I(2) trends) | Error |
| `adl()` | Error: not an operator of Dynare 7 (the 7.1 preprocessor rejects it too) — write the lags out |

`model_replace`, `model_remove` and `var_remove` (Dynare 6+) are
implemented with Dynare's semantics: `model_remove('eq1', key =
'value');` removes the tagged equations, and a variable whose defining
equation was removed becomes exogenous if the model still uses it
(dropped otherwise); `model_replace('eq1'); ... end;` swaps the tagged
equations for new ones appended at the end of the model; `var_remove
a b;` removes declared symbols.  See `?parse_mod`.

**External `*_steadystate.m` files**: Dynare models that compute
parameter values in an external `<name>_steadystate.m` file will
produce a warning listing the unset parameters.  Use
`inject_params(model, named_vec)` to supply them from R.

---

## 11  Shipped fixture models

These `.mod` files ship under `inst/extdata/models/`, and between them
exercise the supported syntax:

| File | Features |
|------|----------|
| `rbc.mod` | Basic nonlinear RBC, `shocks`, `initval`, `stoch_simul` |
| `rbc2shock.mod` | The same with two shocks (variance decomposition, multi-shock filtering) |
| `nk_demo.mod` | Nonlinear 3-equation NK, `varobs`, `estimated_params` — the estimation walkthrough's fixture, with simulated data beside it |
| `nk_2obc.mod` | `model(linear)` with two occasionally-binding constraints, `estimated_params` |
| `nk_zlb_dynare.mod` | `model(linear)` with `occbin_constraints` (ZLB) |
| `sw2007.mod` | Smets-Wouters (2007): `model(linear)`, `varobs`, `estimated_params`, with data, mode and Hessian beside it |

Every shipped model carries a provenance row in
`inst/extdata/models/MODELS_SOURCE.md`.

Other models named in the vignettes — `fs2000.mod`, `caldara_rp.mod`,
`kpr1988.mod`, `nk_canonical.mod`, `nk_hs2016.mod`, `aikman_3obc.mod`,
`nz_base_nonlinear.mod` — are development fixtures the parser is tested
against. They are **not** shipped: several are third-party files whose
licences are incompatible with this package's, which is the whole reason the
provenance record exists.

```{r fixture-path}
system.file("extdata", "models", "rbc.mod", package = "dynhr")
```

---

## See also

- `?parse_mod` — function documentation with argument details
- `vignette("mod-conversion")` — converting Dynare models and using
  dynhr-specific annotations
- `vignette("solving")` — perturbation solver after parsing
- `vignette("estimation")` — Bayesian estimation workflow
