---
title: "SBC coverage matrix"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{SBC coverage matrix}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment  = "#>"
)
```

## Overview

dynhr maintains a **SBC (Simulation-Based Calibration) coverage matrix** that
tracks which sampler × likelihood combinations have been calibrated against
the Talts et al. (2018) rank-uniformity criterion. The matrix is regenerated
at the end of every battery run and stored in `inst/extdata/sbc_matrix.rds`.

### Status vocabulary

| Status | Meaning |
|---|---|
| **certified** | Passed the battery uniformity gate for ≥1 model |
| **to-certify** | Dispatch wired; battery cell not yet run |
| **characterized** | Runs but likelihood is approximate (e.g. Whittle at small T); p-values stored as provenance, not gated |
| **incompatible** | Mathematically impossible: gradient sampler × a stochastic, non-differentiable likelihood (tpf) — no usable gradient exists |
| **needs-gradient** | Gradient sampler × a deterministic likelihood with no analytic-gradient path implemented yet (pskf, student_t) — possible, not done |
| **pending** | Likelihood not yet implemented in dynhr |

```{r load-matrix}
rds_path <- system.file("extdata", "sbc_matrix.rds", package = "dynhr")

if (!nzchar(rds_path) || !file.exists(rds_path)) {
  ## Graceful fallback: the RDS is generated by the battery (DYNHR_SBC_CI=true)
  ## and is not committed. Fall back to the canonical static status table built
  ## by sbc_matrix_result() so this placeholder can never drift from the source.
  message(
    "sbc_matrix.rds not found. Run the SBC battery (DYNHR_SBC_CI=true) to ",
    "generate inst/extdata/sbc_matrix.rds with per-cell verdicts."
  )
  mat <- dynhr:::sbc_matrix_result(out_path = NULL)
} else {
  mat <- readRDS(rds_path)
}
```

```{r render-table}
## Pivot to wide form (samplers as rows, likelihoods as columns)
likelihoods <- unique(mat$likelihood)
samplers    <- unique(mat$sampler)

display_mat <- do.call(cbind, lapply(likelihoods, function(lik) {
  sub <- mat[mat$likelihood == lik, c("sampler", "status"), drop = FALSE]
  sub <- sub[match(samplers, sub$sampler), , drop = FALSE]
  stats::setNames(sub["status"], lik)
}))
rownames(display_mat) <- samplers

knitr::kable(
  display_mat,
  caption = paste0(
    "SBC coverage matrix. ",
    if (nzchar(rds_path) && file.exists(rds_path)) {
      sprintf("Generated %s.", format(file.info(rds_path)$mtime, "%Y-%m-%d"))
    } else {
      "(Static placeholder — run battery to populate.)"
    }
  ),
  format = "html"
)
```

## Notes

The matrix spans **7 samplers** — gradient-free (`rwmh`, `smc`, `dime`) and
gradient-based (`nuts`, `hmc`, `mala`, `chees`) — × **6 likelihoods**.
Gradient-based samplers require an analytic gradient, which only `gaussian_kf`,
`whittle`, and `cumulant` expose. On the other three likelihoods they're
unavailable — for two distinct reasons: `tpf` is **incompatible** (its particle
log-likelihood is stochastic and non-differentiable, so no usable gradient
*exists*), while `pskf` and `student_t` are **needs-gradient** (deterministic and
differentiable in principle, but with no analytic-gradient path implemented yet).

- **gaussian_kf**: Linear Gaussian state space via the Kalman filter. Exact
  posterior; the benchmark column. Analytic gradient → every sampler supported.
- **tpf**: Tempered particle filter (Herbst & Schorfheide 2019; second-order DGP). Pseudo-marginal
  `rwmh`/`smc`/`dime` are valid because the estimator is unbiased (Beaumont 2003;
  Andrieu & Roberts 2009). Gradient samplers are unsupported — the log-likelihood
  is stochastic and non-differentiable.
- **whittle**: Frequency-domain Whittle likelihood using the exact complex
  spectral density S(ω) ∈ ℂ^{n×n} (Hermitian positive definite for multi-observable
  models). Asymptotically approximate (O(1/T) bias), so every cell is
  **characterized** — p-values are stored as provenance, never gated. An analytic
  Whittle gradient exists, so the gradient samplers run. The earlier
  miscalibration of NUTS × Whittle at T = 200 (min p < 1e-4, posterior width)
  was the Whittle ×0.5 / 2π definition bug fixed in 0.9.3.35: re-run
  2026-09-27, 100 replications, min p = 0.35, calibrated (still not gated —
  the approximation is fixture-dependent). Any multi-obs
  cell must use the exact complex path (the default); 1-obs cells (S real) are
  bit-identical to it.
- **pskf**: Pruned Skewed Kalman Filter (CSN state-space recursion for skew-normal
  shocks) — **implemented**. Its CDF-based log-likelihood has fragile
  finite-difference gradients, so gradient samplers are unsupported; the
  gradient-free samplers are wired.
- **cumulant**: Cumulant-matching likelihood (Mutschler 2015), any perturbation
  order ≥ 1; orders ≥ 2 carry skewness/kurtosis. Exposes an analytic gradient
  (`cumulant_loglik_grad`), so **all** samplers — gradient-based included — are
  supported.
- **student_t**: Student-t innovation likelihood (Gaussian KF recursions +
  multivariate-t per-period density). No adjoint/analytic gradient path, so gradient
  samplers are unsupported; the gradient-free samplers are wired.

### Sampler caveats

- **chees** (`gaussian_kf` cell **certified** 2026-09-27: AR(1), T = 100,
  200 replications, min p = 0.44, 0 failed, on the corrected mass
  adaptation; batched runs, 7.6 h). An earlier 50-replication attempt
  (2026-06-27) was abandoned after ~3 h. Investigating it, and re-checking
  after the 0.9.3.50 mass-adaptation fix, gave three findings:
  - The cost is **not** the sampler: it is the per-evaluation Kalman/diffuse
    filter on near-unit-root prior draws (`rho → 0.999`), comparable to `hmc`'s
    2.6 h. chees itself uses short trajectories and is fast per draw.
  - On **ill-conditioned, correlated** posteriors — which the near-unit-root
    replications are — chees is **slow-mixing but not biased**. Its ChEES
    trajectory-length adaptation under-shoots (≈6 leapfrog steps where ≈√κ ≈ 32
    are needed). Re-checked 2026-09-25 on the corrected mass adaptation
    (0.9.3.50), 10 seeds, on a κ=1000 Gaussian rotated 45° (so no diagonal mass
    can precondition it):
    - chees large-axis variance is 0.81 / 0.86 / 0.96 (seed sd 0.16 / 0.11 /
      0.08) at n = 1.5k / 6k / 24k (truth 1.0). Its large-axis ESS is 16 / 50 /
      192.
    - NUTS reaches 1.01 (sd 0.11) at n = 1.5k with ESS ≈ 200. Per gradient,
      chees is about half as efficient (0.0013–0.0022 vs NUTS 0.0027–0.0043
      ESS/grad).

    The earlier single-seed figures (0.58 → 0.69 → 0.77) were mostly seed
    noise, not the mass inversion. Both marginal variances of this target are
    equal, so the inverted diagonal mass was only a scalar rescaling (10-seed
    re-run with the inversion restored: 0.84 / 0.84 / 1.00).
  - Where the inversion **did** matter is heterogeneous *scales*. On an
    axis-aligned κ=1000 Gaussian, corrected chees recovers both variances
    (1.007 ± 0.046) with ESS ≈ 950 of 1500 draws. That is ≈ 2.5× NUTS's ESS per
    gradient on the same target. The inverted adaptation left the large axis
    at 0.03 of its variance.

  The 2026-06 attempt's concern -- that the AR(1) fixture stresses exactly
  chees's weak region (correlated stiffness), so slow mixing could look like
  miscalibration -- did not materialise with enough draws: the batched
  200-replication run is calibrated. Slow mixing costs draws, not validity.

  Which gradient sampler mixes best is **model-dependent**. On the synthetic
  rotated κ=1000 Gaussian above, NUTS beats chees per gradient. On a real
  68-parameter New Zealand DSGE (NZSIM: 20 simulated observables, T = 132; 300 warmup +
  300 draws each, re-measured 2026-09-25 on the corrected mass adaptation),
  chees gave the best worst-parameter mixing of all the gradient samplers:

  | Sampler | Min bulk ESS | Min ESS per gradient |
  |---|---|---|
  | chees | 132 | 0.0078 |
  | NUTS, `"lowrank"` | 88 | 0.0022 |
  | NUTS, `"diagonal"` | 57 | 0.0022 |
  | hmc | 43 | 0.0014 |
  | NUTS, `"fisher_diag"` | 15 | 0.0006 |

  So do not rule out chees on stiff posteriors. Try it alongside NUTS and
  compare the minimum ESS.

## References

Talts, S., Betancourt, M., Simpson, D., Vehtari, A., & Gelman, A. (2018).
Validating Bayesian Inference Algorithms with Simulation-Based Calibration.
*arXiv:1804.06788*.

Beaumont, M. A. (2003). Estimation of population growth or decline in
genetically monitored populations. *Genetics*, 164(3), 1139–1160.

Andrieu, C., & Roberts, G. O. (2009). The pseudo-marginal approach for
efficient Monte Carlo computations. *Annals of Statistics*, 37(2), 697–725.
