---
title: "Solving DSGE models with dynhr"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Solving DSGE models with dynhr}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse  = TRUE,
  comment   = "#>",
  fig.width = 7,
  fig.height = 4,
  eval = TRUE
)
```

## Overview

This vignette covers the **deterministic** and **stochastic** solution
pipeline in dynhr:

1. Parse and compile a model
2. Find the steady state
3. Solve the linearised model (first- through fifth-order perturbation)
4. Compute impulse response functions and theoretical moments
5. Simulate model paths
6. Evaluate the Kalman likelihood (standard, Chandrasekhar, DARE, univariate methods)
7. Solve models with occasionally binding constraints (OBC)

For Bayesian estimation, see `vignette("estimation")`.  For the diagnostic
battery, see `vignette("diagnostics")`.

---

## 1  Model setup

We use the textbook RBC model shipped with the package.

```{r setup-model}
library(dynhr)

mod_path <- system.file("extdata", "models", "rbc.mod", package = "dynhr")
model    <- parse_mod(mod_path, verbose = FALSE)
compiled <- compile_model(model, verbose = FALSE)

cat(sprintf("Endogenous : %d\nExogenous  : %d\nParameters : %d\nEquations  : %d\n",
            length(model$var_names),
            length(model$varexo_names),
            length(model$param_names),
            length(model$equations)))
```

`parse_mod()` returns a `dynhr_mod` list; `compile_model()` builds the
symbolic residual and Jacobian functions used by the steady-state solver and
Kalman filter.

---

## 2  Steady state

### 2.1  Newton solver (default)

For models without a `steady_state_model` block, dynhr uses a Newton
method with backtracking line search:

```{r steady}
params <- model$param_values
steady  <- solve_steady(compiled, params,
                        endo_names = model$var_names,
                        exo_names  = model$varexo_names,
                        verbose    = FALSE)
print(steady)
```

The `$values` slot holds the named steady-state vector; `$converged` is
`TRUE` when the residual norm is below the solver tolerance (default
`1e-10`).

### 2.2  Checking steady-state residuals

```{r ss-residual}
ss <- steady$values
cat("Max |F(ss)|:", format(max(abs(steady$residuals)), scientific = TRUE), "\n")
```

All residuals should be below `1e-10` for reliable perturbation.

### 2.3  Models with an analytical block

When a `.mod` file contains a `steady_state_model` block, dynhr evaluates
it analytically and uses the result as the Newton starting point — or
directly if it already satisfies the model equations.  The RBC model uses
numerical Newton only; models that ship a `steady_state_model` block seed the
solver from that closed-form starting point instead.

---

## 3  First-order perturbation

### 3.1  Computing decision rules

`solve_perturbation()` calls the QZ decomposition (via the `QZ` package,
if installed) and returns the **decision-rule matrices** `ghx` and `ghu`:

```{r dr}
dr <- solve_perturbation(model, compiled, steady$values, params,
                         verbose = FALSE)
print(dr)
```

| Output        | Dimension             | Meaning |
|---------------|-----------------------|---------|
| `dr$ghx`      | `n_endo × n_state`   | State transition: how state variables respond to the previous state |
| `dr$ghu`      | `n_endo × n_exo`     | Impact matrix: how all endogenous variables respond to shocks on impact |
| `dr$bk_satisfied` | logical           | `TRUE` iff Blanchard–Kahn conditions hold |

### 3.2  Blanchard–Kahn conditions

The BK count is reported automatically:

```{r bk}
cat("BK satisfied:", dr$bk_satisfied, "\n")
cat("  n_stable   :", dr$n_stable,    "\n")
cat("  n_unstable :", dr$n_unstable,  "\n")
cat("  n_forward  :", dr$n_forward,   "\n")
```

`n_stable == n_forward` is the standard uniqueness-and-existence
condition.  If it fails, check your calibration for policy rule
coefficients below the Taylor principle.

### 3.3  Eigenvalue spectrum

```{r eigenvalues}
evs <- sort(Mod(dr$eigenvalues))
cat("Largest stable  :", round(evs[dr$n_stable], 4), "\n")
cat("Smallest unstable:", round(evs[dr$n_stable + 1L], 4), "\n")
```

The gap between the largest stable and smallest unstable eigenvalue
measures BK robustness.  A gap below 0.01 warrants attention.

### 3.4  Solver selection

`solve_perturbation()` has no `method` argument.  The solver is chosen
automatically at run time: it tries the `QZ` package (generalized Schur
decomposition) first, falls back to `geigen` if `QZ` is unavailable, and
finally falls back to base-R `eigen` for regular pencils.  There is no
user-facing way to pin a specific solver — the automatic cascade is the
only supported mode.

| Internal backend | Package  | Notes |
|------------------|----------|-------|
| QZ               | `QZ`     | Default; generalized Schur decomposition |
| geigen           | `geigen` | Alternative QZ; used when `QZ` is unavailable |
| eigen            | base R   | Fallback; exact only for regular pencils |

---

## 3.5  Higher-order perturbation

dynhr supports perturbation up to **order 5**, each with its own solver
and decision-rule class:

| Order | Solver | Decision-rule class | Key feature |
|-------|--------|---------------------|-------------|
| 2 | `solve_perturbation_order2()` | `DecisionRules2` | Quadratic terms, pruned state space |
| 3 | `solve_perturbation_order3()` | `DecisionRules3` | Cubic terms, skewness |
| 4 | `solve_perturbation_order4()` | `DecisionRules4` | Compact Sylvester (Levintal 2017) |
| 5 | `solve_perturbation_order5()` | `DecisionRules5` | Quintic terms |

`solve_perturbation()` dispatches automatically: pass `order = 2` (or
`3`, `4`, `5`) to select the solver, or pass a precomputed
`DecisionRules2/3/4/5` to the pipeline functions that accept `dr`.

Higher-order solutions include **σ-correction terms** for stochastic
steady-state shifts (`solve_sigma_cross()`, `solve_sigma_order4()`,
`solve_sigma_order5()`).

### 3.6  Cumulant computation

From a second-order perturbation, `compute_third_cumulant()` and
`compute_fourth_cumulant()` compute the closed-form skewness and excess
kurtosis of the pruned state space (Mutschler 2015), used by
`make_log_posterior_cumulant()` for cumulant-based Bayesian estimation.

---

## 4  Impulse response functions

```{r irfs}
irfs <- compute_irfs(dr, model, n_periods = 40L)
names(irfs)   # one matrix per shock
dim(irfs$eps_a)   # 40 × n_endo
```

Each element is a `T × n_endo` matrix.  Rows are horizons 1…T; columns
are endogenous variables in model order.

For higher-order solutions, use `compute_irfs_order2()`,
`compute_irfs_order3()`, `compute_irfs_order4()`, or
`compute_irfs_order5()`, which apply the pruned state-space
approximation (Andreasen et al. 2018).

### 4.1  Plotting a single shock

```{r irf-plot, fig.height=5}
vars  <- c("y", "c", "k", "l")
irf_a <- irfs$eps_a[, vars]

old_par <- par(mfrow = c(2, 2), mar = c(3, 3, 2, 1))
for (v in vars) {
  plot(irf_a[, v], type = "l", main = v, xlab = "Quarters", ylab = "% dev.")
  abline(h = 0, lty = 2, col = "grey60")
}
par(old_par)
```

---

## 5  Theoretical moments

`compute_moments()` solves the Lyapunov equation for the unconditional
variance–covariance matrix and returns correlations, autocorrelations, and
variance decompositions:

```{r moments}
moms <- compute_moments(dr, model)
cat("Variables:", colnames(moms$var_cov), "\n\n")
cat("Standard deviations (% of SS):\n")
sds <- sqrt(diag(moms$var_cov))
round(sds * 100, 3)
```

```{r correlations}
cat("Contemporaneous correlations:\n")
round(moms$correlation[1:4, 1:4], 3)
```

### 5.1  Variance decomposition

```{r vardec}
if (!is.null(moms$variance_decomposition)) {
  vd <- moms$variance_decomposition
  cat("Variance shares (%) attributed to eps_a:\n")
  round(vd["eps_a", ] * 100, 1)
}
```

For a single-shock model like RBC, all variance comes from `eps_a`.
Multi-shock models show the breakdown across shocks.

---

## 6  Stochastic simulation paths

`simulate_model()` draws forward paths by propagating shocks through
the state equation:

```{r simulate}
set.seed(123)
paths <- simulate_model(dr, n_periods = 200L, model = model, burn_in = 40L)
cat("Simulation shape:", nrow(paths), "x", ncol(paths), "\n")
```

```{r sim-plot, fig.height=3}
old_par <- par(mar = c(3, 3, 2, 1))
plot(paths[, "y"], type = "l", main = "Simulated output (y)",
     xlab = "Quarters", ylab = "Level")
par(old_par)
```

---

## 7  Kalman filter

Given an observed data matrix, `kalman_filter()` evaluates the
log-likelihood under the state-space representation implied by the
decision rules.  Multiple solver methods are available:

| Method | Description | Use case |
|--------|-------------|----------|
| `"auto"` | Picks fastest method automatically | Default |
| `"standard"` | Textbook Kalman filter | General use |
| `"chandrasekhar"` | Chandrasekhar recursions | Large state spaces, T >> n_state |
| `"dare"` | Bit-exact per-step filter (no steady-state shortcut) | Correctness oracle |
| `"reference"` | Alias for `"dare"` | Backward compatibility |
| `"univariate"` | Sequential scalar updates | Stochastically singular models |

```{r kf}
# Use a fragment of the simulated paths as stand-in for observed data
obs_vars <- c("y", "c")
Y_obs    <- as.matrix(paths[1:80, obs_vars])

suppressWarnings(
  kf <- kalman_filter(Y_obs, dr, model, params,
                      obs_vars        = obs_vars,
                      return_filtered = TRUE,
                      me_variance     = 1e-6)
)
cat("Log-likelihood  :", round(kf$loglik, 3), "\n")
cat("Filtered states  : dim", paste(dim(kf$filtered_states), collapse = " x "), "\n")
```

The `me_variance` argument adds a small measurement error on the diagonal
of the observation noise matrix.  This regularises the filter for
stochastically singular models where some observables are exactly linear
combinations of state variables.

### 7.1  Filter convergence

For stationary models the Kalman gain typically converges to a steady
state after 20–40 periods.  You can verify this by comparing
`kf$filtered_states` across time.

### 7.2  OBC Kalman filter

For models with occasionally binding constraints, `kalman_filter_obc_pkf()`
implements the **partially-Kalman filter (PKF)** of Boehl (2020), which
handles regime-dependent state-space representations.  The companion
`pkf_smoother_obc()` and `historical_decomposition_obc()` extend the
narrative to smoothed shocks and historical decompositions under OBC.

---

## 8  Models with occasionally binding constraints (OBC)

dynhr provides a comprehensive toolset for solving and analysing models
with occasionally binding constraints (ZLB, borrowing limits, etc.).

### 8.1  OccBin (piecewise linear)

The `occbin_solve_path()` function implements the OccBin algorithm —
guessing a regime path and verifying it against the model's piecewise-
linear decision rules:

```{r occbin, eval = FALSE}
parsed_tags <- obc_parse_tags(model)
regime_path <- occbin_solve_path(model, compiled, steady$values,
                                  params, parsed_tags, verbose = FALSE)
```

### 8.2  Boehl complementarity (LCP)

`solve_obc_lcp()` formulates the OBC problem as a linear complementarity
problem (LCP), which can detect binding regimes without guessing:

```{r lcp, eval = FALSE}
lcp_result <- solve_obc_lcp(model, compiled, steady$values, params,
                             tags = parsed_tags, verbose = FALSE)
```

### 8.3  Perfect-foresight Newton

`pf_newton_solve()` stacks all periods into a large nonlinear system and
solves with a Newton method.  Suitable for deterministic simulations of
OBC models.

### 8.4  OBC simulation and IRFs

- `simulate_sep()` — simulation with occasionally binding constraints
  using the stacked-epochs approach.
- `occbin_compute_irfs()` — IRFs under OBC, showing the effect of binding
  constraints on shock propagation.

---

## 9  Putting it together: `stoch_simul()`

`stoch_simul()` is a convenience wrapper that runs the perturbation solver,
computes IRFs, moments, and optionally simulates — matching the Dynare
`stoch_simul` command:

```{r stochsimul}
sim <- stoch_simul(model, verbose = FALSE)
names(sim)
```

```{r stochsimul-dr}
cat("BK satisfied :", sim$dr$bk_satisfied, "\n")
cat("IRF shocks   :", names(sim$irfs), "\n")
```

---

## 10  Solver options and tolerances

Key options can be passed via arguments or set globally:

| Concern                  | Argument / default                 |
|--------------------------|------------------------------------|
| Steady-state tolerance   | `tol = 1e-10` in `solve_steady()`  |
| Max Newton iterations    | `max_iter = 100`                   |
| BK strict mode           | `bk_strict = TRUE`                 |
| QZ imaginary tolerance   | `qz_imag_tol = 1e-12`              |
| Measurement error floor  | `me_variance = 0`                  |
| Perturbation order       | `order = 1` in `solve_perturbation()` |
| Numerical FD step        | `h = 1e-6` (adaptive per order)    |
| Rcpp backend             | `dynhr.use_rcpp = TRUE`            |

For models near a unit root, relaxing `qz_imag_tol` to `1e-6` is
sometimes needed.  Set `dynhr.use_rcpp = FALSE` to force the pure-R
Kalman filter fallback.

---

## 11  Changing parameters and re-solving

dynhr has no `set_param_value()` function.  To re-solve a model with
modified parameters, edit the `params` vector directly and re-run the
pipeline:

```{r reparameterise, eval = FALSE}
# Start from the calibrated values
params2 <- model$param_values

# Change one or more parameters
params2["rho"] <- 0.95

# Re-solve — no recompilation needed
ss2 <- solve_steady(compiled, params2,
                    endo_names = model$var_names,
                    exo_names  = model$varexo_names,
                    verbose    = FALSE)
dr2 <- solve_perturbation(model, compiled, ss2$values, params2, verbose = FALSE)
```

`params` is a plain named numeric vector, so standard R subsetting and
assignment apply.  `compile_model()` is called only once; the compiled
symbolic functions accept any `params` vector at evaluation time.

---

## 12  Log-linear models and the level-deviation convention

### 12.1  How dynhr represents decision-rule output

dynhr always works in **levels**.  The perturbation solution is:

```
y_t  ≈  y_ss  +  ghx · (x_{t-1} − x_ss)  +  ghu · ε_t
```

where `y_ss` is the (deterministic) steady-state vector and the matrices
`ghx`, `ghu` are measured in **level** units.  `simulate_model()` and
`compute_moments()` return deviations from the steady state in levels.

### 12.2  Converting a log-linearised Dynare model

Dynare's `loglinear` option log-linearises the model before solving, so its
`dr.ghx` and `dr.ghu` are in **log-deviation** units.  To recover dynhr's
level-deviation solution from Dynare's log-deviation solution, apply the
following element-wise conversion:

```
G_log[i, j]  =  (ys[j] / ys[i]) * G_lev[i, j]
H_log[i, j]  =  (1 / ys[i])     * H_lev[i, j]
```

where `ys` is the steady-state vector, `G_lev` = dynhr's `ghx`, and
`H_lev` = dynhr's `ghu`.  These formulas recover **machine precision** when
applied to a model solved in levels (confirmed across Hansen_1985 and
Ghironi_Melitz_2005 in the replication test suite).

To convert dynhr simulated paths to **log-deviations** (matching Dynare
`loglinear` output), divide each variable column by its steady-state value:

```r
## sim is (n_periods x n_endo) deviations from SS in levels
## ys is the named steady-state vector
sim_logdev <- sweep(sim, 2, dr$ys[colnames(sim)], `/`)
```

Note that `simulate_model()` returns **level deviations** (`sim[t, v]` = the
level-space deviation of variable `v` at time `t` from its steady state),
not log-deviations.

### 12.3  Models with a `loglinear;` statement in the `.mod` file

dynhr does not currently implement the `loglinear;` block option.  To compare
with a Dynare `loglinear` run:

1. Solve the model in dynhr (levels, as usual).
2. Apply the `G_log` / `H_log` conversion above to dynhr's `ghx` / `ghu`
   matrices before comparing.
3. Divide simulation paths by `dr$ys` to convert to log-deviations.

This workflow recovers numerical agreement with Dynare's `loglinear` results
to machine precision for the standard linearised models.

---

## 13  Seeding the steady-state solver

For models with multiple steady states or slow Newton convergence, pass an
initial guess `y0` to `solve_steady()`:

```{r ss-seeding, eval = FALSE}
## y0 is a named numeric vector of starting values for the solver.
## Useful when the default zero-start point misses the intended equilibrium.
ss_seeded <- solve_steady(compiled, params,
                          endo_names = model$var_names,
                          exo_names  = model$varexo_names,
                          y0         = c(C = 0.8, k = 9.2, z = 1.0),
                          verbose    = FALSE)
```

`y0` accepts a **named** vector; only the names present in `endo_names` are
used, so a partial vector is fine.  Unspecified variables start from zero.

This is the recommended approach when:

- dynhr's Newton converges to a different equilibrium than Dynare's (e.g.
  because Dynare's `*_steadystate.m` pins a specific root): seed from Dynare's
  steady-state values.
- The model has multiple steady states (e.g. indeterminacy with multiple
  fixed points): supply the economically-relevant starting point.
- Newton diverges from the zero start: seed from approximate analytical values.

```{r ss-seeding-example, eval = FALSE}
## Example: re-solve after seeding from a known steady state
dynare_ys <- c(Y = 1.099, C = 0.741, k = 14.36, ...)   # from Dynare's oo_.steady_state
ss_seeded <- solve_steady(compiled, params,
                          endo_names = model$var_names,
                          exo_names  = model$varexo_names,
                          y0 = dynare_ys, verbose = FALSE)
```

---

## See also

- `vignette("estimation")` — Bayesian estimation pipeline
- `vignette("diagnostics")` — diagnostic battery D1–D30
- `?solve_perturbation_order2`, `?solve_perturbation_order3` — higher-order solvers
- `?solve_obc_lcp`, `?occbin_solve_path` — OBC tools
- `?solve_sigma_cross`, `?solve_sigma_order4` — sigma-correction terms

---

## References

Anderson, B. D. O., & Moore, J. B. (1979). *Optimal Filtering*. Prentice-Hall.

Andreasen, M. M., Fernandez-Villaverde, J., & Rubio-Ramirez, J. F. (2018). The pruned state-space system for non-linear DSGE models. *Review of Economic Studies*, 85(1), 1-49.

Binning, A. (2013). Underidentified SVAR models: A framework for combining short-run and long-run restrictions. *Norges Bank Working Paper*.

Boehl, G. (2020). Efficient solution and computation of models with occasionally binding constraints. *Deutsche Bundesbank Discussion Paper*, 38/2020.

Cottle, R. W., Pang, J.-S., & Stone, R. E. (2009). *The Linear Complementarity Problem*. SIAM.

Guerrieri, L., & Iacoviello, M. (2015). OccBin: A toolkit for solving dynamic models with occasionally binding constraints easily. *Journal of Monetary Economics*, 70, 22-38.

Kalman, R. E. (1960). A new approach to linear filtering and prediction problems. *Journal of Basic Engineering*, 82(1), 35-45.

Klein, P. (2000). Using the generalized Schur form to solve a multivariate linear rational expectations model. *Journal of Economic Dynamics and Control*, 24(10), 1405-1423.

Levintal, O. (2017). Fifth-order perturbation solution of DSGE models. *Journal of Economic Dynamics and Control*, 80, 1-16.

Mutschler, W. (2015). Identification of DSGE models — The effect of higher-order approximation and pruning. *Journal of Economic Dynamics and Control*, 56, 1-33.

Schmitt-Grohé, S., & Uribe, M. (2004). Solving dynamic general equilibrium models using a second-order approximation to the policy function. *Journal of Economic Dynamics and Control*, 28(4), 755-775.

Strid, I., & Walentin, K. (2011). Block Kalman filtering for large-scale DSGE models. *Computational Economics*, 39(2), 145-160.

Villemot, S. (2011). Solving rational expectations models by first-order perturbation. *Dynare Working Paper Series*, 6.

```{r session}
sessionInfo()
```
