---
title: "`r local({ p <- params$meta_rds; if (!is.null(p) && nzchar(p) && file.exists(p)) readRDS(p)$model_name else params$model_name })` -- dynhr Diagnostic Report"
date: "`r format(Sys.time(), '%Y-%m-%d %H:%M')`"
params:
  diags_rds:         ""     # path to RDS holding the dynhr_diagnostic_suite
  irfs_rds:          ""     # path to RDS holding a named IRF list (optional)
  llm_txt:           ""     # path to pre-rendered LLM summary text
  exec_txt:          ""     # path to pre-rendered executive summary text
  meta_rds:          ""     # path to RDS holding the report-meta list
  escape_r:          ""     # path to the sourced .escape_for() helper
  plotutil_r:        ""     # path to the sourced figure-height / pagination helpers
  model_name:        "model"
  show_explanations: true   # show collapsible methodology blocks per diagnostic
format:
  html:
    toc: true
    toc-depth: 4
    toc-location: left
    embed-resources: true
    code-fold: true
    # Executive-summary / LLM blocks are fixed-width text wider than the
    # callout at 68 parameters: wrap instead of clipping (NZSIM-scale check).
    code-overflow: wrap
    theme: cosmo
    html-math-method: plain
    mainfont: "Source Sans 3, Source Sans Pro, Segoe UI, sans-serif"
    monofont: "JetBrains Mono, Consolas, monospace"
    fontsize: "1.05rem"
    fig-width: 9
    fig-height: 6.5
    fig-dpi: 110
    fig-align: center
    include-in-header:
      text: |
        <style>
          :root {
            --dynhr-blue:     #0077BB;   /* Tol vibrant blue (primary) */
            --dynhr-ink:      #1A1A1A;
            --dynhr-body:     #333333;
            --dynhr-muted:    #777777;
            --dynhr-rule:     #DDDDDD;
            --bs-primary:     #0077BB;
            --bs-link-color:  #0077BB;
          }
          html, body { font-family: 'Source Sans 3', 'Source Sans Pro', 'Segoe UI', sans-serif;
                       font-size: 17px; color: var(--dynhr-body); }
          h1, h2, h3, h4, h5, h6, .title, .subtitle {
            font-family: 'Source Serif 4', 'Source Serif Pro', Georgia, serif;
            color: var(--dynhr-ink); font-weight: 600;
          }
          h1.title { color: var(--dynhr-blue); border-bottom: 3px solid var(--dynhr-blue);
                     padding-bottom: 6px; }
          h2 { color: var(--dynhr-blue); border-bottom: 1px solid var(--dynhr-rule);
               padding-bottom: 4px; margin-top: 1.6em; }
          h3, h4 { color: var(--dynhr-ink); }
          a, a:visited { color: var(--dynhr-blue); }
          code, kbd, pre, samp { font-family: 'JetBrains Mono', Consolas, monospace;
                                  font-size: 0.92em; }
          /* Executive-summary / LLM blocks are fixed-width text that can be
             wider than the callout (68-parameter NZSIM): wrap, never clip.
             The ```text fence is not a sourceCode block, so Quarto's own
             code-overflow rule does not reach it. */
          .callout pre, .callout pre code { white-space: pre-wrap !important;
                                            overflow-wrap: anywhere; }
          .table thead th { background-color: rgba(0,119,187,0.08);
                            color: var(--dynhr-ink); border-bottom: 2px solid var(--dynhr-blue); }
          .table tbody tr:hover { background-color: rgba(0,119,187,0.04); }
          .panel-tabset > .nav-tabs .nav-link.active { color: var(--dynhr-blue);
            border-bottom: 3px solid var(--dynhr-blue); font-weight: 600; }
          details > summary { color: var(--dynhr-blue) !important; }
          blockquote { border-left: 4px solid var(--dynhr-blue); color: var(--dynhr-muted); }
        </style>
execute:
  echo: false
  warning: false
  message: false
---

```{r setup}
`%||%` <- function(a, b) if (is.null(a) || (is.character(a) && !nzchar(a))) b else a
.default_fig_height <- knitr::opts_chunk$get("fig.height") %||% 6.5
.max_fig_height     <- 40   # in; ggsave() refuses > 50 and taller is unreadable anyway

# Many-panel figures are SPLIT rather than squashed: at NZSIM scale a
# 68-parameter facet plot capped at .max_fig_height gave 2.0 in per facet row
# and a 190-level correlation plot 0.2 in per level.  The limits below are the
# HTML ones (the PDF template uses tighter ones for its 9 in page).
.paginate_max_rows <- 12L   # facet rows per page
.paginate_max_y    <- 60L   # discrete y levels per page
diags <- if (nzchar(params$diags_rds %||% "") && file.exists(params$diags_rds))
  readRDS(params$diags_rds) else NULL
irfs  <- if (nzchar(params$irfs_rds  %||% "") && file.exists(params$irfs_rds))
  readRDS(params$irfs_rds)  else NULL

# report-meta.rds is the ONLY source of classification, ordering, colour,
# explanation and provenance information: the template must never re-derive a
# badge from $pass / $warn / $errored (that duplication is what let the PDF
# template report a WARN as a PASS).  write_report() always spools it.
meta <- if (nzchar(params$meta_rds %||% "") && file.exists(params$meta_rds))
  readRDS(params$meta_rds) else NULL
if (is.null(meta) && !is.null(diags))
  stop("report-meta.rds was not supplied (params$meta_rds); ",
       "the template cannot classify diagnostics without it.")

# .escape_for() is spooled next to the data by write_report() because dynhr
# itself may not be installed in the Quarto subprocess.  It is the ONLY way
# text reaches a sink in this template.
if (nzchar(params$escape_r %||% "") && file.exists(params$escape_r)) {
  source(params$escape_r)
}
if (!exists(".escape_for", inherits = TRUE))
  stop("escape.R was not supplied (params$escape_r); ",
       "the template cannot safely interpolate text without .escape_for().")

# .fig_height_for() / .paginate_plot() are spooled the same way, for the same
# reason.  The template keeps NO copy of the height rule.
if (nzchar(params$plotutil_r %||% "") && file.exists(params$plotutil_r)) {
  source(params$plotutil_r)
}
if (!exists(".fig_height_for", inherits = TRUE) ||
    !exists(".paginate_plot", inherits = TRUE))
  stop("plotutil.R was not supplied (params$plotutil_r); ",
       "the template cannot size or paginate figures without it.")

show_explanations <- isTRUE(params$show_explanations)

.LEVELS  <- c("ERROR", "FAIL", "WARN", "INFO", "PASS")
.CALLOUT <- c(ERROR = "important", FAIL = "important", WARN = "warning",
              INFO  = "note",      PASS = "tip")

diag_items <- if (!is.null(diags))
  Filter(function(r) inherits(r, "dynhr_diagnostic"), diags) else list()
if (length(diag_items) > 0L) {
  ord <- meta$order[meta$order %in% names(diag_items)]
  ord <- c(ord, setdiff(names(diag_items), ord))
  diag_items <- diag_items[ord]
}
diag_badges <- if (length(diag_items) > 0L)
  meta$badges[names(diag_items)] else character(0)
counts <- if (is.null(meta)) stats::setNames(integer(length(.LEVELS)), .LEVELS) else
  meta$counts

.colour <- function(level) {
  cc <- meta$colours[[level]]
  if (is.null(cc) || is.na(cc)) "#BBBBBB" else cc
}

.anchor <- function(nm) paste0("diag-", gsub("[^A-Za-z0-9_-]", "-", nm))

# .diag_meta_for() falls back to the diagnostic's own name when a diagnostic
# has no .DIAG_META row; don't print the name twice in that case.
.disp_id <- function(nm) {
  did <- meta$disp_id[[nm]] %||% ""
  if (identical(did, nm)) "" else did
}

# Render multi-line $summary text as a bullet list so each line stays on its
# own row inside the Quarto callout block (otherwise paragraph folding can
# strand the closing ::: and the whole callout renders as raw markdown).
.render_summary <- function(summary_text) {
  if (is.null(summary_text) || !nzchar(summary_text)) return("")
  lines <- strsplit(summary_text, "\n", fixed = TRUE)[[1]]
  lines <- lines[nzchar(lines)]
  if (length(lines) <= 1L) return(.escape_for(summary_text, "html"))
  items <- paste0("<li>", .escape_for(lines, "html"), "</li>", collapse = "")
  paste0("<ul>", items, "</ul>")
}

.badge_html <- function(level) {
  sprintf('<span class="badge" style="background:%s;color:white;">%s</span>',
          .colour(level), level)
}

# Plots are written out explicitly rather than print()ed inside the chunk:
# knitr's figure options (fig.alt / fig.cap / fig.height) are snapshotted at
# chunk start, so a mid-chunk opts_chunk$set() cannot give each plot in a loop
# its own caption, alt text or height.  Writing the PNG here and emitting the
# markdown image gives every figure alt text (B7) and its requested height.
.FIG_DIR <- "report-figs"
dir.create(.FIG_DIR, showWarnings = FALSE)

.print_plot <- function(p, nm = NULL, pnm = NULL) {
  if (!inherits(p, "gg")) return(invisible(NULL))
  ## Split a many-panel / many-level figure first; a plot that cannot be
  ## split comes back as a one-element list holding p itself.
  pages <- .paginate_plot(p, max_rows = .paginate_max_rows,
                          max_y = .paginate_max_y,
                          default = .default_fig_height,
                          maximum = .max_fig_height)
  np <- length(pages)
  for (k in seq_len(np)) {
    pk  <- pages[[k]]
    h   <- attr(pk, "dynhr_fig_height", exact = TRUE)
    h   <- if (!is.null(h) && is.finite(h)) as.numeric(h) else
      .fig_height_for(pk, .default_fig_height, .max_fig_height)
    ## Per-parameter facet plots scale their requested height with the number
    ## of estimated parameters (a 68-parameter model asked for 74 in) and
    ## ggsave() refuses anything over 50 in.  Cap at .max_fig_height: a plot
    ## that needs more than that is still delivered, just denser.
    h   <- min(h, .max_fig_height)
    ttl <- pk$labels$title %||% (pnm %||% "diagnostic plot")
    cap <- if (is.null(nm)) ttl else sprintf("%s -- %s", nm, ttl)
    key <- gsub("[^A-Za-z0-9]+", "-", paste(c(nm, pnm, ttl), collapse = "-"))
    ## The page index is part of the file key: without it every page of a
    ## split figure would overwrite the previous page's PNG.
    key <- paste0(substr(key, 1L, 80L), if (np > 1L) sprintf("-p%02d", k) else "")
    f   <- file.path(.FIG_DIR, paste0(key, ".png"))
    ggplot2::ggsave(f, pk, width = 9, height = h, dpi = 110, units = "in",
                    limitsize = FALSE)
    cat(sprintf('\n\n![%s](%s){fig-alt="%s" width=100%%}\n\n',
                .escape_for(cap, "markdown"), f,
                gsub('"', "'", cap, fixed = TRUE)))
  }
  invisible(NULL)
}

# Methodology text comes from meta$explanations (R/diag-explanations.R), so the
# template never carries a second, drifting copy of the thresholds.
.explanation_block <- function(nm) {
  if (!show_explanations) return(invisible(NULL))
  info <- meta$explanations[[nm]]
  if (is.null(info)) return(invisible(NULL))
  cat('<details>\n')
  cat('<summary style="cursor:pointer;color:#555;font-size:0.85em;">')
  cat('&#9432; What this diagnostic checks</summary>\n')
  cat('<div style="padding:8px 16px;font-size:0.9em;color:#444;">\n')
  cat(sprintf('<p><strong>What:</strong> %s</p>\n',
              .escape_for(info$what %||% "", "html")))
  cat(sprintf('<p><strong>What to look for:</strong> %s</p>\n',
              .escape_for(info$look_for %||% "", "html")))
  cat(sprintf('<p><strong>Reference:</strong> <em>%s</em></p>\n',
              .escape_for(info$citation %||% "", "html")))
  cat('</div>\n</details>\n\n')
  invisible(NULL)
}

.diag_block <- function(nm) {
  r     <- diag_items[[nm]]
  level <- unname(diag_badges[[nm]])
  did   <- .disp_id(nm)
  cat(sprintf('\n#### %s%s -- %s {#%s}\n\n',
              .escape_for(nm, "markdown"),
              if (nzchar(did)) paste0(" -- ", .escape_for(did, "markdown")) else "",
              level, .anchor(nm)))
  cat(sprintf(':::: {.dynhr-lvl-%s}\n', tolower(level)))
  cat(sprintf('::: {.callout-%s title="%s"}\n', .CALLOUT[[level]], level))
  cat(.render_summary(r$summary %||% ""), "\n\n", sep = "")
  act <- meta$actions[[nm]] %||% ""
  if (nzchar(act))
    cat(sprintf('<p><strong>Action:</strong> %s</p>\n\n',
                .escape_for(act, "html")))
  .explanation_block(nm)
  cat(":::\n::::\n\n")
  for (pnm in names(r$plots)) .print_plot(r$plots[[pnm]], nm, pnm)
  invisible(NULL)
}
```

```{r level-css}
#| results: asis
# All level colours come from meta$colours (.badge_colour()); nothing is
# hard-coded here.  The print block makes every tab visible on paper.
cat("<style>\n")
for (lv in .LEVELS) {
  cat(sprintf(".dynhr-lvl-%s .callout { border-left-color: %s !important; }\n",
              tolower(lv), .colour(lv)))
  cat(sprintf(".dynhr-lvl-%s .callout .callout-title { color: %s; }\n",
              tolower(lv), .colour(lv)))
}
cat(".dynhr-legend { font-size: 0.9em; color: #555; margin: 0.4em 0 1.2em 0; }\n")
cat(".dynhr-legend .badge { margin-right: 0.35em; }\n")
cat("@media print {\n")
cat("  .tab-content > .tab-pane { display: block !important; opacity: 1 !important; }\n")
cat("  .panel-tabset > .nav-tabs, .nav-tabs { display: none !important; }\n")
cat("  details { display: block !important; }\n")
cat("  details > div { display: block !important; }\n")
cat("}\n")
cat("</style>\n")
```

## Executive Summary

```{r exec-summary}
#| results: asis
# Read the pre-rendered executive summary text (written by write_report()).
# It goes into a fenced code block: Pandoc escapes the content itself, so no
# hand-escaping (and no chance of a stray tag) is involved.
exec_lines <- if (nzchar(params$exec_txt %||% "") && file.exists(params$exec_txt))
  readLines(params$exec_txt, warn = FALSE) else NULL
if (!is.null(exec_lines)) {
  cat(":::: {.dynhr-lvl-fail}\n")
  cat('::: {.callout-important title="Executive Summary"}\n')
  cat("````text\n")
  cat(paste(exec_lines, collapse = "\n"))
  cat("\n````\n")
  cat(":::\n::::\n")
} else {
  cat("::: {.callout-note}\nNo diagnostic suite loaded (diags_rds not supplied).\n:::\n")
}
```

## Provenance

```{r provenance}
#| results: asis
# A diagnostic report is a claim about a specific model, at a specific commit,
# against specific data: state that up front rather than 30 pages later.
prov_labels <- c(dynhr_version = "dynhr version", git_commit = "Git commit",
                 model_name = "Model", model_file = "Model file",
                 n_obs = "Observations", obs_vars = "Observables",
                 data_hash = "Data hash", sample_span = "Sample span",
                 seed = "Seed", sampler = "Sampler", n_draws = "Draws",
                 n_warmup = "Warmup", n_chains = "Chains",
                 rendered_at = "Rendered at")
prov <- meta$provenance %||% list()
cat('<table class="table table-sm">\n<tbody>\n')
for (k in names(prov_labels)) {
  v <- prov[[k]]
  txt <- if (is.null(v) || length(v) == 0L || all(is.na(v)))
    "not recorded" else paste(as.character(v), collapse = ", ")
  cat(sprintf("<tr><th style=\"width:12em;\">%s</th><td>%s</td></tr>\n",
              .escape_for(prov_labels[[k]], "html"), .escape_for(txt, "html")))
}
cat("</tbody></table>\n")
```

## Summary dashboard

```{r dashboard}
#| results: asis
cat(sprintf(
  "<p><strong>%d PASS &nbsp; %d WARN &nbsp; %d FAIL &nbsp; %d ERROR &nbsp; %d INFO</strong> &mdash; %d diagnostics</p>\n",
  counts[["PASS"]], counts[["WARN"]], counts[["FAIL"]], counts[["ERROR"]],
  counts[["INFO"]], length(diag_items)))

cat('<p class="dynhr-legend">')
for (lv in .LEVELS) cat(.badge_html(lv))
cat(sprintf("<br>%s<br><em>%s</em></p>\n",
            .escape_for(meta$legend %||% "", "html"),
            .escape_for("Palette: Paul Tol vibrant (colour-blind safe).", "html")))

cat('<table class="table table-sm table-hover">\n')
cat('<thead><tr><th>ID</th><th>Status</th><th>Summary</th></tr></thead>\n<tbody>\n')
for (nm in names(diag_items)) {
  r     <- diag_items[[nm]]
  level <- unname(diag_badges[[nm]])
  # Escape FIRST, then truncate, so a truncation can never cut an escape in
  # half and re-open the hazard it was meant to neutralise.
  short <- .escape_for(gsub("\n+", "; ", r$summary %||% ""), "html")
  if (nchar(short) > 110L) short <- paste0(substr(short, 1L, 107L), "...")
  cat(sprintf(
    '<tr><td><a href="#%s"><code>%s</code></a> <small>%s</small></td><td>%s</td><td>%s</td></tr>\n',
    .anchor(nm), .escape_for(nm, "html"), .escape_for(.disp_id(nm), "html"),
    .badge_html(level), short))
}
cat("</tbody></table>\n")
```

---

## Diagnostic details

```{r diag-tabs}
#| results: asis
# Severity order: ERROR (did not run) is strictly worse than FAIL (ran and
# failed), and PASS is the tab nobody needs.  Empty levels are not emitted at
# all, so the document always opens on the most severe NON-EMPTY tab.
present <- .LEVELS[vapply(.LEVELS, function(lv) any(diag_badges == lv), logical(1))]
if (length(present) == 0L) {
  cat("*No diagnostics to report.*\n")
} else {
  cat("::: {.panel-tabset}\n\n")
  for (lv in present) {
    cat(sprintf("\n### %s (%d)\n\n", lv, sum(diag_badges == lv)))
    for (nm in names(diag_items)[diag_badges == lv]) .diag_block(nm)
  }
  cat("\n:::\n")
}
```

---

## Impulse responses

```{r irfs-static}
#| results: asis
if (is.null(irfs) || length(irfs) == 0L) {
  cat("::: {.callout-note}\nNo IRFs were supplied to `write_report()`.\n:::\n")
} else {
  horizons <- vapply(irfs, function(p) {
    d <- if (inherits(p, "gg")) p$data else NULL
    if (is.null(d) || is.null(d$horizon)) NA_integer_ else as.integer(max(d$horizon))
  }, integer(1))
  cat(sprintf("Units: %s\n\n", .escape_for(meta$irf_units %||% "", "markdown")))
  if (length(unique(horizons[!is.na(horizons)])) > 1L)
    cat(sprintf("Horizons differ across shocks (%s periods); the x-ranges below are not comparable.\n\n",
                .escape_for(paste(sort(unique(horizons[!is.na(horizons)])),
                                  collapse = ", "), "markdown")))
  cat("::: {.panel-tabset}\n\n")
  for (shock_nm in names(irfs)) {
    cat(sprintf("### %s\n\n", .escape_for(shock_nm, "markdown")))
    p <- irfs[[shock_nm]]
    # The tab heading already names the shock: drop the duplicated plot title
    # but keep it as the figure caption / alt text.
    if (inherits(p, "gg")) {
      ttl <- p$labels$title %||% shock_nm
      p2  <- p + ggplot2::labs(title = NULL)
      attr(p2, "dynhr_fig_height") <- attr(p, "dynhr_fig_height", exact = TRUE)
      .print_plot(p2, NULL, ttl)
    }
    cat("\n\n")
  }
  cat(":::\n")
}
```

---

## LLM-readable summary

Paste this block into an LLM prompt to get targeted modelling suggestions.

```{r llm-block}
#| results: asis
if (nzchar(params$llm_txt %||% "") && file.exists(params$llm_txt)) {
  cat("````text\n")
  cat(paste(readLines(params$llm_txt, warn = FALSE), collapse = "\n"))
  cat("\n````\n")
}
```

---

## Render environment

This is the environment the *report* was rendered in (a separate Quarto R
subprocess). The environment the *diagnostics* ran in is in Provenance above.

```{r session}
sessionInfo()
```
